---
title: "Lab 10: One-Way ANOVA"
subtitle: "Which Study Method Works Best?"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "PSY 348 - Statistics for Psychology"
format:
  revealjs:
    theme: ../custom.scss
    width: 1600
    height: 900
    incremental: true
    touch: true
    chalkboard: true
---

## One-Way ANOVA

### Concept Review

-   Compares **means across 3+ groups**

-   Used when you want to test if at least one group differs from the others

-   Tests the null hypothesis that all group means are equal: $H_0: \mu_1 = \mu_2 = \mu_3$

## Scenario

> Three groups of students used different methods to study for a quiz:
>
> -   **Video lectures**
> -   **Reading the textbook**
> -   **Flashcards**
>
> Each student then took the same 10-question quiz.

-   What is the independent variable (IV) here? What is the dependent variable (DV)?

## Simulated Data in R

```{r}
library(dplyr)
library(ggplot2)

set.seed(123)

study_data <- data.frame(
  method = rep(c("Video", "Textbook", "Flashcards"), each = 15),
  score = c(
    rnorm(15, mean = 7.5, sd = 1),    # Video
    rnorm(15, mean = 6.5, sd = 1.2),  # Textbook
    rnorm(15, mean = 8.0, sd = 0.8)   # Flashcards
  )
)

head(study_data)
```

## Visualizing Scores

```{r}
ggplot(study_data, aes(x = method, y = score, fill = method)) +
  geom_boxplot() +
  labs(
    title = "Quiz Scores by Study Method",
    x = "\nStudy Method",
    y = "Quiz Score\n"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Running the ANOVA

```{r}
#| echo: true
anova_result <- aov(score ~ method, data = study_data)
summary(anova_result)
```

::: highlight-last
-   Look at the **F value** and **p-value**

-   If $p < .05$, reject the null hypothesis
:::

## Post Hoc: Tukey HSD

```{r}
#| echo: true
TukeyHSD(anova_result)
```

::: highlight-last
-   Shows **which pairs** of study methods differ significantly

-   Adjusts for multiple comparisons, so we don't inflate our Type I error rate by running three separate t-tests
:::

## Discussion Prompts

1.  Which group performed best?

2.  Is the difference statistically significant?

3.  Is the difference **practically** meaningful? (Hint: look back at the boxplot — how much does a 1-point difference on a 10-question quiz actually matter?)

4.  What might influence quiz performance beyond study method?

## Wrap-Up

::: highlight-last
-   One-way ANOVA compares 3+ means

-   A significant F-test tells you *at least one* group differs — Tukey HSD tells you *which* ones

-   Real-world applications: education, psychology, marketing, and beyond
:::

## Submit on Brightspace

Using `study_data`, answer:

1.  Report the F-statistic, degrees of freedom, and p-value from your ANOVA.

2.  Which pair(s) of study methods differ significantly per the Tukey HSD test?

3.  In your own words, explain what it would mean if this ANOVA had *not* been significant.
