---
title: "Exam 1 Review"
subtitle: "Chapters 1 - 4"
author: "Dave Brocker"
institute: "Farmingdale State College"
format: 
  revealjs:
    theme: custom.scss
    incremental: true   
    touch: true
    chalkboard: true
    width: 1600
    height: 900
    scrollable: true
    lightbox: true
    drop:
      engine: webr
      webr:
        packages:
         - ggplot2
         - dplyr
revealjs-plugins:
  - drop
---

## Learning Goals

-   Compute sample mean and standard deviation by hand.
-   Interpret what the mean and SD tell us about data.
-   Standardize raw scores into Z-scores.
-   Interpret Z-scores (direction and distance from mean).
-   Relate Z-scores to probability under a normal curve.

## 1. Mean Calculation

### Sample Mean Calculation

> A psychology researcher collects the following data on the number of hours five participants spent practicing a task: `3, 5, 8, 2, 7`.

**Question:** What is the average number of hours spent practicing?

::::: columns
::: {.column width="50%"}
<details>

<summary>Manual Solution</summary>

<p>$\bar{x} = \frac{\sum(3+5+8+2+7)}{5} = \frac{25}{5} = 5\text{ hours}$</p>

</details>
:::

::: {.column width="50%"}
<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
mean(c(3,5,8,2,7))
```

</details>
:::
:::::

## 2. Mean Calculation

### Sample Mean Calculation

> A study measures the changes in mood scores before and after an intervention. The changes in mood scores are: `-4, -2, 0, 3, 5`.

**Question:** Calculate the mean change in mood score.

::::: columns
::: {.column width="50%"}
<details>

<summary>Manual Solution</summary>

<p>$\bar{x} = \frac{\sum(-4 + (-2) + 0 + 3 + 5)}{5} = \frac{2}{5} = 0.4$</p>

</details>
:::

::: {.column width="50%"}
<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
mean(c(-4,-2,0,3,5))
```

</details>
:::
:::::

## 3. Standard Deviation Calculation

### Sample Standard Deviation

> A psychologist measures the number of stressful events experienced by a small group of participants over a week: `2, 3, 6, 4, 5`.

**Question:** Calculate the standard deviation of stressful events.

<details>

<summary>Manual Solution</summary>

<p>

<ul>

<li>First compute the sample mean: $\bar{x} = \frac{2+3+6+4+5}{5} = \frac{20}{5} = 4.$</li>

<li>Compute squared deviations: $(2-4)^2=4,\ (3-4)^2=1,\ (6-4)^2=4,\ (4-4)^2=0,\ (5-4)^2=1$</li>

<li>Sum of squares = $4+1+4+0+1 = 10$</li>

<li>Sample variance $s^2 = \frac{\sum (x_i-\bar{x})^2}{n-1} = \frac{10}{4} = 2.5)$</li>

<li>Sample SD $s = \sqrt{2.5} \approx 1.58$</li>

</ul>

</p>

</details>

<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
sd(c(2,3,6,4,5))
```

</details>

## 4. Standard Deviation Calculation

### Sample Standard Deviation

> **Sample Standard Deviation** Participants' anxiety scores in a clinical trial are recorded as: `4, 7, 3, 5, 8, 6, 2`.

**Question:** Compute the standard deviation of these anxiety scores.

::::: columns
::: {.column width="50%"}
<details>

<summary>Solution</summary>

<p>First compute the sample mean: $\bar{x} = \frac{4+7+3+5+8+6+2}{7} = \frac{35}{7} = 5$<br> Squared deviations: $(4-5)^2=1\\(7-5)^2=4\\ (3-5)^2=4\\(5-5)^2=0\\(8-5)^2=9\\(6-5)^2=1\\ (2-5)^2=9\\$ <br> Sum = $1+4+4+0+9+1+9 = 28$.</p>

</details>
:::

::: {.column width="50%"}
<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
sd(c(4,7,3,5,8,6,2))
```

</details>
:::
:::::

## 5. Standard Deviation Calculation

### Interpreting Standard Deviation in Psychology

> A researcher finds that the standard deviation of reaction times in a cognitive task is `1.5` seconds. Another researcher finds a standard deviation of `0.5` seconds for a similar task.

**Question:** Which task shows more variability in participants' reaction times, and what might this indicate about task difficulty?

<details>

<summary>Solution</summary>

<p>A standard deviation of `1.5` is is larger than `0.5` is, so the first task shows **more variability**.</p>

</details>

## 6. Z-Score Calculation

### Z-Score for a Single Value

> In a personality study, the average score on an extraversion scale is `50`, with a standard deviation of `10`. One participant scores `70`.

**Question:** What is the Z-score for this participant?

::::: columns
::: {.column width="50%"}
<details>

<summary>Solution</summary>

<p>$z = \frac{X - \mu}{\sigma} = \frac{70 - 50}{10} = \frac{20}{10} = 2$</p>

</details>
:::

::: {.column width="50%"}
<details>

<summary>Code Solution</summary>

```{r}
#| echo: true

(70-50)/10
```

</details>
:::
:::::

## 7. Z-Score Calculation:

### Interpreting Z-Scores

> A participant in a self-esteem survey has a Z-score of `-1.5`. The mean self-esteem score for the survey is `40`, and the standard deviation is `5`.

**Question:** Did this participant score above or below the mean, and by how many standard deviations?

<details>

<summary>Solution</summary>

<p>A Z-score of `-1.5` is **below** the mean. In raw score terms: $X = \mu + \frac{z}{\sigma} = 40 + (-1.5)(5) = 40 - 7.5 = 32.5.$ So the participant scored `32.5`, which is `1.5` standard deviations **below** the mean.</p>

</details>

## 8. Z-Score Calculation

### Z-Score with Negative Value

> The mean stress score of a group is `25`, with a standard deviation of `4`. If a participant has a stress score of `18`...

**Question:** Calculate their Z-score and interpret its meaning in the context of stress levels.

::::: columns
::: {.column width="50%"}
<details>

<summary>Solution</summary>

<p>$z = \frac{18 - 25}{4} = \frac{-7}{4} = -1.75.$ The Z-score is $z \approx -1.75$, meaning this participant's stress score is `1.75` **standard deviations below** the group mean — they have substantially lower stress than average.</p>

</details>
:::

::: {.column width="50%"}
<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
(18-25)/4
```

</details>
:::
:::::

## 9. Z-Score Calculation

### Multiple Z-Scores

> In a clinical assessment, the average depression score is `60`, with a standard deviation of `8`. Two participants have scores of `72` and `48`.

**Question:** Calculate the Z-scores for both participants and explain what the values tell us about their relative positions in the group.

<details>

<summary>Solution</summary>

<p>
  <ul>
    <li>For the first participant (72): $z_1 = \frac{72 - 60}{8} = \frac{12}{8} = 1.5.$</li>
    <li>For the second participant (48): $z_2 = \frac{48 - 60}{8} = \frac{-12}{8} = -1.5.$</li> <li>Interpretation: the first participant is `1.5` SD **above** the mean (higher depression score than average), while the second is `1.5` SD **below** the mean (lower depression score than average). They are symmetric around the mean.</li>
  </ul>
</p>

</details>

<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
(72-60)/8;(48-60)/8
```

</details>

## 10. Normal Distribution & Areas

### Probability Beyond a Given Z-Score

> The distribution of scores on a math test is approximately normal with mean $\mu = 100$ and standard deviation $\sigma = 15$.

Question: What is the score of students who scored above $z = 1.2$?

::::: columns
::: {.column width="50%"}
<details>

<summary>Solution</summary>

<p>$z = \frac{x-\bar{x}}{s}\\ x = z\times\bar{x} + 100\\ 1.2\times15+100 = 118$</p>

</details>
:::

::: {.column width="50%"}
<details>

<summary>Code Solution</summary>

```{r}
#| echo: true
1.2*15 + 100
```

</details>
:::
:::::
