---
title: "Factorial ANOVA"
subtitle: "Lecture 14"
author: "Dave Brocker"
footer: "⬡⬢⬡⬢⬡⬢⬡⬢⬡⬢⬡⬢⬡⬢⬡"
institute: "Farmingdale State College"
execute:
  echo: true
format: 
  revealjs:
    theme: custom.scss
    scrollable: true
    incremental: true   
    touch: true
    chalkboard: true
    lightbox: true
    code-fold: true
    drop:
      engine: webr
      webr:
        packages:
         - ggplot2
         - dplyr
revealjs-plugins:
  - drop
editor: 
  markdown: 
    wrap: 72
---

## AnOVA

#### What does ANOVA do?

ANOVA compares the means of different groups to determine if they differ
significantly from one another.

-   ANOVA can examine independent variables with **more than 2 groups**.

-   ANOVA can also look at multiple independent variables.

## 2-way AnOVA

> Prof Brocker believes the key to happiness involves 2 things: drinking
> coffee and Dark. He wants to see if drinking coffee makes participants
> significantly happier than not drinking coffee. He **also** wants to
> see if watching Dark show makes them significantly happier than not
> watching it.

-   What kind of design is he using?
-   **2 x 2**

## 2-way AnOVA

We can write the design of a study analyzed using ANOVA as:

-   Number of groups "x" the Number of groups

-   Conditions in the first independent variable "x" Conditions in the
    second independent variable, etc.

-   2 x 2

## 2-way AnOVA

> Prof Brocker believes the key to happiness involves 2 things: coffee
> and Dark. He wants to see if drinking coffee makes participants
> significantly happier than not drinking coffee. He also wants to see
> if watching Dark show makes them significantly happier than not
> watching it. This is a **2 x 2** design.

-   IV1 = Number of groups in coffee: 2 (coffee or no coffee)

-   IV2 = Number of groups in Dark: 2 (Dark or No Dark)

-   DV

    -   Happiness

## 2-way Anova

What are Prof Brocker's Alternative Hypotheses?

-   H~1~: Participants who drink coffee will be significantly happier
    than those who don't.

-   H~2~: Participants who watch Dark will be significantly happier than
    those who don't.

## 2-way AnOVA

> Prof Brocker believes the key to happiness involves 2 things: coffee
> and Dark. He wants to see if drinking coffee makes participants
> significantly happier than not drinking coffee. He also wants to see
> if watching Dark show makes them significantly happier than not
> watching it. This is a **2 x 2** design.

-   Participants who drink coffee will be significantly happier than
    those who don't.

-   Participants who watch Dark will be significantly happier than those
    who don't.

-   H~3~: Participants who drink coffee while watching the Dark will be
    the happiest.

## ANOVA

### The Three Types

-   Between-subjects: \[One-Way or Factorial\]

-   Within-subjects: (Repeated Measures ANOVA)

-   Mixed-designs: Combination of Between *and* Within

## 2-way Between subjects ANOVA

-   IVs **must be** nominal/categorical

-   DV **must be** continuous

## 2-way Between subjects ANOVA

-   IVs **must be** nominal/categorical

-   DV **must be** continuous

## 2-way ANOVA

### Two types of variance:

-   Variance among all of the scores *(error variance).*

-   Variance between the groups of the first IV *(between group variance
    1).*

-   Variance between the groups of the second IV *(between group
    variance 2).*

-   Variance between the combination of the groups *(interaction
    variance).*

## 2-way ANOVA

```{r}

# Load libraries
library(dplyr)
library(ggplot2)
library(gt)
library(broom)

# Create a 2x2 grid structure for visualization
matrix_data <- expand.grid(
  Coffee = c("Yes", "No"),
  TV_Show = c("Yes", "No")
) %>%
  mutate(
    x = as.numeric(Coffee == "Yes"),
    y = as.numeric(TV_Show == "Yes"),
    fill_color = c("lightblue", "lightcoral", "lightgreen", "khaki") # Assign colors
  )

# Visualize the 2x2 matrix
fac_mat <- 
  ggplot(matrix_data, aes(xmin = x - 0.5, xmax = x + 0.5, ymin = y - 0.5, ymax = y + 0.5)) +
  geom_rect(aes(fill = fill_color), color = "black", alpha = 0.8) + # Draw rectangles
  scale_fill_identity() + # Use fill_color directly
  labs(
    title = "2x2 Factorial Matrix",
    x = "Coffee", y = "TV Show"
  ) +
  scale_x_continuous(breaks = c(0, 1), labels = c("No", "Yes"), limits = c(-0.5, 1.5)) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No", "Yes"), limits = c(-0.5, 1.5)) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

```

## 2-way ANOVA

```{r}
fac_mat + 
  annotate(
    "text", x = 1, y = 1, label = "Coffee"
  ) +
    annotate(
    "text", x = 0, y = 1, label = "Coffee"
  ) +
    annotate(
    "text", x = 0, y = 0, label = "No Coffee"
  ) +
    annotate(
    "text", x = 1, y = 0, label = "No Coffee"
  ) 
```

## 2-way ANOVA

```{r}
fac_mat + 
  annotate(
    "text", x = 1, y = 1, label = "Coffee\nDark"
  ) +
    annotate(
    "text", x = 0, y = 1, label = "Coffee\nNo Dark"
  ) +
    annotate(
    "text", x = 0, y = 0, label = "No Coffee\nDark"
  ) +
    annotate(
    "text", x = 1, y = 0, label = "No Coffee\nNo Dark"
  ) 
```

## 2-way ANOVA

To compare the means, we analyzing two types of variance:

-   Variance among all of the scores *(within group variance).*

-   Variance between the groups of the first IV *(between group variance
    1).*

-   Variance between the groups of the second IV *(between group
    variance 2).*

-   **Variance between the combination of the groups *(interaction
    variance).***

# Calculating the Fs

## Degrees of freedom

The number of observations (data points) in the data that are free to
vary when estimating a statistic.

-   I own 7 pairs of sock. I want to wear a different pair of socks
    every day.

-   How many times do I get to pick a pair of socks before I get stuck?

-   The degrees of freedom is how many times things can vary before we
    are stuck with what's left over.

## Degrees of freedom

-   Between Groups (df~BG~ ) = number of groups (k) - 1

-   Error Groups (df~Error~) = number of participants (n) - number of
    groups (k)

-   df~Total~ = n - 1

. . .

\*k is the number of groups in the independent variable.

## Degrees of freedom

-   Between Groups 1 (df~BG1~ ) = k~**1**~ - 1

-   Between Groups 2 (df~BG2~ ) = k~**2**~ - 1

-   Interaction (MS~Interaction~i) = (k~**1**~ - 1) x (k~**2**~ - 1)

-   Error (MS~Error~) = n - (k~**1**~ x k~**2**~)

-   Total = n - 1

-   \*k is the number of groups in each independent variable.

## Calculating the MS in 2-way ANOVA

-   Between Group 1 Mean Squared:

    -   $MS_{BG_1} = \frac{SS_{BG_{1}}}{k_1-1}$

-   Between Group 2 Mean Squared:

    -   $MS_{BG_{2}} = \frac{SS_{BG_{2}}}{k_2-1}$

-   Interaction Mean Squared:

    -   $MS_{Interaction} = \frac{SS_{Interaction}}{(k_1-1) \times (k_2-1)}$

-   Error Mean Squared:

    -   $MS_{Error} = \frac{SS{Error}}{n-(k_1 \times k_2)}$

## Calculating 2-way ANOVA anova

-   $F = \frac{MS_{BG_{1}}}{MS{Error}}$

-   $F = \frac{MS_{BG_{2}}}{MS{Error}}$

-   $F = \frac{MS_{Interaction}}{MS{Error}}$

## Interpreting F

::: panel-tabset
## Table

```{r}
set.seed(123)
library(dplyr)
library(afex)
library(gt)
library(broom)
library(forcats)

twoway <- 
  tibble(
    id = 1:100,
    gender = sample(c("Male", "Female"), 100, replace = TRUE),
    education_level = sample(c("High School", "Bachelor's", "Master's", "PhD"), 100, replace = TRUE)
    ) %>%
  mutate(
    # assign numeric values for education level
    edu_num = case_when(
      education_level == "High School" ~ 1,
      education_level == "Bachelor's"  ~ 2,
      education_level == "Master's"    ~ 3,
      education_level == "PhD"         ~ 4
    ),
    # introduce a systematic difference based on gender and education_level interaction
    political_interest = 5 + 
                         edu_num * 1.2 +                     # main effect of education level
                         if_else(gender == "Female", edu_num * 0.8, edu_num * -0.5) +  # interaction
                         rnorm(100, mean = 0, sd = 1)        # some noise
  ) 

# Run factorial ANOVA
aov_ez(
  "id",
  "political_interest",
  twoway,
  between = c("gender","education_level"),
  return = "aov"
) |> 
  tidy() |> 
  mutate(
    p.value = ifelse(p.value < .001, "<.001",p.value),
    statistic = ifelse(is.na(statistic),"--",statistic |> round(2)),
    p.value = ifelse(is.na(p.value),"--",p.value)
  ) |> 
  rename(
    Source = term,
    SS = sumsq,
    MS = meansq,
    `F` = statistic,
    p = p.value
  ) |> 
  gt() |> 
  fmt_auto() |> 
  tab_footnote(footnote = md("*Dependent Variable: Political Interest*")) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## Plot

```{r}

twoway$education_level <- factor(twoway$education_level, labels = c("High School","Bachelors","Masters","PhD"),ordered = TRUE)

twoway |> 
  ggplot(aes(education_level, political_interest,color = gender, group = gender)) + 
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line") + 
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) + 
  theme_minimal() + 
  labs(x = "", y = "Political Interest\n")
```
:::

-   Interpreting the Fs in 2-way ANOVA

    -   IV 1 = Gender

    -   IV 2 = Education Level

## Interpreting the Fs in 2-way ANOVA

-   F1 = Gender

-   F2 = Education Level

-   F3 = Interaction of Gender x Education Level

## Interpreting the Fs in 2-way ANOVA

::: panel-tabset
## Table

```{r}
set.seed(42)

# Create a balanced design
df <- expand.grid(
  diet_type = c("Low Carb", "Low Fat"),
  exercise_regimen = c("Cardio", "Strength"),
  id = 1:25
) %>%
  mutate(
    # simulate main effects: both diet and exercise influence weight loss
    weight_loss = case_when(
      diet_type == "Low Carb" ~ 6,
      diet_type == "Low Fat" ~ 4
    ) + case_when(
      exercise_regimen == "Cardio" ~ 5,
      exercise_regimen == "Strength" ~ 3
    ) +
    rnorm(n(), mean = 0, sd = 1.5)  # add some noise
  )

# Run the two-way ANOVA
anova_result <- aov(weight_loss ~ diet_type * exercise_regimen, data = df)
anova_result |> 
  tidy() |> 
    mutate(
    p.value = ifelse(p.value < .001, "<.001",p.value |> round(2)),
    statistic = ifelse(is.na(statistic),"--",statistic |> round(2)),
    p.value = ifelse(is.na(p.value),"--",p.value)
  ) |> 
  rename(
    Source = term,
    SS = sumsq,
    MS = meansq,
    `F` = statistic,
    p = p.value
  ) |> 
  gt() |> 
  fmt_auto() |> 
  tab_footnote(footnote = md("*Dependent Variable: Political Interest*")) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## Plot

```{r}

df |> 
  ggplot(aes(diet_type, weight_loss,color = exercise_regimen, group = exercise_regimen)) + 
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line") + 
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) + 
  theme_minimal() + 
  labs(x = "", y = "Weight Loss\n")
```
:::

## Interpreting the Fs in 2-way ANOVA

-   F1 =

-   F2 =

-   F3 =

## Interpreting the Fs in 2-way ANOVA

-   F1 = Diet Type

-   F2 = Exercise Regimen

-   F3 = Diet Type x Exercise Regimen

## 2-way AnOVA

> Prof Brocker believes the key to happiness involves 2 things: Coffee
> and Dark. He wants to see if drinking coffee makes participants
> significantly happier than not drinking coffee. He also wants to see
> if watching Dark show makes them significantly happier than not
> watching it. This is a **2 x 2** design.

He randomly assigns 100 participants to **Coffee** and **Dark** or **No
Coffee** x **Dark** or **No** **Coffee** and **No Dark**.

::: panel-tabset
## Table

```{r}
# Make the Dataset
cd <- 
  tibble(
  id = 1:400,
  coffee = rep(c("No Coffee","Yes Coffee"), each = 200),
  dark = rep(c("No Dark","Yes Dark"), 200),
  happiness = c(rnorm(100,4,2) |> round(0),
                rnorm(100,7,1) |> round(0),
                rnorm(100,6,2) |> round(0),
                rnorm(100,8.5,1) |> round(0)
                )
  ) 

# Calculate ANOVA
aov(happiness ~ coffee * dark, data = cd) |> 
  tidy() |> 
  mutate(
    Source = term,
    SS = sumsq, 
    MS = meansq,
    `F` = ifelse(is.na(statistic),"--",statistic |> round(3)),
    p = ifelse(p.value < .05, "<.05",p.value |> round(3)),
    p = ifelse(is.na(p),"--",p)
  ) |> 
  select(Source,df,SS,MS,`F`,p) |> 
  gt() |> 
  fmt_auto() |> 
  cols_align(align = "center",
             columns = !Source) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |> 
  tab_footnote(footnote = "Dependent Variable: Happiness")
```

## Plot

```{r}

cd |> 
  ggplot(aes(coffee, happiness,color = dark)) + 
  geom_jitter(alpha = .2) + 
  stat_summary(
    fun.data = "mean_se",
    geom = "errorbar",
    width = .2
  ) + 
  stat_summary(
    fun.data = "mean_se",
    geom = "line",
    group = 1
  ) + 
  theme_minimal() + 
  labs(
    x = "Coffee Drank\n",
    y = "Level of Happiness\n",
    color = "black"
  ) + 
  theme(
    axis.text = element_text(color = "black")
  ) 
```
:::

## Calculating the Fs in 2-way ANOVA

```{r}
tribble(
~Source, ~SS, ~df, ~MS, ~F,
"Coffee",432, "","","",
"Dark",486,"","","",
"Coffee x Dark",540,"","","",
"Error",5184,"","",""
) |> 
  gt() |> 
    cols_align(align = "center",
             columns = SS:F)
```

## Calculating the Fs in 2-way ANOVA

```{r}
tribble(
~Source, ~SS, ~df, ~MS, ~F,
"Coffee",432, 1,432,"8",
"Dark",486,1,486,"9",
"Coffee x Dark",540,1,540,"10",
"Error",5184,96,54,""
) |> 
  gt() |> 
    cols_align(align = "center",
             columns = SS:F)
```

## 2-way AnOVA

> Prof. Brocker recruit 500 children on Halloween. He assigns half to
> wear a costume with a mask, and the other half to wear a costume
> without a mask. He assigns participants to trick-or-treat alone or in
> groups. Then he measures how much candy each child collects while
> trick-or-treating.

::: panel-tabset
## Table

```{r}
set.seed(1031)  # Halloween-ish number for reproducibility

# Create a balanced factorial design
new_df <- expand.grid(
  mask_condition = c("Mask", "No Mask"),
  group_condition = c("Alone", "Group"),
  n = 1:125  # 125 per cell → 500 total
) %>%
  mutate(
    # Simulate DV with higher values for Mask + Alone
    candy_collected = case_when(
      mask_condition == "Mask" & group_condition == "Alone" ~ rnorm(n(), mean = 50, sd = 5),
      mask_condition == "Mask" & group_condition == "Group" ~ rnorm(n(), mean = 44, sd = 5),
      mask_condition == "No Mask" & group_condition == "Alone" ~ rnorm(n(), mean = 42, sd = 5),
      mask_condition == "No Mask" & group_condition == "Group" ~ rnorm(n(), mean = 40, sd = 5)
    )
  )

# Run a two-way ANOVA
model <- aov(candy_collected ~ mask_condition * group_condition, data = new_df)

model |> 
  tidy() |> 
    mutate(
    p.value = ifelse(p.value < .001, "<.001",p.value),
    statistic = ifelse(is.na(statistic),"--",statistic |> round(2)),
    p.value = ifelse(is.na(p.value),"--",p.value)
  ) |> 
  rename(
    Source = term,
    SS = sumsq,
    MS = meansq,
    `F` = statistic,
    p = p.value
  ) |> 
  gt() |> 
  fmt_auto() |> 
  tab_footnote(footnote = md("*Dependent Variable: Candy Collected*")) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## Plot

```{r}

new_df |> 
  ggplot(aes(mask_condition, candy_collected,color = group_condition, group = group_condition)) + 
  geom_jitter(alpha = .2) + 
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line") + 
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) + 
  theme_minimal() + 
  labs(x = "", y = "Candy Collected\n")
```
:::

## Calculating the Fs in 2-way ANOVA

```{r}
tribble(
~Source, ~SS, ~df, ~MS, ~F,
"IV1",252, "","","",
"IV2",441,"","","",
"Interaction",882,"","","",
"Error",10416,"","",""
) |> 
  gt() |> 
  cols_align(align = "center",
             columns = SS:F)
```

## Calculating the Fs in 2-way ANOVA

```{r}
tribble(
~Source, ~SS, ~df, ~MS, ~F,
"IV1",252, "1","252","12",
"IV2",441,"1","441","21",
"Interaction",882,"1","882","42",
"Error",10416,"496","21",""
) |> 
  gt() |> 
  cols_align(align = "center",
             columns = SS:F)
```

## Calculating 2-way ANOVA anova

-   $F = \frac{MS_{BG_{1}}}{MS{Error}}$

-   $F = \frac{MS_{BG_{1}}}{MS{Error}}$

-   $F = \frac{MS_{Interaction}}{MS{Error}}$

-   \*All MS are divided by $MS_{Error}$

## Reporting F

## Reporting F

If asked to report findings in terms of the Null Hypothesis (H~0~), you
should report findings as:

-   Reject H~0~, or

-   Fail to Reject H~0~

## Reporting F

If asked to report findings in general or for publication, you need
**EACH** F-value:

-   F*(df~BG~, df~W~ )*

    -   = F-value

        -   Corresponding p-value

            -   *For significant results: Means and standard deviations
                of each group*

## Reporting F

> The group that watched Dark (M=8.43, s=1.02) reported significantly
> more happiness compared to their peers in the control group who
> watched Jeopardy (M=6.12, s=0.98), *F(1,496) = 7.12, p \< 0.05*.

## Reporting F

Results IV~1~:

::: fragment
> **If significant**: The group that watched Dark (M=8.43, s=1.02)
> reported significantly more happiness compared to their peers in the
> control group who watched Jeopardy (M=6.12, s=0.98), *F (1, 496) =
> 7.12, p \< 0.05*.
:::

::: fragment
> **If NOT significant**: There was not a significant difference in
> happiness between the groups, F (1, 498) = 1.02, p = 0.07.
:::

## Reporting F

Results IV2:

> **If significant**: The group that drank coffee (M=7.35, s=1.01)
> reported significantly more happiness compared to their peers in the
> control group who drank decaf (M=4.21, s=0.99), *F (1, 496) = 9.12, p
> \< 0.05*.

> **If NOT significant**: There was not a significant difference in
> happiness between the groups, *F (1, 498) = 1.02, p = 0.07*.

## Reporting F

Results Interaction:

> **If significant**: The group that drank coffee AND watched Dark
> (M=7.35, s=1.01) reported significantly more happiness compared to the
> other groups (M=4.21, s=0.99), F *(1, 496) = 11.14, p \< 0.05.*

> **If NOT significant**: There was not a significant difference in
> happiness between the groups, *F (1, 498) = 1.02, p = 0.07*.

<div>

```{r}
library(dplyr)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Create levels for the independent variables
coffee <- c("Yes", "No")
tv_show <- c("Yes", "No")

# Create a data frame with all combinations of coffee and tv_show
factorial_design <- expand.grid(Coffee = coffee, TV_Show = tv_show)

# Simulate happiness scores for each group
# Adjust means to reflect interaction effects
n_per_group <- 30 # Number of participants per condition
data <- factorial_design[rep(1:nrow(factorial_design), each = n_per_group), ]
data$Happiness <- NA

# Define mean happiness scores for each condition
means <- c(
  "Yes_Yes" = 8,  # Coffee + TV Show
  "Yes_No" = 6,   # Coffee + No TV Show
  "No_Yes" = 5,   # No Coffee + TV Show
  "No_No" = 3     # No Coffee + No TV Show
)

# Assign happiness scores with some random noise
data$Happiness <- rnorm(
  n = nrow(data),
  mean = means[paste(data$Coffee, data$TV_Show, sep = "_")],
  sd = 1.5 # Standard deviation for noise
)

# Model
fac_mod <- 
  aov(Happiness ~ Coffee * TV_Show, data  = data)

data |> 
  ggplot(aes(Coffee, Happiness, color = TV_Show)) + 
  geom_jitter(alpha = .4) + 
   stat_summary(
    fun.data = "mean_se",
    geom = "line",
    color = "black",
    aes(group = TV_Show,
        lty = TV_Show)
  ) + 
    stat_summary(
    fun.data = "mean_se",
    geom = "pointrange",
    aes(color = TV_Show)
  ) + 
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  )
  

plot(TukeyHSD(fac_mod))

```

</div>
