---
title: "Data Visualization"
subtitle: "PSY 348: Lecture 2"
author: 
  - name: "Dave Brocker"
    email: brockeda@farmingdale.edu
institute: "Farmingdale State College"
footer: "PSY 348 - Statistics for Psychology"
format:
  revealjs:
    menu: 
      numbers: true
    slide-number: c/t
    preview-links: auto
    theme: custom.scss
    incremental: true
    touch: true
    code-fold: true
    drop:
      engine: webr
      webr:
        packages:
         - ggplot2
         - dplyr
revealjs-plugins:
  - drop
---

## From "Stats Is Everywhere" to "Let's Actually See It"

Last time: statistics isn't just math — it's how you make sense of decisions you already make every day.

::: highlight-last
- A number by itself is hard to *feel* — "the average score was 78" doesn't tell you much on its own

- A well-made graph turns a pile of numbers into something you can understand at a glance

- This lecture is about **choosing and building the right picture** for your data
:::

## Why Visualize Data?

Consider these two ways of showing the same information:

::: fragment
**Option A:** "The average exam score was 78 (SD = 8)."
:::

::: fragment
**Option B:** *A histogram showing the actual shape of the scores — most students clustered in the 70s, with a small group who struggled and a small group who excelled.*
:::

::: fragment
Both are "true." Only one of them lets you *see* the students who didn't fit the average.
:::

## Same Numbers, Very Different Pictures

### Anscombe's Quartet

These four datasets have (essentially) identical descriptive statistics:

```{r}
#| echo: false
library(dplyr)
library(gt)

data(anscombe)

tab <- tibble(
  Set = c("I", "II", "III", "IV"),
  `Mean (x)` = sapply(1:4, function(i) round(mean(anscombe[[paste0("x", i)]]), 2)),
  `Mean (y)` = sapply(1:4, function(i) round(mean(anscombe[[paste0("y", i)]]), 2)),
  `SD (x)` = sapply(1:4, function(i) round(sd(anscombe[[paste0("x", i)]]), 2)),
  `SD (y)` = sapply(1:4, function(i) round(sd(anscombe[[paste0("y", i)]]), 2)),
  Correlation = sapply(1:4, function(i) round(cor(anscombe[[paste0("x", i)]], anscombe[[paste0("y", i)]]), 2))
)

tab |>
  gt(rowname_col = "Set") |>
  tab_header(title = "Anscombe's Quartet: Identical Summary Statistics") |>
  tab_source_note("Anscombe, F. J. (1973). Graphs in statistical analysis.") |>
  opt_row_striping() |>
  tab_options(table.font.size = 24, table.width = pct(90))
```

## Same Numbers, Very Different Pictures

### But Look What They Actually Look Like

```{r}
#| fig-align: center
#| code-summary: "Build the 4-panel Anscombe scatterplot"
library(dplyr)
library(ggplot2)

anscombe_long <- bind_rows(              # stack all four sets into one long tibble
  tibble(set = "Set I",   x = anscombe$x1, y = anscombe$y1),
  tibble(set = "Set II",  x = anscombe$x2, y = anscombe$y2),
  tibble(set = "Set III", x = anscombe$x3, y = anscombe$y3),
  tibble(set = "Set IV",  x = anscombe$x4, y = anscombe$y4)
)

ggplot(anscombe_long, aes(x = x, y = y)) +
  geom_point(size = 3, color = "#5C715E") +                    # the actual data points
  geom_smooth(method = "lm", se = FALSE, color = "#8FAF9E") +   # a straight best-fit line for each set
  facet_wrap(~set) +                                            # one small panel per dataset
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank()) +                   # drop the finer gridlines, keep the major ones
  labs(
    title = "Anscombe's Quartet",
    subtitle = "Same mean, SD, and correlation — four very different pictures",
    x = "x", y = "y"
  )
```

## The Takeaway

::: highlight-last
- Summary statistics (mean, SD, correlation) can hide the actual **shape** of your data

- Outliers, clusters, and non-linear patterns are often invisible in a table of numbers

- **Always look at the picture** — it's not optional, it's part of the analysis
:::

## Choosing the Right Chart

```{r}
#| echo: false
library(dplyr)
library(gt)

chart_guide <- tibble(
  `What are you showing?` = c(
    "One categorical variable (counts per group)",
    "One continuous variable (its shape/spread)",
    "Two continuous variables (a relationship)",
    "A continuous variable, compared across groups",
    "Change over time"
  ),
  `Best chart` = c(
    "Bar chart",
    "Histogram or density plot",
    "Scatterplot",
    "Box plot",
    "Line chart"
  )
)

chart_guide |>
  gt() |>
  tab_header(title = "A Quick Decision Guide") |>
  opt_row_striping() |>
  tab_options(table.font.size = 26, table.width = pct(90)) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## The `ggplot2` Recipe

Every `ggplot2` chart follows the same three-ingredient recipe — we'll come back to this constantly this semester:

::: highlight-last
1.  **Data** — which tibble/data frame are you plotting?

2.  **Aesthetics** (`aes()`) — which columns map to x, y, color, etc.?

3.  **Geometry** (`geom_*()`) — what *shape* represents your data (bars, points, lines...)?
:::

```{r}
#| eval: false
#| echo: true
ggplot(data, aes(x = ..., y = ...)) +  # 1. data + 2. aesthetics (which columns map to x/y)
  geom_something()                      # 3. geometry (the shape that represents your data)
```

We'll build up from this simple recipe all semester — no need to memorize every option today.

## Setting Up Our Example Data

### One Dataset, Several Questions

```{r}
#| code-summary: "Build a fake 'students' dataset we'll reuse all lecture"
library(dplyr)
library(gt)

set.seed(2)  # makes the "random" data reproducible every time this renders
students <- tibble(
  student = 1:60,                                             # just an ID number
  study_time = sample(c("Morning", "Afternoon", "Evening", "Night"), 60,
                       replace = TRUE, prob = c(0.15, 0.30, 0.35, 0.20)),   # categorical variable
  study_hours = pmax(round(rnorm(60, mean = 6, sd = 2), 1), 0.5),          # continuous variable, floor at 0.5
  study_method = rep(c("Flashcards", "Practice Tests"), each = 30),       # two even groups
  exam_score = pmin(round(55 + 2.5 * study_hours +                        # score rises with study hours...
                     ifelse(study_method == "Practice Tests", 4, 0) +     # ...gets a small bump from practice tests...
                     rnorm(60, 0, 5)), 100)                               # ...plus random noise, capped at 100
)

students |>
  slice_head(n = 5) |>       # just preview the first 5 rows
  gt() |>
  opt_row_striping() |>
  tab_options(table.font.size = 22, table.width = pct(80))
```

## Bar Charts

### One Categorical Variable

```{r}
#| fig-align: center
#| fig-height: 4.5
#| code-summary: "Bar chart: counts by study time"
library(ggplot2)

students |>
  count(study_time) |>                                     # tally how many students fall in each category
  ggplot(aes(x = study_time, y = n, fill = study_time)) +   # x = category, y = count, color by category
  geom_col() +                                              # bars sized to the y value
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),    # drop the finer gridlines
    panel.grid.major.x = element_blank()   # vertical gridlines add nothing when x is categorical
  ) +
  labs(
    title = "When Do Students Prefer to Study?",
    x = "", y = "Number of Students"
  )
```

## Histograms

### One Continuous Variable

```{r}
#| fig-align: center
#| fig-height: 4.5
#| code-summary: "Histogram: distribution of exam scores"
library(ggplot2)

students |>
  ggplot(aes(x = exam_score)) +
  geom_histogram(binwidth = 5, fill = "#8FAF9E", color = "white") +  # bin width of 5 points per bar
  theme_minimal(base_size = 18) +
  theme(panel.grid.minor = element_blank()) +   # keep the major gridlines -- useful for reading bar heights
  labs(
    title = "Distribution of Exam Scores",
    x = "Exam Score", y = "Number of Students"
  )
```

## Scatterplots

### Two Continuous Variables

```{r}
#| fig-align: center
#| fig-height: 4.5
#| code-summary: "Scatterplot: study hours vs. exam score"
library(ggplot2)

students |>
  ggplot(aes(x = study_hours, y = exam_score)) +
  geom_point(size = 3, color = "#5C715E", alpha = 0.7) +             # one dot per student
  geom_smooth(method = "lm", se = FALSE, color = "#8FAF9E") +        # a straight best-fit line through the cloud
  theme_minimal(base_size = 18) +
  theme(panel.grid.minor = element_blank()) +   # gridlines here help you read off approximate values
  labs(
    title = "Study Hours vs. Exam Score",
    x = "Study Hours (per week)", y = "Exam Score"
  )
```

::: fragment
We'll come back to this exact question — "how strongly are two variables related?" — when we get to correlation.
:::

## Box Plots

### Comparing a Continuous Variable Across Groups

```{r}
#| fig-align: center
#| fig-height: 4.5
#| code-summary: "Box plot: exam scores by study method"
library(ggplot2)

students |>
  ggplot(aes(x = study_method, y = exam_score, fill = study_method)) +
  geom_boxplot(alpha = 0.6) +    # median, IQR box, and whiskers for each group
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()   # x is categorical here too -- vertical lines aren't meaningful
  ) +
  labs(
    title = "Exam Score by Study Method",
    x = "", y = "Exam Score"
  )
```

::: fragment
This is also a preview — "is the difference between these two groups big enough to matter?" is exactly what a later lecture's t-test answers.
:::

## How Graphs Can Mislead

### Same Data, Different Story

```{r}
#| fig-align: center
#| fig-height: 4.5
#| code-summary: "Two bar charts of the same numbers, different y-axis range"
library(ggplot2)
library(patchwork)   # lets us place p1 and p2 side by side with `p1 + p2`
library(dplyr)

group_means <- students |>
  group_by(study_method) |>
  summarize(mean_score = mean(exam_score))   # one average score per study method

p1 <- ggplot(group_means, aes(x = study_method, y = mean_score, fill = study_method)) +
  geom_col() +
  coord_cartesian(ylim = c(0, 100)) +    # y-axis starts at 0 -- the honest version
  theme_minimal(base_size = 14) +
  theme(legend.position = "none", panel.grid.minor = element_blank(), panel.grid.major.x = element_blank()) +
  labs(title = "Full Scale (0-100)", x = "", y = "Mean Exam Score")

p2 <- ggplot(group_means, aes(x = study_method, y = mean_score, fill = study_method)) +
  geom_col() +
  coord_cartesian(ylim = c(65, 75)) +    # y-axis zoomed in -- makes the gap look huge
  theme_minimal(base_size = 14) +
  theme(legend.position = "none", panel.grid.minor = element_blank(), panel.grid.major.x = element_blank()) +
  labs(title = "Truncated Y-Axis", x = "", y = "Mean Exam Score")

p1 + p2   # patchwork's `+` places the two plots next to each other
```

::: fragment
Same exact numbers. The right panel makes a small difference look dramatic — always check where the y-axis starts.
:::

## Building a Good Graph

::: highlight-last
- **Title** — what am I looking at?

- **Labeled axes** — with units, if relevant

- **An honest scale** — start bar charts at 0; don't cherry-pick a range to exaggerate a difference

- **Only the elements that add information** — no 3D effects, no unnecessary gridlines, no clutter
:::

## Recap: What You Should Walk Away With

::: highlight-last
- A summary statistic can hide the actual shape of your data — that's why we plot it

- You can match a research question to the right chart: one variable vs. two, categorical vs. continuous, one group vs. several

- You've seen the basic `ggplot2` recipe: **data + aesthetics + geometry**

- You know how to spot (and avoid) a few common ways graphs mislead
:::

Next time: back to describing that shape with numbers — measures of central tendency.