---
title: "Dispersion"
subtitle: "Lecture 4"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Dispersion | ⬡⬢⬡⬢"
format: 
  revealjs:
    theme: custom.scss
    incremental: true   
    touch: true
    chalkboard: true
---

## Descriptive Statistics

### The Old and the New

::: {.incremental .highlight-last}
-   $x$ 

-   $\bar{x}$ 

-   $\sum(x)$ 

-   $x^2$ 
:::

## Descriptive Statistics

### The Old and the New

::: {.incremental .highlight-last}
-   $x$ = refers to the value for 1 person (1 datapoint) in the sample

-   $\bar{x}$ = refers to the mean or average value of X in the sample

-   $\sum(x)$ = refers to the sum of all x values (sum of all the datapoints)

-   $x^2$ = refers to the squared value of x (x multiplied by x)
:::

## Descriptive Statistics

### Examples

::: {.incremental .highlight-last}
-   `data = {1,4,3,2,4,6}`

-   $x$ = $x = 1 | 4 | 3 | 2 | 4 | 6$

-   $\bar{x}$ = $3.33\bar{3}$

-   $\sum(x)$ = $1 + 4 + 3 + 2 + 4 + 6 = 20$

-   $x^2$ = $1 | 16 | 9 | 4 | 16 | 36$
:::

## Descriptive Statistics

### Describe the characteristics of a sample in terms of:

::: {.incremental .highlight-last}
-   Central tendency

-   Dispersion (aka Variability aka Variance)

    -   Dispersion: the action or process of distributing things or people over a wide area.
:::

## Central tendency: Mean

### What is it, when should we use it, and how do we calculate it?

::: {.incremental .highlight-last}
-   Mean
    -   Average of all $x$ values
    -   Normal Distribution
    -   $\frac{\sum(x)}{n}$
:::

## Central tendency: Median

### What is it, when should we use it, and how do we calculate it?

::: {.incremental .highlight-last}
-   Median
    -   Middle value in **sorted** data
    -   Skewed distributions
    -   $Med(X) = \frac{n + 1}{2}$
:::

## Central tendency: Mode

### What is it, when should we use it, and how do we calculate it?

::: {.incremental .highlight-last}
-   Mode
    -   Most common value
    -   Bi\|Tri-Modal Distributions
    -   Most common value
:::

## Central Tendency

### What do measures of Central Tendency tell you?

::: {.incremental .highlight-last}
-   The mid-point in the data

-   The answer most of the participants gave.
:::

## Central Tendency

### What do measures of Central Tendency tell you?

```{r}
#| fig-align: center
library(dplyr)
library(ggplot2)
library(plotly)
set.seed(348)

ex <- 
tibble(
  student = 1:30,
  score = rnorm(30,80,10)
) |> 
  ggplot(aes(student,score)) + 
  geom_point() +
  theme_minimal() +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  ) + 
  geom_hline(aes(yintercept = mean(score)),
             color = "maroon",
             lty = "dashed") 

ggplotly(ex)

```

## Central Tendency

### What do measures of Central Tendency tell you?

```{r}
#| fig-align: center
set.seed(348)

ex2 <- 
tibble(
  student = 1:30,
  score = rnorm(30,80,10)
) |> 
  ggplot(aes(student,score)) + 
  geom_point() +
  theme_minimal() +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  ) + 
  geom_hline(aes(yintercept = mean(score)),
             color = "maroon",
              linewidth = 1.5) +
  geom_segment(aes(x = student, xend = student,
                  y = mean(score),
                  yend = score)
              )
ggplotly(ex2)

```

## Measures of Dispersion

### What does dispersion mean?

::: {.incremental .highlight-last}
-   Spread-out-ness

-   How much the participants' responses differ from one another

-   Variance
:::

## Measures of Dispersion

### Tells us about the spread-out-ness of the data, but what does that *actually* mean?

::: fragment
```{r}
#| fig-align: center

library(patchwork)

exs <- 
  tibble(
    student = 1:30,
    exam1 = rnorm(30,85,1),
    exam2 = rnorm(30,70,15),
    exam3 = rnorm(30,80,4)
)

p1 <- 
  exs |> 
    ggplot(aes(student,exam1)) + 
    geom_point() +
    theme_minimal() +
    ylim(0,100) +
    geom_hline(aes(yintercept = mean(exam1),color = "maroon")) +
labs(
  x = "\nStudent",
  y = "Exam Score\n"
  )

p2 <- 
  exs |> 
    ggplot(aes(student,exam2)) + 
    geom_point() +
    theme_minimal()+
    ylim(0,100) +
    geom_hline(aes(yintercept = mean(exam2),color = "maroon")) +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  )

p3 <- 
  exs |> 
    ggplot(aes(student,exam3)) + 
    geom_point() +
    theme_minimal() +
    ylim(0,100) +
    geom_hline(aes(yintercept = mean(exam3)),color = "maroon") +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  )

p1 + p2 + p3

```
:::

## Measures of Dispersion

### Measures of dispersion tell us about the spread-out-ness of the data, but what does that *actually* mean?

::: {.incremental .highlight-last}
-   If I am measuring , *music popularity* what would low dispersion tell me?
-   If I am measuring *music popularity*, what would high dispersion tell me?
:::

## Measures of Dispersion

### Measures of dispersion tell us about consensus.

::: {.incremental .highlight-last}
-   Did participants give similar answers?

    -   `{1,2,2,1,2,2,3,2,1}`

-   Did participants give wildly different answers?

    -   `{1,3,4,5,8,9,21,33}`
:::

## Dispersion

### Measures of dispersion tell us about consensus.

::: {.incremental .highlight-last}
-   Data should have natural dispersion.

-   If everyone gives a similar answer, it's harder to analyze difference.
:::

## Calculating Dispersion

### How do you think we should calculate dispersion?

::: {.incremental .highlight-last}
-   ON AVERAGE, how far is each X-VALUE from the MIDPOINT.

    -   Calculate how far each X-VALUE is from the MIDPOINT.

    -   Take the AVERAGE of those distances.
:::

## Calculating Dispersion

### How do you think we should calculate dispersion?

In psychology, the mid-point we use will be the [***mean***]{.underline}.

## Calculating Dispersion

### Acquire Data

```{r}
library(huxtable)

data.frame(
  `Participant ID` = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2)
) |> hux() |> 
  theme_article() |> 
  set_width(.5) |> 
  set_align("left")
```

## Deviation Scores

### Calculate how far each X-VALUE is from the MEAN.

::: {.incremental .highlight-last}

::: fragment
Step 1. Examine Data
:::

-   `data = {1,3,2,2,3,2,1,3,1,2}`

::: fragment
Step 2. Calculate Mean
:::

-   $\bar{x} = \frac{\sum(x)}{n} = 1 + 3 + 2...$

-   $\bar{x} = \frac{20}{10} = 2$
:::

## Deviation Scores

### Calculate how far each X-VALUE is from the MEAN.

::: {.incremental .highlight-last}
Step 2: Subtract the mean from each X-value.

-   $x - \bar{x}$

-   $(1-2) + (3-2) + (2-2) + (2-2)...$

-   $(-1) + 1 + 0 + 0...$
:::

## Calculating Dispersion

### Visual

```{r}
library(dplyr)
tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X)
) |> hux() |> 
  theme_article() |> 
  set_align("center") |> 
  set_width(.5) |> 
  set_height(.5)
```

## Calculating Dispersion

### Visual

```{r}
tibble(
  ParticipantID = paste0(1:10),
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X)
) |> 
  ggplot(aes(ParticipantID,X)) + 
  geom_point(color = "darkgreen") + 
  geom_hline(aes(yintercept = mean(X)),color = "red") + 
  theme_minimal() +
  geom_segment(aes(x = ParticipantID, xend = ParticipantID,
                  y = mean(X),
                  yend = X)
              ) +
  labs(
  x = "\nParticipant ID",
  y = "X Value\n"
  )
```

## Calculating Dispersion

### Visual

```{r}
tibble(
  ParticipantID = paste0(1:10),
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X)
) |> 
  ggplot(aes(ParticipantID,X)) + 
  geom_hline(aes(yintercept = mean(`x-M`)),lty = "dashed") + 
  geom_segment(aes(x = ParticipantID, xend = ParticipantID,
                  y = mean(`x-M`),
                  yend = `x-M`)
              ) +
  geom_point(aes(ParticipantID,`x-M`, color = "black")) +
  theme_minimal() +
  labs(
  x = "\nParticipant ID",
  y = "X Value\n"
  )

```
## Calculating Dispersion

### Visual

::: {.incremental .highlight-last}
-   The average of the deviation scores will always = 0.

-   On average, the x-values = x.

-   So the average distance of x from x is 0.
:::

## Calculating Dispersion

### How do you think we should calculate dispersion?

::: {.incremental .highlight-last}
1.  Calculate the mean.
2.  Calculate the deviation scores.
3.  What can we do to the deviation scores to prevent them from adding up to 0?
:::

::: {.incremental .highlight-last}
## What is the Mean of the Deviation Scores?

### Stop Being So Negative!

-   The problem is the negative numbers!

-   How do we get rid of negative numbers

-   $(x-\bar{x})^2$
:::

## Calculating Dispersion

### Stop Being So Negative!

```{r}
tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) |> 
  # Bold Last Column
  hux() |> 
  theme_article() |> 
  set_align("center") |> 
  set_width(.5)
```

## Calculating Dispersion

### First 3 Steps

::: {.incremental .highlight-last}
-   Step 1: Calculate the mean.

-   Step 2: Subtract the mean from each X-value.

-   Step 3: Square each deviation score.
:::

## Calculating Dispersion

### Now what: Calculate the average squared deviation.

::: {.incremental .highlight-last}
-   Step 4: Add up the Squared Deviations.

-   $1+1+0+1...$
:::

## Calculating Dispersion

### Now what: Calculate the average squared deviation.

```{r}
tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) |> 
  hux() |> 
  theme_article()
```

## Calculating Dispersion
### Divide by n-1 to Estimate

::: {.incremental .highlight-last}

Step 5: Divide by (n-1).

-   (n-1) refers to the *degrees of freedom*... don't worry about it for now.

-   $\frac{\sum (x-\bar{x})^2}{n-1}$

-   Variance

    -   The fact or quality of being different, divergent, or inconsistent.
    
:::

## Calculating Dispersion

### How do think we should calculate dispersion?

-   When we squared the deviation scores in Step 3, we inflated the deviation.

    -   We have to undo that inflation.

    -   How do you undo squaring?

    -   We take the square root.

## Calculating Dispersion
### Inflation

```{r}
tibble(
  ParticipantID = 1:6,
  X = runif(6,30,50),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) |> 
  hux() |> 
  theme_article()
```

## Calculating Dispersion

Step 6: Take the square root.

-   $\frac{\sum(x-\bar{x})^2}{n-1} = \sqrt{.667} = .816$

## Calculating standard deviation
### Step-by-Step

::: {.incremental .highlight-last}

1.  Mean: Calculate the mean.

    -   $\bar{x}$

2.  Deviation scores: Subtract the mean from each x-value.

    -   $(x-\bar{x})$

3.  Squared deviations: Square each deviation score.

    -   $(x-\bar{x})^2$

4.  Sum of squares: Add up the squared deviations.

    -   $\sum(x-\bar{x})^2$
    
:::

## Calculating standard deviation
### Step-by-Step

::: {.incremental .highlight-last}

5.  Variance: Divide the sum of squares by (n-1).

    -   $\frac{\sum(x-\bar{x})^2}{n-1}$

6.  Standard Deviation: Take the square root of the variance.

    -   $\sqrt{\frac{\sum(x-\bar{x})^2}{n-1}}$

:::

## Formulas for Dispersion

### Variance and Standard Deviation

::: {.callout-important}
$$
s^2 = \frac{\sum(x-\bar{x})^2}{N-1}
$$

:::

::: {.callout-important}

$$
s = \sqrt{\frac{\sum(x-\bar{x})^2}{N-1}}
$$

:::

## Formulas for Dispersion
### Standard Deviation

-   Variance is a measure of dispersion.

    -   It's not helpful, because it's inflated (from squaring the deviation scores).

-   Standard deviation is a better measure of dispersion, because it is standardized.

    -   This means that a standard deviation of 1 means a distance of 1 on the scale used to measure x.

    -   HYPE scale: 1 to 10: standard deviation of 1 means the distance from one rating to the next, from 1 to 2.

    -   If the mean = 5, SD = 1, most scores fall between 4 and 6.

## Variance & Standard deviation

-   The Variance is the standard deviation squared: s²

-   The Standard Deviation is the square root of the variance: s
