---
title: "Standard Normal Distribution"
subtitle: "PSY 348: Lecture 4"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Dispersion | ⬡⬢⬡⬢"
filters: 
  - linkate
format: 
  revealjs:
    theme: custom.scss
    chalkboard: true
    slide-number: c/t
    preview-links: true
    incremental: true   
    touch: true
    width: 1600
    height: 900
    lightbox: true
    menu:
      side: right
      width: wide
---

## Descriptive Statistics

### The Old and the New

-   $x$

-   $\bar{x}$

-   $\sum(x)$

-   $x^2$

## Descriptive Statistics

### The Old and the New

::: {.incremental .highlight-last}
-   $x$ = refers to the value for 1 person (1 datapoint) in the sample

-   $\bar{x}$ = refers to the mean or average value of X in the sample

-   $\sum(x)$ = refers to the sum of all x values (sum of all the datapoints)

-   $x^2$ = refers to the squared value of x (x multiplied by x)
:::

## Descriptive Statistics

### Examples

-   $data = \{1,4,3,2,4,6\}$

-   $x$ = $1 | 4 | 3 | 2 | 4 | 6$

-   $\bar{x}$ = $3.33\bar{3}$

-   $\sum(x)$ = $1 + 4 + 3 + 2 + 4 + 6 = 20$

-   $x^2$ = $1 | 16 | 9 | 4 | 16 | 36$

## Descriptive Statistics

### Describe the characteristics of a sample in terms of:

-   Central tendency

    -   The mid-point of the data

    -   The most typical/frequent answers

-   Dispersion (aka Variability aka Variance)

    -   Dispersion: the action or process of distributing things or people over a wide area.

## Central tendency: Mean

### What is it, when should we use it, and how do we calculate it?

-   Mean
    -   Average of all $x$ values
    -   Normal Distribution
    -   Formula: $\frac{\sum(x)}{n}$

## Central tendency: Median

### What is it, when should we use it, and how do we calculate it?

-   Median
    -   Middle value in **sorted** data
    -   Skewed distributions
        -   Is the data pulled to the left or the right?
        -   $Med(X) = \frac{n + 1}{2}$

## Central tendency: Mode

### What is it, when should we use it, and how do we calculate it?

-   Mode
    -   Most common value
    -   Bi/Tri-Modal Distributions
    -   Most common value(s)
        -   Shows us that data is clustered

## Central Tendency

### What do measures of Central Tendency tell you?

::::: columns
::: {.column width="50%"}
-   The mid-point in the data

-   The answer most of the participants gave.
:::

::: {.column width="50%"}
```{r}
library(dplyr)
library(ggplot2)
set.seed(348)

tibble(
  student = 1:30,
  score = rnorm(30,80,10)
) |> 
  ggplot(aes(student,score)) + 
  geom_point() +
  theme_minimal() +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  ) + 
  geom_hline(aes(yintercept = mean(score)),
             color = "maroon",
             lty = "dashed") 
```
:::
:::::

## Central Tendency

### What do measures of Central Tendency tell you?

```{r}
set.seed(348)

tibble(
  student = 1:30,
  score = rnorm(30,80,10)
) |> 
  ggplot(aes(student,score)) + 
  geom_point() +
  theme_minimal() +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  ) + 
  geom_hline(aes(yintercept = mean(score)),
             color = "maroon",
              linewidth = 1.5) +
  geom_segment(aes(x = student, xend = student,
                  y = mean(score),
                  yend = score)
              )
```

## Measures of Dispersion

### What does dispersion mean?

-   Spread-out-ness

-   How much the participants' responses differ from one another

-   Variance

## Measures of Dispersion

### Tells us about the spread-out-ness of the data, but what does that *actually* mean?

::: fragment
```{r}
#| fig-align: center

library(patchwork)

exs <- 
  tibble(
    student = 1:30,
    exam1 = rnorm(30,85,1),
    exam2 = rnorm(30,70,15),
    exam3 = rnorm(30,80,4)
)

p1 <- 
  exs |> 
    ggplot(aes(student,exam1)) + 
    geom_point() +
    theme_minimal() +
    geom_hline(aes(yintercept = mean(exam1),color = "maroon")) +
labs(
  x = "\nStudent",
  y = "Exam Score\n"
  ) + theme(legend.position = "none")
p2 <- 
  exs |> 
    ggplot(aes(student,exam2)) + 
    geom_point() +
    theme_minimal()+
    geom_hline(aes(yintercept = mean(exam2),color = "maroon")) +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  )+ theme(legend.position = "none")

p3 <- 
  exs |> 
    ggplot(aes(student,exam3)) + 
    geom_point() +
    theme_minimal() +
    geom_hline(aes(yintercept = mean(exam3)),color = "maroon") +
  labs(
  x = "\nStudent",
  y = "Exam Score\n"
  ) + theme(legend.position = "none")

p1 + p2 + p3

```
:::

## Measures of Dispersion

### Measures of dispersion tell us about the spread-out-ness of the data, but what does that *actually* mean?

-   If I am measuring *phone usage* what would low dispersion tell me?
-   If I am measuring *phone usage*, what would high dispersion tell me?

## Measures of Dispersion

### Measures of dispersion tell us about consensus.

-   Did participants give similar answers?

    -   $\{1,2,2,1,2,2,3,2,1\}$

-   Did participants give wildly different answers?

    -   $\{1,3,4,5,8,9,21,33\}$

## Dispersion

### Measures of dispersion tell us about consensus.

-   Data should have natural dispersion.

-   If everyone gives a similar answer, it's harder to analyze difference.

## Calculating Dispersion

### How do you think we should calculate dispersion?

-   On average, how far is each x-value from the midpoint

    -   Calculate how far each x-value is from the midpoint (mean)

    -   Take the average of those distances.

## Calculating Dispersion

### How do you think we should calculate dispersion?

In psychology, the mid-point we use will be the mean

## Calculating Dispersion

### Acquire Data

```{r}
library(gt)

data.frame(
  Participant = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2)
) |> gt() |> 
  opt_row_striping() |> 
  cols_align(align = "center") |> 
        tab_options(
    table.font.size = px(25), # Adjust font size
    table.width = pct(80)   # Set table width as percentage
  ) |> 
    tab_style(
    style = 
      cell_text(
        weight = "bold"),
        locations = 
      cells_column_labels()
    )
```

## Deviation Scores

### Calculate how far each x-value is from the mean

::: fragment
Step 1. Examine Data
:::

-   $data = \{1,3,2,2,3,2,1,3,1,2\}$

::: fragment
Step 2. Calculate Mean
:::

-   $\bar{x} = \frac{\sum(x)}{n} = 1 + 3 + 2...$

-   $\bar{x} = \frac{20}{10} = 2$

## Deviation Scores

### Calculate how far each x-value is from the mean

Step 2: Subtract the mean from each X-value.

-   $x - \bar{x}$

-   $(1-2) + (3-2) + (2-2) + (2-2)...$

-   $(-1) + 1 + 0 + 0...$

## Calculating Dispersion

### Visual

```{r}
tibble(
  ParticipantID = 1:10,
  x = c(1,3,2,2,3,2,1,3,1,2),
  xM = x-mean(x)
) |> gt() |> 
  opt_row_striping() |> 
  cols_label(xM = md("$x-\\bar{x}$")) |> 
  cols_align(align = "center") |> 
  tab_options(
    table.font.size = px(25), # Adjust font size
    table.width = pct(80)   # Set table width as percentage
  ) |> 
  tab_style(
    style = 
      cell_text(
        weight = "bold"),
        locations = 
      cells_column_labels()
    )
```

## Calculating Dispersion

### Visual

```{r}
tibble(
  ParticipantID = factor(1:10),
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X)
) |> 
  ggplot(aes(ParticipantID,X)) + 
  geom_point(color = "darkgreen") + 
  geom_hline(aes(yintercept = mean(X)),color = "red") + 
  theme_minimal() +
  geom_segment(aes(x = ParticipantID, xend = ParticipantID,
                  y = mean(X),
                  yend = X)
              ) +
  labs(
  x = "\nParticipant ID",
  y = "X Value\n"
  ) + 
  theme(panel.grid.major = element_blank())
```

## Calculating Dispersion

### Visual

```{r}
library(forcats)

tibble(
  ParticipantID = factor(1:10,ordered = TRUE),
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X)
) |> 
  ggplot(aes(ParticipantID,X)) + 
  geom_hline(aes(yintercept = mean(`x-M`)),lty = "dashed") + 
  geom_segment(aes(x = ParticipantID, xend = ParticipantID,
                  y = mean(`x-M`),
                  yend = `x-M`)
              ) +
  geom_point(aes(ParticipantID,`x-M`, color = "black")) +
  theme_minimal() +
  labs(
  x = "\nParticipant ID",
  y = "X Value\n"
  ) + 
  theme(
    panel.grid.major = element_blank(),
    legend.position = "none"
    )

```

## Calculating Dispersion

### Visual

-   The average of the deviation scores will always = 0.

-   On average, the x-values = x.

-   So the average distance of x from x is 0.

## Calculating Dispersion

### How do you think we should calculate dispersion?

1.  Calculate the mean.
2.  Calculate the deviation scores.
3.  What can we do to the deviation scores to prevent them from adding up to 0?

## What is the Mean of the Deviation Scores?

### Stop Being So Negative!

-   The problem is the negative numbers!

-   How do we get rid of negative numbers

-   $(x-\bar{x})^2 = (1-2) = 0 | (1-2)^2 = 1$

## Calculating Dispersion

### Stop Being So Negative!

```{r}
tibble(
  ParticipantID = 1:10,
  x = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = x-mean(x),
  `(x-M)^2` = `x-M`^2
  ) |> 
  # Bold Last Column
  gt() |> 
  opt_row_striping() |> 
  cols_align(align = "center") |> 
  cols_label(
    `x-M` = md("$(x-\\bar{x})$"),
    `(x-M)^2` = md("$(x-\\bar{x}^2)$")
  ) |> 
  tab_options(
    table.font.size = px(25), # Adjust font size
    table.width = pct(80)   # Set table width as percentage
  ) |> 
    tab_style(
    style = 
      cell_text(
        weight = "bold"),
        locations = 
      cells_column_labels()
    )
```

## Calculating Dispersion

### First 3 Steps

-   Step 1: Calculate the mean.

-   Step 2: Subtract the mean from each X-value.

-   Step 3: Square each deviation score.

## Calculating Dispersion

### Now what: Calculate the average squared deviation.

-   Step 4: Add up the Squared Deviations.

-   $1+1+0+1...$

## Calculating Dispersion

### Now what: Calculate the average squared deviation.

```{r}
tibble(
  ParticipantID = 1:10,
  x = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = x-mean(x),
  `(x-M)^2` = `x-M`^2
  ) |> 
  # Bold Last Column
  gt() |> 
  opt_row_striping() |> 
  cols_align(align = "center") |> 
  cols_label(
    `x-M` = md("$(x-\\bar{x})$"),
    `(x-M)^2` = md("$(x-\\bar{x}^2)$")
  ) |> 
    tab_options(
    table.font.size = px(25), # Adjust font size
    table.width = pct(80)   # Set table width as percentage
  ) |> 
    tab_style(
    style = 
      cell_text(
        weight = "bold"),
        locations = 
      cells_column_labels()
    )
```

## Calculating Dispersion

### Divide by n-1 to Estimate

Step 5: Divide by (n-1).

-   (n-1) refers to the *degrees of freedom*... don't worry about it for now.

-   $\frac{\sum (x-\bar{x})^2}{n-1}$

-   Variance

    -   The fact or quality of being different, divergent, or inconsistent.

## Calculating Dispersion

### How do think we should calculate dispersion?

-   When we squared the deviation scores in Step 3, we inflated the deviation.

    -   We have to undo that inflation.

    -   How do you undo squaring?

    -   We take the square root.

## Calculating Dispersion

### Inflation

```{r}
set.seed(348)

infl <- tibble(
  ParticipantID = 1:6,
  X = runif(6,30,50),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) |> 
  gt() |> 
    grand_summary_rows(
    # Specify the single column to summarize
    columns = `(x-M)^2`,
    # Provide the aggregation function
    fns = list(
      Sum = ~sum(.),
      Mean = ~sum(.)/5
    )
  ) |> 
  cols_label(
  `x-M` = md("$(x-\\bar{x})$"),
  `(x-M)^2` = md("$(x-\\bar{x}^2)$")
  ) |> fmt_auto() |> 
      tab_options(
    table.font.size = px(25), # Adjust font size
    table.width = pct(80)   # Set table width as percentage
  ) |> 
    tab_style(
    style = 
      cell_text(
        weight = "bold"),
        locations = 
      cells_column_labels()
    )

infl
```

## Calculating Dispersion

Step 6: Take the square root.

-   $\sqrt{\frac{\sum(x-\bar{x})^2}{n-1}}$

## Calculating standard deviation

### Step-by-Step

1.  Mean: Calculate the mean.

    -   $\bar{x}$

2.  Deviation scores: Subtract the mean from each x-value.

    -   $(x-\bar{x})$

3.  Squared deviations: Square each deviation score.

    -   $(x-\bar{x})^2$

4.  Sum of squares: Add up the squared deviations.

    -   $\sum(x-\bar{x})^2$

## Calculating standard deviation

### Step-by-Step

5.  Variance: Divide the sum of squares by (n-1).

    -   $\frac{\sum(x-\bar{x})^2}{n-1}$

6.  Standard Deviation: Take the square root of the variance.

    -   $\sqrt{\frac{\sum(x-\bar{x})^2}{n-1}}$

## Formulas for Dispersion

### Variance and Standard Deviation

::::::: columns
:::: {.column width="50%"}
::: callout-important
## Variance

$$
\large{\color{gold}{s^2} = \frac{\sum(x-\bar{x})^2}{n-1}}
$$
:::
::::

:::: {.column width="50%"}
::: callout-important
## Standard Deviation

$$
\large{\color{orange}{s} = \sqrt{\frac{\sum(x-\bar{x})^2}{n-1}}}
$$
:::
::::
:::::::

## Formulas for Dispersion

### Standard Deviation

-   Variance is a measure of dispersion.

    -   It's not helpful, because it's inflated (from squaring the deviation scores).

-   Standard deviation is a better measure of dispersion, because it is standardized.

    -   This means that a standard deviation of 1 means a distance of 1 on the scale used to measure x.

## Variance & Standard deviation

-   The Variance is the standard deviation squared: s²

-   The Standard Deviation is the square root of the variance: s

## Practice 1 — Basic mean, deviations, variance, SD

**Problem (do by hand first):**\
Compute the **sample mean**, each **deviation** from the mean, the **sum of squared deviations**, the **sample variance**, and the **sample standard deviation** for the dataset:

`x = [4, 2, 5, 6, 3]`

## Practice 1

### Solution (steps)

::::: columns
::: {.column width="50%"}
1.  Sum: $(\sum x = 4+2+5+6+3 = 20)$
2.  Sample size: $(n = 5)$
3.  Mean: $(\bar{x} = \frac{20}{5} = 4)$
4.  Deviations: $(x_i - \bar{x} = [0, -2, 1, 2, -1])$
5.  Squared deviations: $([0, 4, 1, 4, 1])$
6.  Sum of squared deviations: $(0+4+1+4+1 = 10)$
7.  Sample variance: $(s^2 = \frac{\sum (x_i-\bar{x})^2}{n-1} = \frac{10}{4} = 2.5)$
8.  Sample SD: $(s = \sqrt{2.5} \approx 1.5811)$
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| output-location: column

x <- c(4,2,5,6,3)
n <- length(x)
sum_x <- sum(x)
mean_x <- mean(x)
deviations <- x - mean_x
sq_dev <- deviations^2
ssd <- sum(sq_dev)
s2 <- var(x)  # sample variance by default (n-1)
s <- sd(x)
list(
  n = n,
  sum = sum_x,
  mean = mean_x,
  deviations = deviations,
  squared_deviations = sq_dev,
  sum_squared_deviations = ssd,
  sample_variance = s2,
  sample_sd = s
)
```
:::
:::::

## Problem 2: Comparing dispersion

### Mathematically

> Two instructors have class scores: - Class A: scores_A = $[80, 82, 78, 85, 75]$ - Class B: scores_B = $[60, 90, 75, 85, 85]$ 1. Compute the mean and sample SD for each class. 2. Both classes have similar means. Which class has greater dispersion? What does that imply about reporting a “typical” student score?

## Problem 2: Comparing dispersion

### Mathematically

::::: columns
::: {.column width="50%"}
-   **Step 1: Compute the mean**

    -   $\bar{x}_A = \frac{80+82+78+85+75}{5} = \frac{400}{5} = 80$

-   **Step 2: Compute deviations and squared deviations**

-   $x_i - \bar{x}_A = [0, 2, -2, 5, -5]$

-   $(x_i - \bar{x}_A)^2 = [0, 4, 4, 25, 25]$

-   **Step 3: Compute variance and SD**

-   $s_A^2 = \frac{\sum (x_i - \bar{x}_A)^2}{n-1} = \frac{58}{4} = 14.5$

-   $s_A = \sqrt{14.5} \approx 3.81$
:::
::: {.column width="50%"}
-   **Step 1: Compute the mean**

    -   $\bar{x}_B = \frac{60+90+75+85+85}{5} = \frac{395}{5} = 79$

-   **Step 2: Compute deviations and squared deviations**

    -   $x_i - \bar{x}_B = [-19, 11, -4, 6, 6]$

    -   $(x_i - \bar{x}_B)^2 = [361, 121, 16, 36, 36]$

-   **Step 3: Compute variance and SD**

    -   $s_B^2 = \frac{\sum (x_i - \bar{x}_B)^2}{n-1} = \frac{570}{4} = 142.5$

    -   $s_B = \sqrt{142.5} \approx 11.94$
:::
:::::

## Interpretation

-   Class A: $\bar{x} = 80, s \approx 3.81$
-   Class B: $\bar{x} = 79, s \approx 11.94)$

## Problem 2: Comparing dispersion

### Computationally

```{r}
#| echo: true
#| output-location: column
scores_A <- c(80,82,78,85,75)
scores_B <- c(60,90,75,85,85)

stats_A <- list(mean = mean(scores_A), sd = sd(scores_A), n = length(scores_A))
stats_B <- list(mean = mean(scores_B), sd = sd(scores_B), n = length(scores_B))

stats_A

stats_B
```
