---
format: 
  revealjs:
    theme: custom.scss
    incremental: true   
    touch: true
    height: 1080
    width: 1920
    highlight-style: a11y
---

:::{.title-center}
Welcome to Statistics for Psychology!
:::

:::{.author}
Dave Brocker
Lecturer
Farmingdale State College
:::

## Information

### Who Am I?

**Name:** Professor Brocker

**Background:** 7+ Years of Teaching Undergraduate Statistics, Researcher and Laboratory Manager

**Education**: M.A. in Experimental Psychology (specializing in Morality and Emotion).

**Current Research**: Morbid Curiosity, Creepiness, Disgust and Anger, and Dehumanization

## My Teaching Style

### What Can You Expect?

-   **Humor:** I (try) to be funny

-   **Interactive:** Whenever possible, I favor dynamic content over static content

-   **Real-World Examples:** We will use a mixture of real and fabricated data to help illustrate ideas

## What Are Students Saying?

### My Course Evaluations

```{r}
library(readxl)
library(dplyr)
library(gt)
ev <- read_excel("statscomments.xlsx")
```

::: panel-tabset
## Instructor Comments

```{r}
ev_c <- 
  ev |> 
  filter(Type == "Instructor Comments") |> 
  select(Type,Comment)

ev_c |> gt(rowname_col = "Type")
```

```{=html}
<ul>
<li>Smart and funny guy. Answered questions to the best of his ability. Made 9am class feel like a mid-day/evening class, not boring. Weirdly made me think statistics was fun and interesting. And that research is cool. Excited to take classes with him in the future.</li>
<li>this professor likes telling jokes and telling people to leave class which is cringey. he talks too fast and it’s very hard to keep up. very friendly, nice, caring, and accommodating but the jokes and talking speed is a bit much for me personally, not really my taste.</li>
<li>Professor Brocker is in tune with the class and meets us at the level we are at and tries to pull us up to the level we are supposed to be. The only suggestion is that we need homework to reinforce what we have learned. Overall, Prof. Brocker was inspiring, I never thought I would understand anything related to Statistics, but he made it clear. He was ALWAYS PREPARED, made time for us and helped us any way he could. He has such great energy, approachable, knows his material, finds ways to deliver the material. Really great experience, in spite of my grade. It was me, NOT him.</li>
<li>The professor is greatly appreciated. He made the course interesting and was very funny. I think his fun personality definitely helped keep us awake and interested in the course regardless of the time.</li>
<li>Very knowledgeable and professional, very engaging with students and available for extra help</li>
<li>He is a great instructor but if you are not a stats lover you it can be very confusing, he is funny and has a personality.</li>
<li>Prof. Brocker is a wonderful professor and I am so happy I had this class with him. He could slow down in how he speaks as I don’t get every word he says cuz he speaks like the flash. His jokes are so funny and makes me now want to kill myself in math as much</li>
<li>i love how fun you made the class, funny guy great teacher.  material was great and you explained things very well.</li>
<li>With peace and love, he can slow down while he teaches</li>
<li>Great instructor hate statistics</li>
<li>Professor is very nice and calm man. He teaches the class very well and it shows how much he cares for his students. Is very lenient and easy going. Always can ask him any question when needed. I would take him again.</li>
<li>Prior to taking this course, I strongly disliked math of any kind, but I have found a genuine love for statistics after this class. The class was taught well, and Professor Brocker was extremely helpful in teaching the material. Thanks for a great semester!</li>
<li>Best professor I’ve ever had</li>
<li>Overall you're a great person and your jokes were actually funny. You were available after class and was really generous with points for the exams.</li>
<li>Awesome professor, did a great job teaching this class. Also loves riffing a lot.</li>
<li>Super approachable, understanding and just overall cares about his students. I would absolutely have him as a professor again.</li>
<li>Professor Brocker is amazing. He truly takes his time to make sure the students understand the material, even if that means that he has to go over the same thing over and over again. Math is not my best subject, but he helped me understand statistics more than I ever knew before.</li>
<li>I genuinely appreciated having professor Brocker this semester. Even though the subject wasn't something I was particularly interested in, he made the class engaging and enjoyable. He has a great sense of humor and a very caring, approachable personality that made students feel comfortable participating and asking questions.</li>
<li>What stood out most was how understanding and supportive he was, he always made time to help. His ability to connect with students and bring a positive energy to the classroom really made a difference. Overall, he's one of the most personable and helpful professors I've had, and I would definitely recommend his class to others.</li>
<li>Dr. Brocker was a very cool, down-to-earth professor.  He utilizes his knowledge for our gain.  I appreciated him as a professor.</li>
<li>I was so nervous going into this class because math is very difficult for me. My anxiety was immediately relieved when I met professor Brocker. He will do anything he can to make sure you actually understand the information. He cares about every student and will go to lengths to help every single student excel.</li>
<li>David Brocker has a genuine passion for teaching and a strong commitment to helping his students succeed. Despite me growing up disliking math, he has managed to make the subject interesting and approachable. He is truly the best professor I have ever had; I cannot say anything negative about him as both a professor and a person.</li>
<li>Favorite professor I have had so far.</li>
<li>Prof. Brocker really cares about his students and it is obvious. He goes out of his way to get his students to understand the material. The material is hard and confusing at times but he tries to get you to understand using different approaches. </li>
<li>Great instructor</li>
<li>He took his time in teaching and made the lectures and labs very understandable. Before his exams, he always provided great reviews such as Kahoot and review sheets.</li>
<li>I'm not sure how I felt while learning but am thankful for what I've retained.</li>
<li>Professor Brocker ensured that we demonstrated competence before progressing to the next topic. He welcomed all questions and provided thorough answers to ensure everyone understood before moving on. His slides were simplified to their most essential form, offering a clear understanding of each topic. He also included additional resources in his slides for students to reference after class. Overall, his teaching was very organized and beneficial for students.</li>
<li>Learning was easy in this class. Professor made really took the the time to make sure we all understood the material.</li>
<li>This class is difficult. I spent most of the semester frustrated and confused. Then at the end of semester I was shocked that things were actually making sense. I am very pleased with the result. </li>
<li>great</li>
</ul>
```

## Learning Process

## Course Suggestions
:::

## Outside of Class

### Media and Music

```{r}
library(bslib)

c1 <- 
    card(
    card_header("Favorite TV Show"),
    card_image(src ="https://resizing.flixster.com/lpJkDxnEFNQT1OWJjnmYfvpAHJ0=/ems.cHJkLWVtcy1hc3NldHMvdHZzZXJpZXMvUlRUVjI2NjgyOS53ZWJw")
    )

c2 <- 
  card(
    card_header("Favorite Movie"),
    card_image(src ="https://m.media-amazon.com/images/M/MV5BY2VhODM5OTUtZDJhMi00MTc5LThjNjYtZWY1M2NlNWU0N2NjXkEyXkFqcGc@._V1_FMjpg_UX1000_.jpg")
    )

c3 <- 
      card(
    card_header("Favorite Music"),
    card_image(src ="https://f4.bcbits.com/img/a0356251795_16.jpg")
    )


layout_column_wrap(
  c1,c2,c3
)
```

## Test Slide 2

```{r}
library(leaflet)
leaflet() %>%
  addTiles() %>% 
  setView(lat=40.750543,lng=-73.429724, zoom = 15) |> 
  addCircleMarkers(lat=40.750543,lng=-73.429724,50)
```

## What is Statistics?

### Definition

**Statistics** is the science of collecting, analyzing, and interpreting data to make informed decisions.

## Why Should You Care?

### "Why Do I Even Need This Course?"

-   As a practitioner, your practices need to be informed by evidence-based treatments
-   As a researcheer, your claims need to be supported by evidence that can assert a high probability or relationship or cause and effect.
-   Because data is everywhere, even in your **Netflix** recommendations!

## What You'll Learn This Semester

### Introduction to Research

-   Descriptive Statistics
-   Probability is Paramount
-   Testing hypotheses
-   Correlation, regression, and causation

## How This Class Works

### Course Structure

-   **Participation:** Asking questions helps you understand and will likely help out your classmates!
-   **Homework:** I want you to have practice outside of the classroom (low-stakes)
-   **Exams:** This is a decent (not perfect) method of seeing how well you understand and can apply the material

## Any Questions?

### I Am an Open Book!
