---
title: "One-Sample T-Test"
subtitle: "PSY 348: Lecture 9"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "One-Sample T-Test | ⬡⬢⬡⬢⬡⬢"
format:
  revealjs:
    theme: ../custom.scss
    incremental: true
    touch: true
    chalkboard: true
    lightbox: true
    scrollable: true
    code-fold: true
    slide-number: c/t
    include-after-body: ../title-bg.html
    width: 1600
    height: 900
    drop:
      engine: webr
revealjs-plugins:
  - drop
  - codewindow
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(broom)
library(gt)

# Site tokens (see custom.scss / DESIGN_SYSTEM.md)
sage   <- "#5C715E"  # $accent
sage2  <- "#8FAF9E"  # $accent2
sage3  <- "#A7C7B5"  # $accent3
ink    <- "#2F3E46"  # $body-color
hot    <- "#B5483B"  # "look here" highlight
amber  <- "#E8A33D"

theme_deck <- function(base_size = 22) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.title = element_text(color = ink),
      axis.text = element_text(color = ink),
      plot.title = element_text(color = ink, face = "bold"),
      legend.text = element_text(color = ink)
    )
}
```

## Recap

### How did we get here?

:::::: badge-row
::: content-badge
[straighten]{.material-icons}

#### Measurement

Measuring abstract concepts
:::

::: content-badge
[category]{.material-icons}

#### Variables

Types and scales
:::

::: content-badge
[show_chart]{.material-icons}

#### The Normal Curve

Probability from z-scores
:::

::: content-badge
[groups]{.material-icons}

#### Sampling Theory

Distribution of sample means
:::
::::::

::: {.callout-tip appearance="minimal"}
If we can estimate the probability that a sample mean came from a particular distribution, can we also estimate whether it came from the null or the alternative hypothesis?
:::

## From Probability to Decision-Making

- Once we know how likely a sample mean is to occur, we can begin to make decisions about what that likelihood means.

- We introduced two competing ideas:

::::: columns
::: {.column width="50%"}
::: callout-tip
## Null hypothesis $H_0$

No real difference or effect.
:::
:::

::: {.column width="50%"}
::: callout-important
## Alternative hypothesis $H_1/H_A$

A true difference or effect exists.
:::
:::
:::::

- Because samples vary, we can never be 100% certain, only confident to a degree.

## Errors in Estimation

### Type I and II Error

- That degree of confidence is defined by $\alpha$ (alpha), our threshold for error when rejecting $H_0$.

- $\beta$ (beta) is the probability of missing a real effect (failing to reject $H_0$ when $H_1$ is true).

::::: term-list
::: fragment
[Type I Error ($\alpha$)]{.term-pill}

: Finding a difference that isn't really there. (False Positive)
:::

::: fragment
[Type II Error ($\beta$)]{.term-pill}

: Missing a difference that actually exists. (False Negative)
:::
:::::

## The Truth Table

```{r}
#| echo: false
error_table <- data.frame(
  "Decision" = c("$$\\text{Reject }H_0$$", "$$\\text{Fail to Reject } H_0$$"),
  "null_true" = c("$$\\text{Type I Error }(\\alpha)$$", "$$\\text{Correct Decision}\\\\1-\\alpha$$"),
  "null_false" = c("$$\\text{Correct Decision}\\\\1-\\beta$$", "$$\\text{Type II Error } (\\beta)$$")
)

error_table |>
  gt(rowname_col = "Decision") |>
  tab_header(title = md("**Decision Outcomes in Hypothesis Testing**")) |>
  cols_label(
    null_true = md("$$H_0\\text{ True}$$"),
    null_false = md("$$H_0\\text{ False}$$")
  ) |>
  fmt_markdown(columns = everything(), rows = everything()) |>
  tab_style(
    style = cell_fill(color = "#EBC9C3"),
    locations = list(cells_body(columns = 2, rows = 1),
                     cells_body(columns = 3, rows = 2))
  ) |>
  tab_style(
    style = cell_fill(color = "#DCE8E1"),
    locations = list(cells_body(columns = 2, rows = 2),
                     cells_body(columns = 3, rows = 1))
  ) |>
  tab_options(table.font.size = 25, table.width = pct(70))
```

## Alpha and Beta

### Seeing the two errors

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4.4
x <- seq(-4, 8, length.out = 900)
crit <- qnorm(.95)
d <- bind_rows(
  tibble(x, y = dnorm(x, 0, 1), curve = "H0 true"),
  tibble(x, y = dnorm(x, 3, 1), curve = "H1 true (a real effect)")
)
ggplot(d, aes(x, y, color = curve)) +
  geom_area(data = filter(d, curve == "H0 true", x >= crit),
            aes(fill = "alpha"), alpha = .75, color = NA) +
  geom_area(data = filter(d, curve != "H0 true", x <= crit),
            aes(fill = "beta"), alpha = .75, color = NA) +
  geom_line(linewidth = 1.3) +
  geom_vline(xintercept = crit, linetype = "dashed", color = ink) +
  scale_color_manual(values = c("H0 true" = sage, "H1 true (a real effect)" = ink)) +
  scale_fill_manual(values = c(alpha = hot, beta = amber)) +
  labs(x = "Test statistic", y = NULL, color = NULL, fill = NULL) +
  theme_deck(20) +
  theme(axis.text.y = element_blank(), panel.grid = element_blank(),
        legend.position = "top")
```

- The dashed line is the **critical value**: results beyond it lead us to reject $H_0$.

- $\alpha$ (red): the part of the $H_0$ curve beyond the line, so a true null gets rejected (Type I error).

- $\beta$ (orange): the part of the $H_1$ curve before the line, so a real effect is missed (Type II error).

- **Power** is $1 - \beta$: the chance of detecting a real effect.

## Why Compare Means?

- We often want to test whether a sample mean differs from a known population mean.

- The one-sample t-test helps determine if the sample mean is significantly different from the hypothesized population mean.

:::::: badge-row
::: content-badge
[school]{.material-icons}

#### Exam scores

Class average vs. a national average of 75
:::

::: content-badge
[timer]{.material-icons}

#### Reaction times

Average vs. a published 250 ms standard
:::

::: content-badge
[linear_scale]{.material-icons}

#### Scale midpoints

Ratings vs. the scale midpoint
:::
::::::

## Why Not z?

### From z to t

- In lecture 8 we computed $z = \dfrac{\bar{x} - \mu}{\sigma / \sqrt{n}}$, which needs the **population** standard deviation $\sigma$.

- In practice we almost never know $\sigma$, so we substitute the sample's $s$.

- But $s$ changes from sample to sample, especially when $n$ is small. That is **extra uncertainty** the normal curve does not account for.

- So the sampling distribution has **heavier tails** than the normal curve: the t-distribution. A result must be more extreme to count as unusual.

## Why Not z?

### Same $\alpha = .05$, bigger cutoff for small samples

Two-tailed critical values:

:::::::::::: side-by-cols
::::: stat-card
::: stat-num
1.96
:::

::: stat-sub
z ($\sigma$ known)
:::
:::::

::::: stat-card
::: stat-num
2.09
:::

::: stat-sub
t, $df = 19$ ($n = 20$)
:::
:::::

::::: stat-card
::: stat-num
2.78
:::

::: stat-sub
t, $df = 4$ ($n = 5$)
:::
:::::
::::::::::::

- The smaller the sample, the more the t cutoff moves away from the z cutoff. With large $n$, they are nearly identical.

## The t-test

### Three types, three cases

::::: columns
::: {.column width="33%"}
::: callout-tip
## One-Sample

$t=\dfrac{\bar{x}-\mu}{\sigma/\sqrt{n}} \;|\; \dfrac{\bar{x}-\mu}{s/\sqrt{n}}$
:::
:::

::: {.column width="33%"}
::: callout-tip
## Paired-Samples

$t=\dfrac{\bar{D}}{\sigma_D/\sqrt{n}} \;|\; \dfrac{\bar{D}}{s_D/\sqrt{n}}$
:::
:::

::: {.column width="33%"}
::: callout-tip
## Independent-Samples

$t = \dfrac{\bar{x}_1 - \bar{x}_2}{\sqrt{s_p^2\left(\frac{1}{n_1} + \frac{1}{n_2}\right)}}$

$s_p^2 = \dfrac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}$
:::
:::
:::::

- We use the sample standard deviation ($s$) to estimate the standard error ($s/\sqrt{n}$) since population $\sigma$ is typically unknown.

- $s_D$ is the standard deviation of the **difference scores**; $s_p^2$ is the **pooled variance**, a weighted average of the two groups' variances.

- In all cases, the t-distribution's shape (and so the critical value) depends on the **degrees of freedom** (df).

  - One-sample and paired: $df = n - 1$. Independent samples: $df = n_1 + n_2 - 2$

## The t-distribution

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4.6
df <- tibble(
  t.values = seq(-4, 4, .1),
  df3 = dt(t.values, 3),
  df10 = dt(t.values, 10),
  df30 = dt(t.values, 30),
  df80 = dt(t.values, 80),
  normal = dnorm(t.values)
) |>
  pivot_longer(cols = -t.values, names_to = "distribution", values_to = "density") |>
  mutate(distribution = factor(distribution,
                               levels = c("df3", "df10", "df30", "df80", "normal")))

ggplot(df, aes(x = t.values, y = density, color = distribution, linetype = distribution)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c(df3 = hot, df10 = amber, df30 = sage3,
                                df80 = sage, normal = ink)) +
  scale_linetype_manual(values = c(df3 = "solid", df10 = "solid", df30 = "solid",
                                   df80 = "solid", normal = "dashed")) +
  labs(title = "t-distributions with increasing degrees of freedom",
       x = "t-value", y = NULL, color = NULL, linetype = NULL) +
  theme_deck(20) +
  theme(legend.position = "top", axis.text.y = element_blank())
```

- The t-distribution has **heavier tails** than the normal curve when df is small, because $s$ is only an estimate of $\sigma$.

- As df grows, the t-distribution approaches the normal curve.

## Assumptions of the One-Sample t-Test

::: callout-important
## Assumptions

1.  The data are continuous (interval or ratio scale).
2.  The sample is randomly selected.
3.  The data are normally distributed (or the sample size is large enough for the Central Limit Theorem to apply).
:::

## The One-Sample t-Test Formula

::: callout-tip
## One-Sample t-Test

$\large{t = \dfrac{\color{blue}{\bar{x}} - \mu}{\color{orange}{s}/\sqrt{n}}}$
:::

Where:

- $\color{blue}{\bar{x}}$ = sample mean

- $\mu$ = population mean (null hypothesis value)

- $\color{orange}{s}/\sqrt{n}$ = the standard error (SE), estimated from the sample since population $\sigma$ is typically unknown

  - $\color{orange}{s}$ = sample standard deviation

  - $n$ = sample size

## Hypothesis Testing Steps

::: callout-tip
## Four steps

1.  **Set hypotheses:** define both the null and alternative.
2.  **Collect sample data:** use the appropriate sampling method and collect the needed data.
3.  **Compute the t-statistic:** conduct the correct t-test.
4.  **Compare to the critical value:** use the t-statistic, sample size (df), and alpha level to determine statistical significance. Equivalently, compare the p-value to $\alpha$.
:::

## Reading the t-Table

### Critical values by degrees of freedom

::: z-excerpt
| df | One-tailed .05<br>(two-tailed .10) | One-tailed .025<br>(two-tailed .05) | One-tailed .005<br>(two-tailed .01) |
|:----------------:|:----------------:|:----------------:|:----------------:|
| 15 | 1.753 | 2.131 | 2.947 |
| **19** | 1.729 | [2.093]{.z-hit} | 2.861 |
| 24 | 1.711 | 2.064 | 2.797 |
| 29 | 1.699 | 2.045 | 2.756 |
| ∞ (z) | 1.645 | 1.960 | 2.576 |

: Excerpt of the t-table (critical values)
:::

- Find your **df** ($n - 1$ for a one-sample test), then pick the column for your $\alpha$ and your **tails**.

- If $|t|$ is **beyond** the critical value, reject $H_0$.

  - Example: $t = 2.68$ with $df = 19$, two-tailed $\alpha = .05$: $2.68 > 2.093$, so reject.

- In R: `qt(.975, df = 19)` returns `2.093`.

## Hypothesis Testing Steps

### Manual: hypotheses and data

> A sample of `20` students has a mean exam score of `78` with a standard deviation of `5`. Does the class differ from the national average of `75`?

::: callout-tip
## 1. Set hypotheses

- $H_0$: The average exam score is 75 ($\mu = 75$).
- $H_A$: The average exam score is not 75 ($\mu \neq 75$).
:::

::: callout-tip
## 2. Collect sample data

- Sample size: $n = 20$
- Sample mean: $\color{blue}{\bar{x}} = 78$
- Sample standard deviation: $\color{orange}{s} = 5$
:::

## Hypothesis Testing Steps

### Manual: test and decide

::: callout-tip
## 3. Compute the t-statistic

$t = \dfrac{78 - 75}{\frac{5}{\sqrt{20}}} = \dfrac{3}{1.118} = 2.68$
:::

::: callout-tip
## 4. Compare to the critical value

- At $\alpha = .05$ (two-tailed) and $df = 19$: $t_{critical} = \pm 2.093$.
- Since $2.68 > 2.093$, we **reject** $H_0$.
:::

## Hypothesis Testing Steps

### Programmatic

::: {.codewindow .r .codewindow-dark}
one_sample.R

```{r}
#| echo: true
#| code-fold: false

# For Reproducibility
set.seed(20)

n = 20   # sample size
m = 78   # sample mean
mu = 75  # population mean (H0)
s = 5    # sample SD

# Simulate data, then rescale so the sample has exactly mean 78 and SD 5
dat <- rnorm(n)
dat <- m + s * (dat - mean(dat)) / sd(dat)

t.test(dat, mu = mu)

# By hand
(m - mu) / (s / sqrt(n))

# Critical value (two-tailed, alpha = .05)
qt(.975, df = n - 1)
```
:::

## Reporting Results

### APA-style

> A one-sample t-test was conducted to compare exam scores to the national average of `75`. Results showed a significant difference, `t(19) = 2.68, p = .015`, indicating that students scored significantly higher than the national average (`M = 78, SD = 5`).

- What kind of test?
- What is being compared?
- Is it significant?
- What are the test results?

## Reporting Results

### APA-style

> A [one-sample t-test]{.hl} was conducted to [compare exam scores to the national average]{.hl} of `75`. Results showed a [significant difference]{.hl}, [`t(19) = 2.68, p = .015`]{.hl}, indicating that students scored significantly higher than the national average (`M = 78, SD = 5`).

- **What kind of test?** one-sample t-test
- **What is being compared?** exam scores to the national average
- **Is it significant?** significant difference
- **What are the test results?** `t(19) = 2.68, p = .015`

## Interpreting Results

### Sample summary table

- A significant result ($p < .05$) suggests that the sample mean is different from the population mean.

- A non-significant result ($p > .05$) means we fail to reject $H_0$ (insufficient evidence of a difference).

```{r}
#| echo: false
t.test(dat,mu = mu) |>
  tidy() |>
  rename(
    M = estimate,
    t = statistic,
    df = parameter,
    p = p.value
    ) |>
  gt(rowname_col = "method") |>
  fmt_auto() |>
  opt_row_striping() |>
  tab_options(table.font.size = 20)
```

## Visualizing Data

### Scores against the national average

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5.6
data <- data.frame(Scores = c(72, 74, 76, 78, 80, 81, 77, 79, 78, 76,
                              80, 82, 74, 75, 78, 77, 76, 79, 80, 78))
ggplot(data, aes(y = Scores)) +
  geom_boxplot(fill = sage3, color = ink) +
  geom_hline(yintercept = mean(data$Scores), lty = "dashed", color = hot) +
  geom_hline(yintercept = 75, lty = "dotted", color = ink) +
  theme_deck(18) +
  labs(title = "Exam scores",
       subtitle = "dashed = sample mean, dotted = national average (75)") +
  annotate("text", x = .45, y = median(data$Scores),
           label = "Median", hjust = 0, color = sage, size = 5) +
  annotate("text", x = 1.2, y = quantile(data$Scores, 3/4),
           label = "Q3 (75th percentile)", hjust = 0, color = ink, size = 5) +
  annotate("text", x = 1.2, y = quantile(data$Scores, 1/4),
           label = "Q1 (25th percentile)", hjust = 0, color = ink, size = 5) +
  annotate("text", x = 0.8, y = max(data$Scores),
           label = "Max (upper whisker)", hjust = 1, color = ink, size = 5) +
  annotate("text", x = 0.8, y = min(data$Scores),
           label = "Min (lower whisker)", hjust = 1, color = ink, size = 5) +
  theme(axis.text.x = element_blank())
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5.6
set.seed(40)
vdata <- tibble(scores = rnorm(40, 78, 3))

ggplot(vdata, aes(x = "", y = scores)) +
  geom_violin(fill = sage3, color = ink) +
  geom_hline(yintercept = mean(vdata$scores), lty = "dashed", color = hot) +
  geom_hline(yintercept = 75, lty = "dotted", color = ink) +
  geom_jitter(height = 0, width = .01, color = ink) +
  theme_deck(18) +
  labs(title = "Exam scores", x = NULL, y = "Scores")
```
:::
:::::

## Exam Scores vs. National Average

### Hypotheses

> A professor wants to test if students in their class scored differently from the national average of `75` on an exam. A sample of `25` students has a mean score of `78` with a standard deviation of `10`.

- Null Hypothesis ($H_0$): $\mu = 75$ (no difference)

- Alternative Hypothesis ($H_A$): $\mu \neq 75$ (scores differ)

## Exam Scores vs. National Average

### Manual

::: callout-tip
## Compute t

$t = \dfrac{\color{blue}{\bar{x}} - \mu}{\frac{\color{orange}{s}}{\sqrt{n}}} = \dfrac{78 - 75}{\frac{10}{\sqrt{25}}} = \dfrac{3}{2} = 1.5$
:::

::: callout-tip
## Decide

For $df = 24$ ($25 - 1$), the critical value at $\alpha = .05$ (two-tailed) is $\pm 2.064$. Since $1.5 < 2.064$, we **fail to reject** $H_0$.
:::

## Exam Scores vs. National Average

### Programmatic

::: {.codewindow .r .codewindow-dark}
exam_scores.R

```{r}
#| echo: true
#| code-fold: false

# For Reproducibility
set.seed(131)

# Simulate data, then rescale so the sample has exactly mean 78 and SD 10
x <- rnorm(25)
x <- 78 + 10 * (x - mean(x)) / sd(x)

t.test(
  x,
  # The population mean to test against
  mu = 75
  )
```
:::

## APA Reporting

> A one-sample t-test was conducted to determine whether students' exam scores differed from the national average ($\mu$ = 75). Results were not statistically significant, `t(24) = 1.50, p = .147`, indicating that students performed similarly to the national average (`M = 78, SD = 10`).

## Daily Coffee Consumption

### Hypotheses

> A coffee company claims that people drink an average of `3 cups` of coffee per day. A researcher samples `16 individuals`, finding a mean of `2.5 cups` and a standard deviation of `1 cup`.

- Null Hypothesis ($H_0$): $\mu = 3$

- Alternative Hypothesis ($H_A$): $\mu < 3$ (people drink less)

## Daily Coffee Consumption

### Manual

::: callout-tip
## Compute t

$t = \dfrac{2.5 - 3}{1 / \sqrt{16}} = \dfrac{-0.5}{0.25} = -2.0$
:::

::: callout-tip
## Decide

For $df = 15$ ($16 - 1$), the critical t-value for a one-tailed test at $\alpha = .05$ is $-1.753$. Since $-2.0 < -1.753$, we **reject** $H_0$.
:::

## Daily Coffee Consumption

### Programmatic

::: {.codewindow .r .codewindow-dark}
coffee.R

```{r}
#| echo: true
#| code-fold: false

# For Reproducibility
set.seed(131)

# Simulate data, then rescale so the sample has exactly mean 2.5 and SD 1
x <- rnorm(16)
x <- 2.5 + 1 * (x - mean(x)) / sd(x)

t.test(
  x,
  # The value claimed by the company
  mu = 3,
  # Our alternative is directional: fewer than 3 cups
  alternative = "less"
  )
```
:::

## APA Reporting

> A one-sample t-test was conducted to test whether daily coffee consumption was lower than the reported average of `3 cups`. Results showed a significant difference, `t(15) = -2.00, p = .032` (one-tailed), suggesting that people consume significantly fewer cups of coffee per day than reported (`M = 2.5, SD = 1`).

## Sleep Duration Among College Students

### Hypotheses

> A health researcher believes college students sleep less than `7 hours` per night. A sample of `30 students` reports a mean of `6.5` hours with a standard deviation of `1.2` hours.

- Null Hypothesis ($H_0$): $\mu = 7$

- Alternative Hypothesis ($H_A$): $\mu < 7$

## Sleep Duration Among College Students

### Manual

::: callout-tip
## Compute t

$t = \dfrac{6.5 - 7}{1.2 / \sqrt{30}} = \dfrac{-0.5}{0.219} = -2.28$
:::

::: callout-tip
## Decide

For $df = 29$, the critical t-value for a one-tailed test at $\alpha = .05$ is $-1.699$. Since $-2.28 < -1.699$, we **reject** $H_0$.
:::

## Sleep Duration Among College Students

### Programmatic

::: {.codewindow .r .codewindow-dark}
sleep.R

```{r}
#| echo: true
#| code-fold: false

# For Reproducibility
set.seed(728)

# Simulate data, then rescale so the sample has exactly mean 6.5 and SD 1.2
x <- rnorm(30)
x <- 6.5 + 1.2 * (x - mean(x)) / sd(x)

t.test(
  x,
  # The value we are testing against
  mu = 7,
  # Our alternative is directional: less than 7 hours
  alternative = "less"
  )
```
:::

## APA Reporting

> A one-sample t-test was conducted to test whether college students sleep fewer than 7 hours per night. Results were statistically significant, `t(29) = -2.28, p = .015` (one-tailed), indicating that students get significantly less sleep than the recommended amount (`M = 6.5, SD = 1.2`).
