---
title: "Linear Regression"
subtitle: "PSY 348: Lecture 16"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Linear Regression | ⬡⬢⬡⬢⬡⬢"
format:
  revealjs:
    theme: ../custom.scss
    scrollable: true
    width: 1600
    height: 900
    incremental: true
    touch: true
    chalkboard: true
    lightbox: true
    code-fold: true
    slide-number: c/t
    include-after-body: ../title-bg.html
    drop:
      engine: webr
revealjs-plugins:
  - drop
  - codewindow
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)
library(gt)
library(gtsummary)

# Site tokens (see custom.scss / DESIGN_SYSTEM.md)
sage  <- "#5C715E"  # $accent
sage2 <- "#8FAF9E"  # $accent2
sage3 <- "#A7C7B5"  # $accent3
ink   <- "#2F3E46"  # $body-color
hot   <- "#B5483B"  # "look here" highlight
amber <- "#E8A33D"

theme_deck <- function(base_size = 22) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.title = element_text(color = ink),
      axis.text = element_text(color = ink),
      plot.title = element_text(color = ink, face = "bold"),
      legend.position = "none"
    )
}

f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fp <- function(p) ifelse(p < .001, "< .001", paste("=", sub("^0", "", sprintf("%.3f", p))))
# coefficient helpers: b and p for one term of an lm
b_of <- function(m, term) unname(coef(m)[term])
p_of <- function(m, term) summary(m)$coefficients[term, 4]
apa_b <- function(m, term) sprintf("b = %s, p %s", f2(b_of(m, term)), fp(p_of(m, term)))
model_p <- function(m) {
  f <- summary(m)$fstatistic
  pf(f[1], f[2], f[3], lower.tail = FALSE)
}

# ---- Line of best fit demo (two correlated variables) ------------------
set.seed(51)
n0 <- 35
x0 <- rnorm(n0)
bf <- tibble(X = x0, Y = .51 * x0 + sqrt(1 - .51^2) * rnorm(n0))
mod_bf <- lm(Y ~ X, data = bf)
bf <- bf |> mutate(yhat = predict(mod_bf) |> as.vector())

# ---- Example 1: social media use (age + sense of belonging) -----------
set.seed(123)
age <- sample(18:50, 30, replace = TRUE)
belonging <- round(runif(30, min = 0, max = 10), 2)
social_media_use <- 5 - (0.1 * age) + (0.3 * belonging) + rnorm(30, mean = 0, sd = 0.5)
df_sob <- data.frame(Participant = paste0("P", 1:30), Age = age,
                     Sense_of_Belonging = belonging,
                     Social_Media_Use = round(social_media_use, 2))
sm_fit <- lm(Social_Media_Use ~ Age + Sense_of_Belonging, data = df_sob)
sm_sum <- summary(sm_fit)

# ---- Example 2: energy levels with a categorical predictor -------------
set.seed(123)
n <- 30
duration_tv <- round(runif(n, 0.5, 5), 2)
cups_of_coffee <- sample(0:5, n, replace = TRUE)
prior_sleep <- round(rnorm(n, mean = 7, sd = 1.5), 2)
tv_genre <- sample(c("Drama", "Comedy", "Documentary"), n, replace = TRUE)
# More sleep and coffee raise energy; more TV lowers it; calmer genres lower it a bit
genre_effect <- ifelse(tv_genre == "Drama", -1, ifelse(tv_genre == "Documentary", -1.5, 0))
energy_level <- 2 + (-0.6 * duration_tv) + (0.8 * cups_of_coffee) +
  (0.6 * prior_sleep) + genre_effect + rnorm(n, mean = 0, sd = 1)
df_cat <- data.frame(Participant = paste0("P", 1:n), Duration_TV = duration_tv,
                     Cups_of_Coffee = cups_of_coffee, Prior_Sleep = prior_sleep,
                     TV_Genre = factor(tv_genre, levels = c("Comedy", "Drama", "Documentary")),
                     Energy_Level = round(energy_level, 2))
model_3 <- lm(Energy_Level ~ Duration_TV + Cups_of_Coffee + Prior_Sleep + TV_Genre, data = df_cat)

# ---- Example 3: horror movies ----------------------------------------
library(janitor)
hm <- read.csv("../Data_Files/Excel Files/csv files/Horror Movie Data.csv") |> clean_names()
hm_fit <- lm(rotten_tomato_score ~ scare_score + year + this_film_is_a_sequel, data = hm)
hm_sum <- summary(hm_fit)
hm_full <- lm(scare_score ~ avg_movie_heart_rate_bpm + hrv_difference + highest_spike_bpm +
                rotten_tomato_score + year + this_film_is_a_sequel, data = hm)
```

## Types of Analysis

:::::: badge-row
::: content-badge
[looks_one]{.material-icons}

#### Univariate

One variable, like the mean
:::

::: content-badge
[looks_two]{.material-icons}

#### Bivariate

Two variables, like correlation
:::

::: content-badge
[looks_3]{.material-icons}

#### Multivariate

More than 2 variables
:::
::::::

## Regression

::: callout-tip
## How regression relates to correlation

-   Regression is the **multivariate** version of correlation.

-   Correlation is the **bivariate** version of regression.

-   We're doing the same thing in regression as we do in correlation, **BUT** we can have more than 2 variables.
:::

## Regression terminology

### Outcome and predictors

::::: term-list
::: fragment
[$Y$]{.term-pill}

: The **outcome** variable: the one we choose to predict.
:::

::: fragment
[$\hat{Y}$]{.term-pill}

: The **predicted** value of $Y$ from the model.
:::

::: fragment
[$A$, $B$, etc.]{.term-pill}

: The **predictor** variables we have reason to believe predict $Y$ (one or more).
:::
:::::

## Regression terminology

::: hero-def
[Regress]{.term-pill}

: When we predict $Y$ with $X$, we say that we **regress** $Y$ on $X$.
:::

## Regression Equation

### Step-by-step

$$
\large{\color{red}{Y_i} = \color{green}{\beta_0} + \color{blue}{\beta_1(A)}+\color{orange}{e_i}}
$$

::: fragment
[1. The **observed** value of $Y$ for a particular participant ($i$)]{style="color:red"}
:::

::: fragment
[2. Intercept]{style="color:green"}: the predicted $Y$ when all predictors are 0
:::

::: fragment
[3. Coefficient (weight)]{style="color:blue"}: the predicted change in $Y$ for a one-unit increase in $A$
:::

::: fragment
[4. The value of a particular participant on the measure of $A$]{style="color:black"}
:::

::: fragment
[5. Error]{style="color:orange"}: random, and therefore cannot be predicted
:::

## Regression Equation

### A few ways to write the same thing

$$
\large{\color{red}{Y_i} = \color{green}{\beta_0} + \color{blue}{\beta_1(A)} + e_i}
$$

::: fragment
$$\hat{Y} = \beta_0 + \beta_1(A)$$
:::

::: fragment
$$Y = a + bX$$
:::

## Linear Regression

### Line of best fit

::: panel-tabset
## Scatterplot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.2
ggplot(bf, aes(X, Y)) +
  geom_point(color = sage, size = 3) +
  labs(title = "Two variables", x = "X", y = "Y") +
  theme_deck(20)
```

## Linear regression

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.2
ggplot(bf, aes(X, Y)) +
  geom_point(color = sage, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = hot, linewidth = 1.4) +
  labs(title = "The line of best fit", x = "X", y = "Y") +
  theme_deck(20)
```

## Residuals

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.2
ggplot(bf, aes(X, Y)) +
  geom_segment(aes(xend = X, yend = yhat), linetype = "dashed", color = ink) +
  geom_point(color = sage, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = hot, linewidth = 1.4) +
  labs(title = "Residuals: observed minus predicted", x = "X", y = "Y") +
  theme_deck(20)
```
:::

## Linear Regression

### Least squares

-   The **line of best fit** is the line that makes the **residuals** as small as possible.

-   A **residual** is the distance between an observed $Y$ and its predicted $\hat{Y}$: $e_i = Y_i - \hat{Y}_i$.

-   **Least squares** picks the line that minimizes the sum of the **squared** residuals, $\sum e_i^2$.

## Regression

### Equation component meanings

-   $Y$ = the actual, observed $y$ value for a participant

-   $\hat{Y}$ = the model's predicted $y$ value for a participant

-   $\beta_0$ = the intercept: the predicted $y$ when $A$ is 0

-   $A$ = the $A$ value for a participant

-   $\beta_1$ = the coefficient for $A$: it is multiplied by $A$ and added to $\beta_0$ to estimate $y$

## Regression Equation

### More than one X

::: callout-tip
## Multiple regression

$\large{\hat{Y}_i = \beta_0 + \beta_1(A) + \beta_2(B) + \cdots}$

-   $\beta_1$ and $\beta_2$ are the coefficients for predictors $A$ and $B$.

-   Each coefficient is the predicted change in $Y$ for a one-unit increase in that predictor, **holding the other predictors constant**.
:::

## Regression in R

### The lm() function

::: {.codewindow .r .codewindow-dark}
regression.R

```{r}
#| echo: true
#| code-fold: false
# lm(outcome ~ predictor1 + predictor2, data)
fit <- lm(Social_Media_Use ~ Age + Sense_of_Belonging, data = df_sob)

summary(fit)
```
:::

# Interpreting Coefficients

## Regression Model Table

### Two predictor variables

```{r}
#| echo: false
sm_fit |> tbl_regression(intercept = TRUE) |> as_gt() |>
  tab_options(table.font.size = 26)
```

## Interpreting coefficients

::::: term-list
::: fragment
[Beta ($b$)]{.term-pill}

: The predicted change in $y$ for a **one-unit increase** in the predictor, holding the other predictors constant.
:::

::: fragment
[t-value]{.term-pill}

: The coefficient divided by its standard error: $t = b / SE_b$.
:::

::: fragment
[p-value]{.term-pill}

: Is this a **significant predictor** of $y$?
:::
:::::

::: {.callout-tip appearance="minimal"}
To compare which predictor matters more, use **standardized** betas. Unstandardized betas depend on the units of each predictor (years vs. a 0 to 10 scale).
:::

## $R^2$: Coefficient of Determination

### How much can we account for?

-   $R^2$ is the correlation coefficient ($r$) squared, and is always between 0 and 1.

-   In correlation, it is the amount of $y$ accounted for by $x$.

-   In regression, $R^2$ is the amount of $y$ accounted for by **all** of the predictors together.

-   Did we account for a **significant** amount of the variance in $y$?

## Interpreting R Squared

::::: term-list
::: fragment
[$R^2$]{.term-pill}

: The percentage of $y$ accounted for by the predictors.
:::

::: fragment
[Adjusted $R^2$]{.term-pill}

: The same, but penalized for using too many predictors.
:::

::: fragment
[F-value and p-value]{.term-pill}

: Is the model **as a whole** a significant predictor of $y$?
:::
:::::

## Assumptions of Linear Regression

::: callout-important
## Assumptions

1.  **Linearity:** the relationship between each predictor and $y$ is linear.
2.  **Independence:** the observations (and residuals) are independent.
3.  **Homoscedasticity:** the residuals have about the same spread at every predicted value.
4.  **Normality:** the residuals are approximately normal.
5.  **No severe multicollinearity:** the predictors are not nearly duplicates of one another.
:::

::: {.codewindow .r .codewindow-dark}
diagnostics.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
# Four standard diagnostic plots for an lm
plot(fit)
```
:::

# Linear Regression Example

## Linear Regression

### Example

> Professor Brocker wants to be able to predict social media use. The Social Media Use Scale measures different motivations and reasons for using Social Media platforms.

What are some predictors of social media use?

-   Sense of belonging

-   Age

## Regression Equation

### Theoretical

$$\large{\color{red}{Y_i} = \color{green}{\beta_0} + \color{orange}{\beta_1}\color{blue}{(Age)} + \color{purple}{\beta_2(SOB)} + e_i}$$

1.  The observed degree of social media use for a participant "$i$"

2.  Intercept: predicted social media use when all other predictors are 0

3.  Coefficient for Age ($\beta_1$)

4.  The participant's age

5.  Coefficient for Sense of Belonging ($\beta_2$)

6.  The participant's sense of belonging

7.  Randomness

## Example

### Model coefficients

```{r}
#| echo: false
sm_fit |>
  tbl_regression(intercept = TRUE) |>
  as_gt() |>
  tab_header(title = "Social Media Use Model") |>
  tab_options(table.font.size = 26)
```

## Example

### Model summary

```{r}
#| echo: false
tibble(
  Metric = c("Residual Standard Error", "Multiple R-squared", "Adjusted R-squared", "F-statistic", "p"),
  Value = c(sm_sum$sigma, sm_sum$r.squared, sm_sum$adj.r.squared,
            unname(sm_sum$fstatistic[1]), model_p(sm_fit))
) |>
  gt() |>
  fmt_number(decimals = 3) |>
  tab_header(title = md("**Regression Model Summary**"),
             subtitle = md("*Predicting social media use from age and sense of belonging*")) |>
  tab_options(table.font.size = 26)
```

# Reporting Regression Findings

## Reporting

### The overall model

First, report adjusted $R^2$ and its corresponding $p$:

> Age and sense of belonging predicted a significant amount of variance in social media use, adjusted $R^2$ = `r f2(sm_sum$adj.r.squared)`, *F*(`r sm_sum$fstatistic[2]`, `r sm_sum$fstatistic[3]`) = `r f2(sm_sum$fstatistic[1])`, *p* `r fp(model_p(sm_fit))`.

## Reporting

### The individual predictors

Then report individual betas and their $p$'s:

> Sense of belonging significantly predicted social media use (*`r apa_b(sm_fit, "Sense_of_Belonging")`*), and so did age (*`r apa_b(sm_fit, "Age")`*).

-   It's common to report betas and $p$'s in the text **and** in a table.

-   Interpretation: for each additional year of age, predicted social media use drops by `r f2(abs(b_of(sm_fit, "Age")))` hours per day, holding sense of belonging constant.

# Categorical Variables in Regression

## Multiple Predictors

### Understanding the factors that influence energy levels

> What determines how energized you feel throughout the day? Is it the number of cups of coffee you drink? The amount of sleep you got the night before? Or maybe even the type of TV shows you watch before bed?

## Multiple Predictors

### Four predictors

We explore how TV watching habits (duration and genre), caffeine intake, and prior sleep predict energy levels:

-   **Duration of TV watching** (continuous): does more TV drain your energy?

-   **Cups of coffee** (continuous): does caffeine actually help?

-   **Prior sleep** (continuous): does more sleep mean more energy?

-   **TV genre** (categorical: Comedy, Drama, Documentary): can what you watch change how you feel?

## Categorical Variables

### Dummy coding

::: callout-tip
## How R handles a categorical predictor

A variable with $k$ categories becomes $k - 1$ **dummy variables** (0/1). One category is the **reference** group, and each coefficient compares a category to the reference.
:::

| TV genre | Drama | Documentary |
|:---------|:-----:|:-----------:|
| Comedy (reference) | 0 | 0 |
| Drama | 1 | 0 |
| Documentary | 0 | 1 |

## Fit the Model

```{r}
#| echo: false
tbl_regression(model_3, intercept = TRUE) |>
  as_gt() |>
  tab_header(title = md("**Energy Level Model**"),
             subtitle = md("*What is the best predictor?*")) |>
  tab_options(table.font.size = 22)
```

## Interpreting a Categorical Predictor

::: callout-tip
## Reading the genre coefficients

-   **Drama:** drama watchers have energy `r f2(abs(b_of(model_3, "TV_GenreDrama")))` points `r ifelse(b_of(model_3, "TV_GenreDrama") < 0, "lower", "higher")` than comedy watchers (the reference), holding the other predictors constant.

-   **Documentary:** documentary watchers are `r f2(abs(b_of(model_3, "TV_GenreDocumentary")))` points `r ifelse(b_of(model_3, "TV_GenreDocumentary") < 0, "lower", "higher")` than comedy watchers.
:::

-   The reference group does not get its own coefficient. Its mean is folded into the **intercept**.

## Horror Movies

### A real dataset

> Does a film's "scare score" predict its Rotten Tomatoes score, once we account for the year it was made and whether it is a sequel?

```{r}
#| echo: false
hm_fit |>
  tbl_regression(label = list(scare_score ~ "Scare score", year ~ "Year",
                              this_film_is_a_sequel ~ "Sequel (vs. not)")) |>
  as_gt() |>
  tab_options(table.font.size = 26)
```

## Horror Movies

### Reading the model

-   The overall model is **not significant**: adjusted $R^2$ = `r f2(hm_sum$adj.r.squared)`, *F*(`r hm_sum$fstatistic[2]`, `r hm_sum$fstatistic[3]`) = `r f2(hm_sum$fstatistic[1])`, *p* `r fp(model_p(hm_fit))`.

-   None of the predictors reaches significance (the sequel coefficient is closest, *p* `r fp(p_of(hm_fit, "this_film_is_a_sequelyes"))`).

-   Not every plausible predictor works: how scary a film is does not predict how critics rate it.

## Horror Movies

### A warning about circular predictors

::: callout-important
## Be careful what goes in the model

If we predict the **scare score** from the **heart-rate measures** in the same dataset, we get a nearly perfect model: $R^2$ = `r f3(summary(hm_full)$r.squared)`.

That looks **circular**, not like a discovery: the scare score appears to be computed from those measures. Predictors that are derived from the outcome (like a movie's *ranking*) or are exact combinations of one another (like *overall difference* = average movie heart rate minus resting heart rate) also cause problems, such as **multicollinearity**.
:::

# Review Questions

## Multiple Choice

### 1. What does the slope in a simple linear regression equation represent?

A)  The value of Y when X = 0

B)  The predicted change in Y for a one-unit increase in X

C)  The strength of the correlation between X and Y

D)  The proportion of variance in Y explained by X

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer: B

The slope is the predicted change in Y for a one-unit increase in X. (A is the intercept, C is $r$, and D is $R^2$.)
:::
:::

## Multiple Choice

### 2. If a linear regression model has an R² value of 0.85, what does this mean?

A)  85% of the variability in X is explained by Y

B)  The regression model is 85% accurate

C)  85% of the variability in Y is explained by X

D)  The relationship between X and Y is statistically significant

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer: C

$R^2$ is the proportion of the variability in the outcome (Y) accounted for by the predictor(s). It says nothing by itself about statistical significance.
:::
:::

## Multiple Choice

### 3. A researcher runs a regression and finds the equation $\hat{Y} = 5 + 2X$. What does the number 5 represent?

A)  The slope of the regression line

B)  The predicted value of Y when X = 0

C)  The effect size of X on Y

D)  The p-value of the regression

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer: B

5 is the intercept: the predicted Y when X = 0. (The slope is 2.)
:::
:::

## Multiple Choice

### 4. Which of the following would indicate that a multiple regression model is overfitting?

A)  A high R² value on the training data but poor performance on new data

B)  A non-significant p-value for the intercept

C)  The presence of a negative coefficient in the model

D)  A low standard error for the slope

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer: A

An overfit model captures noise in the sample, so it looks great on the data it was fit to and predicts new data poorly.
:::
:::

## Multiple Choice

### 5. A model predicts salary from years of experience. The p-value for the slope is 0.0003. What does this mean?

A)  Years of experience does not significantly predict salary

B)  Years of experience significantly predicts salary at the α = 0.05 level

C)  The slope coefficient is 0.0003

D)  The model is not linear

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer: B

$p < .05$, so years of experience is a significant predictor of salary.
:::
:::

# Open-Ended Questions (Show Your Work)

## Open-Ended

### 6. A multiple regression model

$\hat{Y} = 3 + 1.5X_1 - 0.7X_2$

-   $X_1$ = number of hours spent studying

-   $X_2$ = number of hours spent watching TV

**a)** Predict the outcome if a student studies for 6 hours and watches TV for 2 hours.

**b)** Interpret the coefficient of $X_2$.

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer

**a)** $\hat{Y} = 3 + 1.5(6) - 0.7(2) = 3 + 9 - 1.4 = 10.6$

**b)** For each additional hour of TV, the predicted outcome **decreases** by 0.7, holding hours of studying constant.
:::
:::

## Open-Ended

### 7. A simple linear regression output

| Predictor | Coefficient | Std. Error | p-value |
|-----------|-------------|------------|---------|
| Intercept | 4.2         | .5         | .001    |
| X         | 1.8         | .3         | .0005   |

**a)** Write the equation. **b)** Interpret the slope. **c)** Predict Y if X = 10.

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer

**a)** $\hat{Y} = 4.2 + 1.8X$

**b)** For each one-unit increase in X, the predicted Y increases by 1.8, and the slope is significant ($p = .0005$).

**c)** $\hat{Y} = 4.2 + 1.8(10) = 22.2$
:::
:::

## Open-Ended

### 8. Sleep (X) and energy (Y): $\hat{Y} = 2 + 3X$

**a)** Interpret the slope.

**b)** Predict energy for a person who sleeps 7 hours.

**c)** Interpret the intercept. Is it meaningful here?

::: {.fragment}
::: {.callout-important .nonincremental}
## Answer

**a)** Each additional hour of sleep predicts 3 more points of energy.

**b)** $\hat{Y} = 2 + 3(7) = 23$

**c)** The intercept is the predicted energy for 0 hours of sleep. It is probably **not meaningful**, because nobody in the data sleeps 0 hours (that would be extrapolating).
:::
:::

## Summary: Linear Regression

-   Regression predicts an **outcome** ($Y$) from one or more **predictors**: $\hat{Y} = \beta_0 + \beta_1 A + \beta_2 B + \cdots$

-   The line of best fit **minimizes the squared residuals**.

-   Each coefficient is the predicted change in $Y$ for a one-unit increase in that predictor, **holding the others constant**.

-   $R^2$ and adjusted $R^2$ tell us how much of $y$ the model accounts for, and the F-test tells us whether the model as a whole is significant.

-   Categorical predictors become **dummy variables** compared to a reference group.

-   Check the assumptions, and watch out for **circular** or **collinear** predictors.
