---
title: "Analysis of Variance"
subtitle: "PSY 348: Lecture 12"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Analysis of Variance | ⬡⬢⬡⬢⬡⬢"
format:
  revealjs:
    theme: ../custom.scss
    scrollable: true
    width: 1600
    height: 900
    incremental: true
    touch: true
    chalkboard: true
    lightbox: true
    code-fold: true
    slide-number: c/t
    include-after-body: ../title-bg.html
    drop:
      engine: webr
revealjs-plugins:
  - drop
  - codewindow
editor:
  markdown:
    wrap: 72
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)
library(gt)

# Site tokens (see custom.scss / DESIGN_SYSTEM.md)
sage  <- "#5C715E"  # $accent
sage2 <- "#8FAF9E"  # $accent2
sage3 <- "#A7C7B5"  # $accent3
ink   <- "#2F3E46"  # $body-color
hot   <- "#B5483B"  # "look here" highlight
amber <- "#E8A33D"
grp3  <- c(sage, amber, sage3)

theme_deck <- function(base_size = 22) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(color = ink),
      axis.text = element_text(color = ink),
      plot.title = element_text(color = ink, face = "bold"),
      legend.position = "none"
    )
}

f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fp <- function(p) ifelse(p < .001, "< .001", paste("=", sub("^0", "", sprintf("%.3f", p))))
exact <- function(n, m, sd) m + sd * scale(rnorm(n))[, 1]   # exactly mean m and SD sd

# ---- Coffee Vol. 2: happiness scores, 30 per group -------------------
set.seed(2024)
cof <- tibble(
  condition = factor(rep(c("Coffee", "Placebo", "Control"), each = 30),
                     levels = c("Coffee", "Placebo", "Control")),
  score = c(exact(30, 8, 1.2), exact(30, 6, 1.2), exact(30, 5, 1.2))
)

# ---- Super Secret Limitless Drug (SSLD): 100 per group, IQ (SD = 15) --
set.seed(12)
ssld <- tibble(
  group = factor(rep(c("SSLD", "Placebo", "Control"), each = 100),
                 levels = c("SSLD", "Placebo", "Control")),
  iq = c(exact(100, 106, 15), exact(100, 100, 15), exact(100, 99, 15))
)
aov_ssld <- aov(iq ~ group, data = ssld)
tab_ssld <- tidy(aov_ssld)
F_ssld   <- tab_ssld$statistic[1]
p_ssld   <- tab_ssld$p.value[1]
df1      <- tab_ssld$df[1]
df2      <- tab_ssld$df[2]
ssb      <- tab_ssld$sumsq[1]
sse      <- tab_ssld$sumsq[2]
msb      <- tab_ssld$meansq[1]
mse      <- tab_ssld$meansq[2]
eta2     <- ssb / (ssb + sse)
fcrit    <- qf(.95, df1, df2)
ms_ssld  <- ssld |> group_by(group) |> summarise(M = mean(iq), SD = sd(iq))
tuk      <- tidy(TukeyHSD(aov_ssld))
grand    <- mean(ssld$iq)
```

## Recap

### How did we get here?

```{dot}
graph {
  node[shape = box, style = "rounded,filled", fillcolor = "white", color = "#5C715E", fontname = "Fira Sans"]
  edge[color = "#5C715E"]
  bgcolor="#E9F1EC"
  label="Independent Variable: \nCaffeine Content"
  labelloc="t"

  subgraph cluster_tt {
    bgcolor="white"
    style = rounded
    color = "#8FAF9E"
    label="T Test"

    Coffee_T [label="Coffee"]
    Decaf_T [label="Decaf"]
  }
}
```

## Recap

### How did we get here?

```{dot}
graph {
  node[shape = box, style = "rounded,filled", fillcolor = "white", color = "#5C715E", fontname = "Fira Sans"]
  edge[color = "#5C715E"]
  bgcolor="#E9F1EC"
  label="Independent Variable \nCaffeine Content"
  labelloc="t"

  subgraph cluster_tt {
    style = rounded
    color = "#8FAF9E"
    bgcolor="white"
    label="T Test"

    Coffee_T [label="Coffee"]
    Decaf_T [label="Decaf"]

    Stress_T[label="Happiness"]
    Stress_T2[label="Happiness"]

    Coffee_T -- Stress_T
    Decaf_T  -- Stress_T2
  }

  subgraph cluster_anova {
    style = rounded
    color = "#8FAF9E"
    label="ANOVA"
    bgcolor="white"
    Coffee_A [label="Coffee"]
    Decaf_A [label="Decaf"]
    Water [label="Water"]

    Stress_A1[label="Happiness"]
    Stress_A2[label="Happiness"]
    Stress_A3[label="Happiness"]

    Coffee_A -- Stress_A1
    Decaf_A -- Stress_A2
    Water -- Stress_A3
  }
}
```

## Analysis of Variance

### ANOVA

ANOVA, like t-tests, compares the means of different groups to determine
if they differ significantly from one another.

::::: columns
::: {.column width="50%"}
-   ANOVA can examine independent variables with **more than 2 groups**.

-   ANOVA can also be used for just 2 groups (then $F = t^2$ and it gives the same answer as the t-test).
:::

::: {.column width="50%"}
```{dot}
graph {
  node[shape = box, style = "rounded,filled", fillcolor = "white", color = "#5C715E", fontname = "Fira Sans"]
  edge[color = "#5C715E"]
  bgcolor="#E9F1EC"
  label="Independent Variable \nCaffeine Content"
  labelloc="t"

  subgraph cluster_anova {
    style = rounded
    color = "#8FAF9E"
    label="ANOVA"
    bgcolor="white"
    Coffee_A [label="Coffee"]
    Decaf_A [label="Decaf"]
    Water [label="Water"]

    Stress_A1[label="Happiness"]
    Stress_A2[label="Happiness"]
    Stress_A3[label="Happiness"]

    Coffee_A -- Stress_A1
    Decaf_A -- Stress_A2
    Water -- Stress_A3
  }
}
```
:::
:::::

## ANOVA

### Three types

:::::: badge-row
::: content-badge
[groups]{.material-icons}

#### Between subjects

Different participants at each level of the IV
:::

::: content-badge
[repeat]{.material-icons}

#### Within subjects

The same participants at every level (repeated measures)
:::

::: content-badge
[grid_view]{.material-icons}

#### Factorial

Levels of two or more IVs combined
:::
::::::

## ANOVA

### Between and within

```{dot}
graph {
  node[shape = box, style = "rounded,filled", fillcolor = "white", color = "#5C715E", fontname = "Fira Sans"]
  edge[color = "#5C715E"]
  center="t"
  bgcolor="#E9F1EC"
  label="ANOVA Types"
  labelloc="t"

  subgraph cluster_anova {
    style = rounded
    color = "#8FAF9E"
    label="Between Subjects ANOVA"
    bgcolor="white"
    IV

    Level1[label="IV Level 1"]
    Level2[label="IV Level 2"]
    Level3[label="IV Level 3"]

    Participants1[label = "Participants"]
    Participants2[label = "Participants"]
    Participants3[label = "Participants"]

    IV -- Participants1 -- Level1
    IV -- Participants2 --Level2
    IV -- Participants3 --Level3
  }

  subgraph cluster_wanova {
    style = rounded
    color = "#8FAF9E"
    label="Within Subjects ANOVA"
    bgcolor="white"
    ParticipantsW [label="Participants"]
    IVW [label = "IV"]

    Level1W[label="IV Level 1"]
    Level2W[label="IV Level 2"]
    Level3W[label="IV Level 3"]

    ParticipantsW -- IVW
    IVW -- Level1W
    IVW -- Level2W
    IVW -- Level3W
  }
}
```

## ANOVA

### Factorial (multiple IVs)

```{dot}
graph {
  node [shape=box, style="rounded,filled", fillcolor="white", color="#5C715E", fontname="Fira Sans"];
  edge [color="#5C715E"];
  bgcolor="#E9F1EC";
  labelloc="t";
  label="Factorial ANOVA (3 × 2 Design)";

  subgraph cluster_IV1 {
    style=rounded; color="#8FAF9E";
    label="IV₁: Type of Beverage";
    bgcolor="white";
    IV1_L1 [label="Coffee"];
    IV1_L2 [label="Decaf"];
    IV1_L3 [label="Water"];
  }

  subgraph cluster_IV2 {
    style=rounded; color="#8FAF9E";
    label="IV₂: Time of Day";
    bgcolor="white";
    IV2_L1 [label="Morning"];
    IV2_L2 [label="Evening"];
  }

  subgraph cluster_cells {
    style=rounded; color="#8FAF9E";
    label = ""
    bgcolor="white";
    C1 [label="Coffee + Morning"];
    C2 [label="Coffee + Evening"];
    C3 [label="Decaf + Morning"];
    C4 [label="Decaf + Evening"];
    C5 [label="Water + Morning"];
    C6 [label="Water + Evening"];
  }

  Outcome [label="Happiness", shape=ellipse, style=filled, fillcolor="#DCE8E1"];

  IV1_L1 -- C1;
  IV1_L1 -- C2;
  IV1_L2 -- C3;
  IV1_L2 -- C4;
  IV1_L3 -- C5;
  IV1_L3 -- C6;
  IV2_L1 -- C1;
  IV2_L2 -- C2;
  IV2_L1 -- C3;
  IV2_L2 -- C4;
  IV2_L1 -- C5;
  IV2_L2 -- C6;

  C1 -- Outcome;
  C2 -- Outcome;
  C3 -- Outcome;
  C4 -- Outcome;
  C5 -- Outcome;
  C6 -- Outcome;
}
```

## Between subjects ANOVA

### Coffee Example Vol. 2

> A researcher wants to see whether it is the taste of coffee that is
> causing higher levels of happiness. He assigns 30 people to each
> condition. The experimental group drinks caffeinated coffee. The
> placebo group drinks decaf. The control group drinks water. All `3`
> groups complete a survey to measure their happiness on a scale of `1`
> to `10`.

::::: columns
::: {.column width="50%"}
::: callout-tip
## Null hypothesis $H_0$

$\mu_\text{Coffee} = \mu_\text{Placebo} = \mu_\text{Control}$
:::
:::

::: {.column width="50%"}
::: callout-important
## Alternative hypothesis $H_A$

**Not all** the means are equal (at least one group mean differs).
:::
:::
:::::

## Coffee Example Vol. 2

### The group means

- Mean happiness: coffee `8`, placebo `6`, control `5`.

  - Is the group that drank coffee significantly happier than the other groups?

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.2
cof |>
  group_by(condition) |>
  summarise(happy = mean(score)) |>
  ggplot(aes(condition, happy, fill = condition)) +
  geom_col(color = ink, width = .6) +
  scale_fill_manual(values = grp3) +
  labs(x = "Caffeine content", y = "Happiness score") +
  theme_deck(20)
```

## Coffee Example Vol. 2

### Why not do more t-tests?

Recall that our $\alpha$ level is set at .05. If we want to find out whether coffee affects happiness, we'd need **three t-tests**:

::: callout-tip
## Three pairwise comparisons

- $t_1$: Coffee vs. Placebo
- $t_2$: Coffee vs. Control
- $t_3$: Placebo vs. Control
:::

## Why Not Do More T-Tests?

### The problem

-   Each test is done *independently*.

-   None accounts for the shared variability across all groups.

-   By running multiple t-tests, we **increase the probability of a
    false positive (Type I error)** and **ignore the combined
    structure** of the data.

-   ANOVA solves both problems by analyzing **all groups at once**,
    partitioning total variance into *between-group variance*
    (differences due to the treatment) and *within-group variance*
    (random variation).

## Why Not Do More T-Tests?

### Alpha inflation

-   Each test has a 5% chance of a false positive, and those chances **accumulate**.

::: callout-tip
## Familywise error rate

$\alpha_\text{FWE} = 1 - (1 - \alpha)^k$

For three tests: $1 - (0.95)^3 = 0.1426 \approx 14\%$
:::

:::::::: side-by-cols
::::: stat-card
::: stat-num
`r f2(100 * (1 - .95^1))`%
:::

::: stat-sub
1 test
:::
:::::

::::: stat-card
::: stat-num
`r f2(100 * (1 - .95^3))`%
:::

::: stat-sub
3 tests
:::
:::::

::::: stat-card
::: stat-num
`r f2(100 * (1 - .95^10))`%
:::

::: stat-sub
10 tests
:::
:::::
::::::::

## Why Not Do More T-Tests?

### The takeaway

> ANOVA protects against this inflation by giving us one overall test of
> group differences (F-statistic), keeping our Type I error at the
> intended 5%.

```{r}
#| echo: false
tt_row <- function(drop) {
  t.test(score ~ condition, data = cof |> filter(condition != drop) |> droplevels(),
         var.equal = TRUE) |>
    tidy() |>
    rename(Mean1 = estimate1, Mean2 = estimate2)
}
bind_rows(tt_row("Control"), tt_row("Placebo"), tt_row("Coffee")) |>
  mutate(Comparison = c("Coffee vs. Placebo", "Coffee vs. Control", "Placebo vs. Control"),
         p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) |>
  select(Comparison, D = estimate, t = statistic, df = parameter, p) |>
  gt() |>
  fmt_number(columns = c(D, t), decimals = 2) |>
  fmt_number(columns = df, decimals = 0) |>
  cols_align(align = "center") |>
  opt_row_striping() |>
  tab_options(table.width = pct(75), table.font.size = 30)
```

## Between subjects ANOVA

-   Is the variability between groups large enough, relative to the
    variability within groups, to suggest that the group means truly
    differ?

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4.4
cof |>
  ggplot(aes(condition, score, color = condition)) +
  geom_jitter(width = .15, alpha = .6, size = 2.5) +
  stat_summary(fun = mean, geom = "crossbar", width = .5, color = ink) +
  scale_color_manual(values = grp3) +
  labs(x = "Condition", y = "Score", subtitle = "dark bar = group mean") +
  theme_deck(20)
```

## Between subjects ANOVA

### Group means and variances

```{r}
#| echo: false
cof |>
  group_by(Condition = condition) |>
  summarise(Mean = mean(score), Variance = var(score), SD = sd(score)) |>
  gt() |>
  fmt_number(columns = c(Mean, Variance, SD), decimals = 2) |>
  opt_row_striping() |>
  tab_options(table.width = pct(60), table.font.size = 30)
```

## Between subjects ANOVA

### Why "analysis of variance"?

- If it's all about the means, why is it called Analysis of Variance?

- To compare the means, we analyze two types of variance:

  - Variance *within* each group, around its own mean *(within-group or error variance)*

  - Variance *between* the group means *(between-group variance)*

## The F Statistic

<!-- TODO (Dave): consider a short, fact-based ethics aside here on Fisher and eugenics. Framing: the science is not wrong; ethics are the cornerstone of any scientific endeavor. Check primary sources first. -->

::: hero-def
[F Statistic]{.term-pill}

: The ratio of **between-group variance** to **within-group (error) variance**. Named in honor of **Sir Ronald Fisher**, who developed ANOVA.
:::

## Calculating F

::: callout-tip
# ANOVA Formula

$F = \dfrac{MS_B}{MS_E} \qquad MS_B = \dfrac{SS_B}{df_B} \quad MS_E = \dfrac{SS_E}{df_E} \qquad df_B = k-1 \quad df_E = N - k$
:::

::::: term-list
::: fragment
[MS]{.term-pill}

: Mean Square: a sum of squares divided by its degrees of freedom (a variance).
:::

::: fragment
[$_B$ and $_E$]{.term-pill}

: **B**etween-group variance and within-group (**E**rror) variance.
:::

::: fragment
[$SS$, $k$, $N$]{.term-pill}

: Sum of Squares, number of groups, number of participants.
:::
:::::

## Calculating F

### Calculating the variance

ANOVA looks at variance. To calculate the variance:

::: callout-tip
# Variance

$\Large{s^2 = \dfrac{\sum(x-\bar{x})^2}{n-1}}$
:::

-   Divide by (degrees of freedom).

    -   **This is the variance**, but in ANOVA we also call it a Mean
        Square (MS).

## Calculating the Mean Square (MS)

-   $MS = \dfrac{SS}{df}$

-   Calculate the SS (sum of squared deviations).

-   Divide the SS by the degrees of freedom.

## Partitioning the variance

::: callout-tip
## Total = between + within

$SS_{Total} = SS_B + SS_E$
:::

-   $SS_B$ (between): how much variance comes from **group differences**?

    -   $SS_B = \sum n_j(\bar{x}_j - \bar{x}_{grand})^2$

-   $SS_E$ (error, within): how much variance is **random variation within the groups**?

    -   $SS_E = \sum\sum (x_{ij} - \bar{x}_j)^2$

## Degrees of Freedom

### The group project

> Imagine a group of 5 students is submitting a project, and the total
> score for the group must be 100 points combined.

-   The first `4` students can choose any score they want, say `15`,
    `20`, `25`, and `10`.

-   But the 5th student's score isn't free anymore. It must be whatever
    makes the total = `100`. In other words, only 4 scores are "free to
    vary."

-   That's `n − 1 = 4` degrees of freedom.

## Degrees of Freedom

::: callout-tip
## Rules for a one-way ANOVA

-   Between groups: $df_{B} = k - 1$

-   Within groups: $df_E = N - k$

-   Total: $df_{Total} = N - 1$

($k$ is the number of groups in the independent variable.)
:::

## Calculating the Mean Squares and F

::: callout-tip
## The F ratio

$MS_B = \dfrac{SS_B}{df_B} \qquad MS_E = \dfrac{SS_E}{df_E} \qquad F = \dfrac{MS_B}{MS_E}$
:::

## Between subjects ANOVA Example

> `100` participants take the **Super Secret Limitless Drug (SSLD)**, `100` take a **placebo**, and `100` take **nothing**. All complete an IQ test. Do those who took the SSLD have significantly higher IQ than the placebo and control groups?

-   What is the IV?

    -   Drug taken

-   How many levels does it have?

    -   3 (SSLD, Placebo, Nothing)

-   What is the DV?

    -   IQ score

## Between subjects ANOVA Example

### Group means

```{r}
#| echo: false
ms_ssld |>
  rename(Group = group) |>
  gt() |>
  fmt_number(columns = c(M, SD), decimals = 2) |>
  opt_row_striping() |>
  tab_options(table.font.size = 30, table.width = pct(50))
```

## ANOVA in R

### Fitting the model

::: {.codewindow .r .codewindow-dark}
anova.R

```{r}
#| echo: true
#| code-fold: false
# aov(DV ~ IV, data): the group column must be a factor
fit <- aov(iq ~ group, data = ssld)

summary(fit)
```
:::

-   `Df`: degrees of freedom ($k - 1$ between, $N - k$ within)

-   `Sum Sq`, `Mean Sq`: the SS and MS ($MS = SS/df$)

-   `F value`: $MS_B / MS_E$ and `Pr(>F)`: its p-value

## ANOVA in R

### The ANOVA table

```{r}
#| echo: false
tab_ssld |>
  mutate(term = c("Between groups", "Within groups (error)")) |>
  rename(Source = term, SS = sumsq, MS = meansq, F = statistic, p = p.value) |>
  gt() |>
  fmt_number(columns = c(SS, MS, F), decimals = 2) |>
  fmt_number(columns = df, decimals = 0) |>
  fmt_number(columns = p, decimals = 3) |>
  sub_missing(missing_text = "") |>
  opt_row_striping() |>
  tab_options(table.font.size = 30, table.width = pct(85)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels())
```

::: callout-tip
## By hand

$SS_B = 100(`r f2(ms_ssld$M[1])` - `r f2(grand)`)^2 + 100(`r f2(ms_ssld$M[2])` - `r f2(grand)`)^2 + 100(`r f2(ms_ssld$M[3])` - `r f2(grand)`)^2 = `r f2(ssb)`$

$MS_B = \dfrac{`r f2(ssb)`}{`r df1`} = `r f2(msb)`$, $MS_E = \dfrac{`r f2(sse)`}{`r df2`} = `r f2(mse)`$, so $F = \dfrac{`r f2(msb)`}{`r f2(mse)`} = `r f2(F_ssld)`$
:::

## Interpreting F

$MS_B$ is made up of $MS_E$ + the theoretical difference between groups:

-   $MS_B = \text{group difference} + MS_E$

-   $MS_B = \text{group difference (0)} + MS_E$

-   $MS_B = MS_E$

-   $F = \dfrac{MS_B}{MS_E}$

-   $F = 1$

## Interpreting F

-   When $F < 1$, it's not likely to be significant.

-   When there **is** a group difference:

    -   $MS_B = \text{(Number > 0)} + MS_E$

    -   $F = \dfrac{\text{Difference} + MS_E}{MS_E} = 1 + \dfrac{\text{Difference}}{MS_E}$

    -   The bigger the group difference relative to the error, the bigger $F$ gets (larger than 1).

## The F Distribution

### Is our F big enough?

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4.2
fx <- seq(0, 8, length.out = 800)
fd <- tibble(x = fx, y = df(fx, df1, df2))
ggplot(fd, aes(x, y)) +
  geom_area(data = filter(fd, x >= fcrit), fill = hot, alpha = .75) +
  geom_line(color = sage, linewidth = 1.4) +
  geom_vline(xintercept = fcrit, linetype = "dotted", color = ink, linewidth = 1.1) +
  geom_vline(xintercept = F_ssld, color = hot, linewidth = 1.3) +
  annotate("text", x = fcrit, y = .9, label = paste0("critical F = ", f2(fcrit)),
           hjust = 1.05, color = ink, size = 6) +
  annotate("text", x = F_ssld, y = .75, label = paste0("observed F = ", f2(F_ssld)),
           hjust = -.05, color = hot, size = 6) +
  labs(x = paste0("F (", df1, ", ", df2, ")"), y = NULL,
       subtitle = "red area = the 5% of F values we would call unusual under H0") +
  theme_deck(20) +
  theme(axis.text.y = element_blank())
```

- The F distribution is **right-skewed** and never negative.

- In R: `qf(.95, df1 = 2, df2 = 297)` gives the critical value `r f2(fcrit)`. Since $`r f2(F_ssld)` > `r f2(fcrit)`$, we **reject** $H_0$.

## Hypothesis testing: ANOVA

::: callout-tip
## Decision rule

-   $F \le 1$: fail to reject the null hypothesis.

-   $F > 1$: compare to the critical value (or use the p-value). Reject $H_0$ only if $F > F_{critical}$, equivalently if $p < \alpha$.
:::

## Assumptions of One-Way ANOVA

::: callout-important
## Assumptions

1.  **The DV is continuous** (interval or ratio scale).
2.  **Observations are independent**, and each person is in only one group.
3.  **Normality:** the scores (residuals) in each group are approximately normal.
4.  **Homogeneity of variance:** the groups have similar variances (Levene's Test).
:::

::: {.codewindow .r .codewindow-dark}
assumptions.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
library(car)

# Equal variances? (H0: the variances are equal)
leveneTest(iq ~ group, data = ssld)

# Normal residuals? (H0: the residuals are normal)
shapiro.test(residuals(fit))
```
:::

## Effect Size: Eta Squared

### How much of the variance is explained by the groups?

::: callout-tip
## $\eta^2$

$\eta^2 = \dfrac{SS_B}{SS_{Total}} = \dfrac{`r f2(ssb)`}{`r f2(ssb + sse)`} = `r f3(eta2)`$
:::

```{r}
#| echo: false
tibble(
  `η²` = c("0.01", "0.06", "0.14"),
  Interpretation = c("Small effect", "Medium effect", "Large effect")
) |>
  gt() |>
  opt_row_striping() |>
  tab_options(table.width = pct(50), table.font.size = 22)
```

- Our groups explain about `r f2(100 * eta2)`% of the variance in IQ: a **small-to-medium** effect.

## Post-hoc Tests

### A significant F tells us *some* means differ, not *which*

::: {.codewindow .r .codewindow-dark}
tukey.R

```{r}
#| echo: true
#| code-fold: false
# Tukey's HSD compares every pair and adjusts for multiple comparisons
TukeyHSD(fit)
```
:::

-   `diff` is the difference between the two group means.

-   `p adj` is adjusted for the number of comparisons, so the familywise error rate stays at 5%.

## Post-hoc Tests

### Reading Tukey's HSD

```{r}
#| echo: false
tuk |>
  mutate(Comparison = contrast,
         p = ifelse(adj.p.value < .001, "<.001", sprintf("%.3f", adj.p.value))) |>
  select(Comparison, Difference = estimate, `CI low` = conf.low,
         `CI high` = conf.high, `p adj` = p) |>
  gt() |>
  fmt_number(columns = c(Difference, `CI low`, `CI high`), decimals = 2) |>
  opt_row_striping() |>
  tab_options(table.font.size = 28, table.width = pct(80))
```

-   A pair is significant when its **p adj** is below .05 (or its confidence interval does not include 0).

## Reporting F

> If asked to report findings in terms of the Null Hypothesis (H0), you
> should report findings as:

-   Reject $H_0$, or

-   Fail to Reject $H_0$

## Reporting F

If asked to report findings in general or for publication, you need to
report 5 things:

-   $F$(df for the between group MS, df for the error MS)

-   $F$-value

-   $p$-value

-   Mean and standard deviation of each group

-   An effect size ($\eta^2$)

## Reporting F

> IQ scores differed significantly across the three groups, *F*(`r df1`, `r df2`) = `r f2(F_ssld)`, *p* `r fp(p_ssld)`, $\eta^2$ = `r f2(eta2)`. The SSLD group had the highest mean (*M* = `r f2(ms_ssld$M[1])`, *SD* = `r f2(ms_ssld$SD[1])`), followed by the placebo group (*M* = `r f2(ms_ssld$M[2])`, *SD* = `r f2(ms_ssld$SD[2])`) and the control group (*M* = `r f2(ms_ssld$M[3])`, *SD* = `r f2(ms_ssld$SD[3])`). Tukey's HSD showed that the SSLD group scored higher than the placebo group (*p* `r fp(tuk$adj.p.value[tuk$contrast == "Placebo-SSLD"])`) and the control group (*p* `r fp(tuk$adj.p.value[tuk$contrast == "Control-SSLD"])`), which did not differ from each other (*p* `r fp(tuk$adj.p.value[tuk$contrast == "Control-Placebo"])`).

## Reporting F

> There was a significant difference in IQ between the groups, *F*(`r df1`, `r df2`) = `r f2(F_ssld)`, *p* `r fp(p_ssld)`.

## Identifying Degrees of Freedom

### One-way ANOVA, equal groups

> A researcher compares three different study methods (flashcards,
> rereading, and practice testing) on exam performance. There are 15
> students in each group.

-   What is $df_B$?

    -   $k-1 = 3-1 = 2$

-   What is $df_E$?

    -   $N = 15 \times 3 = 45$

    -   $df_E = N-k = 45-3 = 42$

## Identifying Degrees of Freedom

### One-way ANOVA, unequal groups

> A psychologist compares four types of therapy for anxiety. Group sizes
> are `10`, `12`, `14`, and `9` participants.

-   What is $df_B$?

    -   $k-1 = 4-1 = 3$

-   What is $df_E$?

    -   $N = 10+12+14+9 = 45$, so $df_E = N-k = 45-4 = 41$

## Summary: One-Way ANOVA

-   Compares **three or more** group means with **one** overall test.

-   $F = MS_B / MS_E$: between-group variance relative to within-group variance.

-   Follow a significant $F$ with a **post-hoc test** (Tukey's HSD) to find which groups differ.

-   Check assumptions: independence, normality, equal variance.

-   Report $F$(df, df), $p$, group means and SDs, and an effect size ($\eta^2$).
