---
title: "Lab: Z-Scores & Probability in Practice"
subtitle: "Standardizing Data and Finding Probabilities"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "PSY 348 - Statistics for Psychology"
format:
  revealjs:
    theme: ../custom.scss
    width: 1600
    height: 900
    incremental: true
    touch: true
    chalkboard: true
---

## Quick Recap

$$z = \frac{x - \bar{x}}{s}$$

::: highlight-last
-   A z-score tells you how many standard deviations a value is from the mean

-   Once we have a z-score, we can find **probabilities** — what percentage of scores fall above, below, or between values
:::

## Scenario

> Professor Brocker's PSY 348 midterm had `40` students. The exam scores were approximately normally distributed with a mean of `76` and a standard deviation of `8`.

::: highlight-last
-   Nadia scored an `88`.

-   Before we calculate anything — do you think that's a good score relative to the class? Great? Just okay?
:::

## By Hand First

$$z = \frac{x - \bar{x}}{s} = \frac{88 - 76}{8} = 1.5$$

::: highlight-last
-   Nadia's score is **1.5 standard deviations above the mean**

-   Using the 68/95/99.7 rule: roughly 93% of students scored *below* Nadia
:::

## Now in R

```{r}
#| echo: true
library(dplyr)

# Simulate a class of 40 exam scores
set.seed(348)
exam_scores <- rnorm(40, mean = 76, sd = 8) |> round(1)

# Nadia's score
nadia <- 88

# Z-score by hand vs. by function
z_hand <- (nadia - mean(exam_scores)) / sd(exam_scores)
z_hand
```

::: highlight-last
-   Notice: this won't be *exactly* 1.5 — why not? (Hint: think about what `set.seed()` and `rnorm()` are actually doing.)
:::

## Standardizing an Entire Dataset

R can convert every score to a z-score at once with `scale()`:

```{r}
#| echo: true
z_scores <- scale(exam_scores)[,1] |> round(2)

data.frame(
  score = exam_scores,
  z = z_scores
) |>
  head(8)
```

-   What do you notice about the *mean* of `z_scores`? The *standard deviation*?

## From Z-Score to Probability

R's `pnorm()` gives the probability of scoring *at or below* a given z-score, using the theoretical normal curve (not just our 40 students).

```{r}
#| echo: true
# Probability of scoring 88 or LOWER
pnorm(88, mean = 76, sd = 8)
```

::: highlight-last
-   About `pnorm(1.5)` = 93.3% of the theoretical distribution falls at or below a z of 1.5

-   So only about **6.7%** of students would be expected to outscore Nadia
:::

## Visualizing It

```{r}
library(ggplot2)

dat <- data.frame(x = seq(50, 102, by = 0.1))
dat$y <- dnorm(dat$x, mean = 76, sd = 8)

ggplot(dat, aes(x, y)) +
  geom_line(linewidth = 1.1) +
  geom_area(
    data = subset(dat, x >= 88),
    fill = "orange", alpha = 0.6
  ) +
  geom_vline(xintercept = 88, lty = "dashed") +
  theme_minimal(base_size = 18) +
  labs(
    title = "Probability of Scoring 88 or Higher",
    x = "\nExam Score",
    y = "Density\n"
  ) +
  theme(panel.grid = element_blank())
```

## Going the Other Direction

`qnorm()` does the reverse — give it a probability, get back a score.

```{r}
#| echo: true
# What score marks the top 10% of the class?
qnorm(0.90, mean = 76, sd = 8)
```

::: highlight-last
-   A student would need roughly an `86` to be in the top 10% of this class
:::

## Your Turn

Using `exam_scores` from earlier in this lab:

1.  Calculate the z-score for a student who scored a `70`. Is that above or below average?

2.  Using `pnorm()`, what percentage of the theoretical distribution scored below a `70`?

3.  Using `qnorm()`, what raw score marks the bottom 25% of the distribution?

4.  In your own words: what's the difference between using `mean(exam_scores)`/`sd(exam_scores)` (our actual class) versus plugging in `76` and `8` directly into `pnorm()`/`qnorm()`?

## Submit on Brightspace

-   Your answers to the four questions above (show your R code and output)

-   One sentence: is a z-score of 0 a "good" or "bad" score? Explain.
