---
title: "Z-Scores and Probability Under the Normal Curve"
subtitle: "PSY 348: Lecture 6"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Z-Scores & Probability | ⬡⬢⬡⬢⬡⬢"
resources:
  - assets/standardnormaltable.pdf
format:
  revealjs:
    theme: ../custom.scss
    incremental: true
    touch: true
    chalkboard: true
    scrollable: true
    lightbox: true
    include-after-body: ../title-bg.html
    slide-number: c/t
    width: 1600
    height: 900
    drop:
      engine: webr
      webr:
        packages:
         - ggplot2
         - dplyr
revealjs-plugins:
  - drop
  - codewindow
---

```{r}
#| include: false
# Math colors follow lec4's symbol-coloring convention (trial, 9/21):
#   x-bar (mean) -> blue, s (standard deviation) -> orange.
#   Plain x, z, and raw numbers stay uncolored.
# NOTE: lec5's formula slide uses a different mapping (x orange, x-bar red,
# s pink) -- pick one before promoting either to DESIGN_SYSTEM.md.
library(dplyr)
library(ggplot2)
library(ggiraph)

# Site tokens (see custom.scss / DESIGN_SYSTEM.md)
sage   <- "#5C715E"  # $accent
sage2  <- "#8FAF9E"  # $accent2
sage3  <- "#A7C7B5"  # $accent3
ink    <- "#2F3E46"  # $body-color
hot    <- "#B5483B"  # "look here" highlight -- deliberate exception, like lec4's red mean line

# One normal curve, drawn from the theoretical density (not simulated data),
# so the dashed mean line and the z-axis land exactly where the slide says.
# shade = c(lower, upper) fills that region; marks = raw scores to draw lines at.
curve_plot <- function(mu, sigma, shade = NULL, marks = NULL,
                       xlab = "Exam Score") {
  d <- tibble(
    x = seq(mu - 4 * sigma, mu + 4 * sigma, length.out = 800),
    y = dnorm(x, mu, sigma)
  )
  p <- ggplot(d, aes(x, y))
  if (!is.null(shade)) {
    p <- p + geom_area(
      data = filter(d, x >= shade[1], x <= shade[2]),
      fill = sage2, alpha = .7
    )
  }
  p <- p +
    geom_line(color = sage, linewidth = 1.4) +
    geom_vline(xintercept = mu, linetype = "dashed", color = ink) +
    scale_x_continuous(
      breaks = seq(mu - 4 * sigma, mu + 4 * sigma, by = sigma),
      sec.axis = sec_axis(~ (. - mu) / sigma, name = "z", breaks = -4:4)
    ) +
    labs(x = xlab, y = NULL) +
    theme_minimal(base_size = 22) +
    theme(
      axis.text.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.title = element_text(color = ink),
      axis.text.x = element_text(color = ink)
    )
  if (!is.null(marks)) {
    p <- p + geom_vline(xintercept = marks, color = hot, linewidth = 1.2)
  }
  p
}
```

## Review

### Proportional Breakdown

:::::::::::: side-by-cols
::::: stat-card
::: stat-num
68.3%
:::

::: stat-sub
of scores fall within **1 SD** of the mean
:::
:::::

::::: stat-card
::: stat-num
95.4%
:::

::: stat-sub
of scores fall within **2 SD** of the mean
:::
:::::

::::: stat-card
::: stat-num
99.7%
:::

::: stat-sub
of scores fall within **3 SD** of the mean
:::
:::::
::::::::::::

## Review

### What's a z-score again?

::: hero-def

[Z-Score]{.term-pill}

: How many **standard deviations** a single score (an $x$) sits above or below the mean.
:::

## Z-scores

### Why do we care?

Z-scores re-express the original data points (the $x$'s) so we can see, at a glance:

- How close that $x$ is to the mean
- Where it falls in the spread of the distribution
- How similar this participant is to the rest of the sample
- How different this participant is from the majority of the sample

## Z-scores

### Why do we care?

```{r}
#| fig-width: 12
#| fig-height: 6.5
set.seed(616)
z1 <- tibble(
  student = 1:1000,
  x = rnorm(1000, 75, 2) |> round(2),
  z = (x - mean(x)) / sd(x)
) |>
  mutate(
    band = case_when(
      abs(z) <= 1 ~ "Within 1 SD",
      abs(z) <= 2 ~ "1-2 SD",
      abs(z) <= 3 ~ "2-3 SD",
      TRUE        ~ "Beyond 3 SD"
    ),
    band = factor(band, levels = c("Within 1 SD", "1-2 SD", "2-3 SD", "Beyond 3 SD"))
  )

z1 |>
  ggplot(aes(z, student, color = band)) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("Within 1 SD" = sage, "1-2 SD" = sage2, "2-3 SD" = "#D9A75B", "Beyond 3 SD" = hot),
    name = NULL
  ) +
  scale_x_continuous(breaks = -4:4) +
  theme_minimal(base_size = 22) +
  theme(panel.grid.minor = element_blank(), legend.position = "top") +
  labs(x = "\nZ-Score", y = "Student\n")
```

## Z-scores

### Who's more than 2 SD *below* the mean?

```{r}
#| fig-width: 12
#| fig-height: 6.5
z1 |>
  ggplot(aes(z, student)) +
  geom_point(aes(color = ifelse(z <= -2, hot, "grey75")), size = 2) +
  geom_vline(xintercept = -2, color = ink, linetype = "dashed") +
  scale_color_identity() +
  scale_x_continuous(breaks = -4:4) +
  theme_minimal(base_size = 22) +
  theme(panel.grid.minor = element_blank()) +
  labs(x = "\nZ-Score", y = "Student\n")
```

## Z-scores

### Going backwards: Imagine that $\color{orange}{s} = 2$ and $\color{blue}{\bar{x}} = 25$

::: callout-tip
## Formula

$$x = \color{blue}{\bar{x}} + z\,\color{orange}{s}$$
:::

- $z = 1 \Rightarrow x = 25 + (1)(2) = 27$
- $z = -2 \Rightarrow x = 25 + (-2)(2) = 21$
- $z = 0.5 \Rightarrow x = 25 + (0.5)(2) = 26$

## Calculating a Z-score

### Step-by-Step

::: callout-tip
## Two steps

1.  **Mean-center** the $x$ values: $x - \color{blue}{\bar{x}}$ — this makes the mean $= 0$.
2.  **Divide** by the standard deviation: $\dfrac{x - \color{blue}{\bar{x}}}{\color{orange}{s}}$ — this makes the SD $= 1$.
:::

## Calculating Probability

### Use the normal distribution

> The average grade on the exam was an `86` with a standard deviation of `4`.

```{r}
#| fig-width: 13
#| fig-height: 4.4
curve_plot(86, 4)
```

## Calculating Probability

### What is the probability of scoring a `94` or higher?

> The average grade on the exam was an `86` with a standard deviation of `4`.

```{r}
#| fig-width: 13
#| fig-height: 4.4
curve_plot(86, 4, shade = c(94, Inf), marks = 94)
```

## Calculating Probability

### What is the probability of scoring a `94` or higher?

::: callout-tip
## Solution

- $z = \dfrac{x-\color{blue}{\bar{x}}}{\color{orange}{s}} = \dfrac{94-86}{4} = \dfrac{8}{4} = 2$
- Beyond $z = 2$ is the `2.1%` + `0.1%` slivers of the curve ≈ **`2.3%`** chance.
:::

## Example 2

### Test Time!

> The mean exam score was `86` with a standard deviation of `4`. What is the probability of scoring between an `82` and a `90`?

```{r}
#| fig-width: 13
#| fig-height: 4.4
curve_plot(86, 4)
```

## Example 2

### Test Time!

> The mean exam score was `86` with a standard deviation of `4`. What is the probability of scoring between an `82` and a `90`?

```{r}
#| fig-width: 13
#| fig-height: 4.4
curve_plot(86, 4, shade = c(82, 90), marks = c(82, 90))
```

## Example 2

### Test Time!

::: callout-tip
## Solution

- $z = \dfrac{82-86}{4} = \dfrac{-4}{4} = -1$
- $z = \dfrac{90-86}{4} = \dfrac{4}{4} = 1$
- Between $z = -1$ and $z = 1$: a **`68.3%`** chance of scoring between `82` and `90`.
:::

## Example 3

### Esmeralda's Second Test

> Esmeralda's stats professor tells the class that the average score on the exam was a `72` with a standard deviation of `6`, and the distribution of scores was normal. Esmeralda wants to calculate the probability that she scored below `60`.

```{r}
#| fig-width: 14
#| fig-height: 3.2
curve_plot(72, 6)
```

## Example 3

### Esmeralda's Second Test

> Esmeralda's stats professor tells the class that the average score on the exam was a `72` with a standard deviation of `6`, and the distribution of scores was normal. Esmeralda wants to calculate the probability that she scored below `60`.

```{r}
#| fig-width: 14
#| fig-height: 3.2
curve_plot(72, 6, shade = c(-Inf, 60), marks = 60)
```

## Example 3

### Esmeralda's Second Test: $\color{blue}{\bar{x}} = 72 \mid \color{orange}{s} = 6$

::: callout-tip
## Solution

- $z = \dfrac{60-72}{6} = \dfrac{-12}{6} = -2$
- About a **`2.3%`** chance that she failed.
- What's the chance she passed?
- $100\% - 2.3\% = 97.7\%$
:::

## Z-scores and Probability

### How does it apply?

In real life, we are usually working with numbers with long decimals rather than nice whole numbers.

- Because the standard normal curve is *standard* and *normal*, we can look up the **exact** probability for a z-score with decimals.

## Z-scores and Probability

### The Z-Table

::::: columns
::: {.column width="45%"}
- Locate the whole number and first decimal along the **left side** of the table.
- Locate the second decimal along the **top** of the table.
- Where they meet is the proportion of the curve at or **below** that z.
:::

::: {.column width="55%"}
{{< pdf file="assets/standardnormaltable.pdf" height="600px" download-link="false">}}
:::
:::::

## Z-scores and Probability

### The Z-Table

::: callout-important
## Rule

The z-table gives you the probability of that z-score **or less** (the area to the *left*).
:::

- If you need that z-score **or higher**, subtract the table value from 1.
  - Esmeralda: `.023` chance of failing, so a `1 - .023 = .977` → **`97.7%`** chance of passing.

## Z-scores and Probability

### Life Satisfaction

> In a recent study on world happiness, participants were asked to evaluate their current lives on a scale from `0` to `10`, where `0` represents the worst possible life and `10` represents the best possible life. The responses were normally distributed, with a mean of `5.4` and a standard deviation of `2.2`. Find the probability that a randomly selected study participant's response was:

::: {.incremental .highlight-last}
- Less than `4`
- More than `8`
:::

## Z-scores and Probability

### Life Satisfaction: $\color{blue}{\bar{x}} = 5.4 \mid \color{orange}{s} = 2.2$

::: callout-tip
## Convert to z first

- **Less than 4:** $z = \dfrac{4-5.4}{2.2} = \dfrac{-1.4}{2.2} = -0.64$
- **More than 8:** $z = \dfrac{8-5.4}{2.2} = \dfrac{2.6}{2.2} = 1.18$
:::

## Life Satisfaction {.z-slide}

### Less than 4 ($z = -0.64$)

::: z-excerpt
|    z     |  .02   |  .03   |       .04        |  .05   |  .06   |
|:--------:|:------:|:------:|:----------------:|:------:|:------:|
|   -0.5   | 0.3015 | 0.2981 |      0.2946      | 0.2912 | 0.2877 |
| **-0.6** | 0.2676 | 0.2643 | [0.2611]{.z-hit} | 0.2578 | 0.2546 |
|   -0.7   | 0.2358 | 0.2327 |      0.2296      | 0.2266 | 0.2236 |

: Excerpt of the z-table (area to the left of z)
:::

::::: panel-tabset
### By Hand

- Find **-0.6** on the left side of the table — use the negative-z page.
- Find the **.04** column.
- Where they meet is the area to the left — that **is** the probability we want.
- Report it as a percentage (the proportion $\times 100$).

<details>

<summary>Reveal Answer</summary>

::: nonincremental
- $.2611 \times 100 =$ **26.1%** of participants rated their life below a 4.
:::

</details>

### In R

::: {.codewindow .r .codewindow-dark}
pnorm.R

``` r
pnorm(-0.64)                        # .2611 (z-table)
pnorm(4, mean = 5.4, sd = 2.2)      # .2623 (unrounded z)
```
:::

::: {.callout-tip appearance="minimal"}
**`pnorm`** (**p**robability) is the z-table: z in, area to the left out. **`qnorm`** (**q**uantile) is the reverse: probability in, z out.
:::

:::::

## Life Satisfaction {.z-slide}

### More than 8 ($z = 1.18$)

::: z-excerpt
|    z    |  .06   |  .07   |       .08        |  .09   |
|:-------:|:------:|:------:|:----------------:|:------:|
|   1.0   | 0.8554 | 0.8577 |      0.8599      | 0.8621 |
| **1.1** | 0.8770 | 0.8790 | [0.8810]{.z-hit} | 0.8830 |
|   1.2   | 0.8962 | 0.8980 |      0.8997      | 0.9015 |

: Excerpt of the z-table (area to the left of z)
:::

::::: panel-tabset
### By Hand

- Find **1.1** on the left side of the table.
- Find the **.08** column.
- Note where they intersect.
- We want **8 or more**, so subtract that proportion from 1.
- Report it as a percentage (the proportion $\times 100$).

<details>

<summary>Reveal Answer</summary>

::: nonincremental
- $1 - .8810 = .1190$
- **11.9%** of participants rated their life above an 8.
:::

</details>

### In R

::: {.codewindow .r .codewindow-dark}
pnorm.R

``` r
1 - pnorm(1.18)                                              # .1190 (z-table)
pnorm(8, mean = 5.4, sd = 2.2, lower.tail = FALSE)           # .1186 (unrounded z)
```
:::

::: {.callout-tip appearance="minimal"}
**`pnorm`** (**p**robability) is the z-table: z in, area to the left out. **`qnorm`** (**q**uantile) is the reverse: probability in, z out.
:::

:::::

## Z-scores and Probability

### MENSA

> Scores on an IQ test are approximately normal with a mean of `100` and a standard deviation of `15`. The organization MENSA, which calls itself the "high IQ society," requires a score of `130` or higher.

- What percent of adults would qualify for membership?
  - Find the probability of scoring `130` or higher.
  - $z = \dfrac{x-\color{blue}{\bar{x}}}{\color{orange}{s}} = \dfrac{130-100}{15} = \dfrac{30}{15} = 2$

## MENSA {.z-slide}

### $\color{blue}{\bar{x}} = 100 \mid \color{orange}{s} = 15$ — `130` or higher ($z = 2.00$)

::: z-excerpt
|    z    |       .00        |  .01   |  .02   |
|:-------:|:----------------:|:------:|:------:|
|   1.9   |      0.9713      | 0.9719 | 0.9726 |
| **2.0** | [0.9772]{.z-hit} | 0.9778 | 0.9783 |
|   2.1   |      0.9821      | 0.9826 | 0.9830 |

: Excerpt of the z-table (area to the left of z)
:::

::::: panel-tabset
### By Hand

- Find **2.0** on the left side of the table.
- Find the **.00** column.
- Note where they intersect.
- We want **130 or more**, so subtract that proportion from 1.
- Report it as a percentage (the proportion $\times 100$).

<details>

<summary>Reveal Answer</summary>

::: nonincremental
- $1 - .9772 = .0228$
- About **2.3%** of adults would qualify for MENSA.
:::

</details>

### In R

::: {.codewindow .r .codewindow-dark}
pnorm.R

``` r
1 - pnorm(2)                                                 # .0228 (z-table)
pnorm(130, mean = 100, sd = 15, lower.tail = FALSE)          # .0228
```
:::

::: {.callout-tip appearance="minimal"}
**`pnorm`** (**p**robability) is the z-table: z in, area to the left out. **`qnorm`** (**q**uantile) is the reverse: probability in, z out.
:::

:::::

## Z-table clarity {.z-slide}

### The table only ever looks **left**

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 3.4
# Hover-to-shade: every slice at or left of the cursor fills in, and the
# tooltip reads out the z-table value. Shade slices (s_i) and invisible
# full-height hit targets (h_i) share an index; the JS below fills s_j for
# all j <= hovered i. Slices are .05 wide, so every z shown is a real table row.
zs <- seq(-3.5, 3.5, by = .05)
sl <- tibble(
  i    = seq_len(length(zs) - 1),
  xmin = head(zs, -1), xmax = tail(zs, -1),
  z    = round(xmax, 2),
  ymax = dnorm((xmin + xmax) / 2),
  tip  = sprintf("z = %.2f\nTable: %.4f  (area to the left)\nRight tail: 1 - %.4f = %.4f",
                 z, pnorm(z), pnorm(z), 1 - pnorm(z))
)
g <- ggplot(sl) +
  geom_rect_interactive(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = ymax,
                            data_id = paste0("s_", i)),
                        fill = "white", color = NA) +
  geom_line(data = tibble(x = seq(-3.5, 3.5, length.out = 400), y = dnorm(x)),
            aes(x, y), color = sage, linewidth = 1.4) +
  geom_rect_interactive(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = .42,
                            data_id = paste0("h_", i), tooltip = tip),
                        fill = "#ffffff", alpha = .01, color = NA) +
  scale_x_continuous(breaks = -3:3) +
  labs(x = "z", y = NULL) +
  theme_minimal(base_size = 22) +
  theme(axis.text.y = element_blank(), panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title = element_text(color = ink), axis.text.x = element_text(color = ink))
w <- girafe(ggobj = g, width_svg = 11, height_svg = 3.4, bg = "transparent",
            options = list(
              opts_tooltip(css = paste0("background:", ink, ";color:white;padding:8px 12px;",
                                        "border-radius:8px;font-size:18px;white-space:pre;"),
                           use_cursor_pos = TRUE, offx = 14, offy = -10),
              opts_hover(css = ""), opts_hover_inv(css = ""),
              opts_toolbar(saveaspng = FALSE), opts_sizing(rescale = TRUE, width = 1)))
w <- htmlwidgets::onRender(w, sprintf("function(el) {
  var fill = '%s';
  function paint(i) {
    el.querySelectorAll('[data-id^=\"s_\"]').forEach(function(n) {
      var j = +n.getAttribute('data-id').slice(2);
      n.style.fill = (i !== null && j <= i) ? fill : 'white';
      n.style.fillOpacity = (i !== null && j <= i) ? .75 : 0;
    });
  }
  el.addEventListener('mouseover', function(e) {
    var id = e.target.getAttribute && e.target.getAttribute('data-id');
    if (id && id.indexOf('h_') === 0) paint(+id.slice(2));
  });
  el.addEventListener('mouseleave', function() { paint(null); });
}", sage2))
w
```

::: {.callout-tip appearance="minimal"}
**Hover over the curve.** Shaded = the table value. **Left of z:** read it directly. **Right of z:** 1 minus the table value. **Between:** big area minus small area. The "1 minus" applies to the *area*, never to $z$.
:::

## Z-scores

### "Compared to what?" is everywhere

```{r}
#| include: false
set.seed(348)
ex_plot <- function(mu, sd, val, xlab, std = FALSE, n = 500) {
  x <- rnorm(n, mu, sd)
  zv <- (val - mu) / sd
  d <- if (std) tibble(x = (x - mu) / sd) else tibble(x = x)
  v <- if (std) zv else val
  ggplot(d, aes(x)) +
    geom_histogram(bins = 30, fill = sage3, color = "white") +
    geom_vline(xintercept = v, color = hot, linewidth = 1.4) +
    scale_x_continuous(n.breaks = 4) +
    labs(x = if (std) "z" else xlab, y = NULL) +
    theme_minimal(base_size = 22) +
    theme(axis.text.y = element_blank(), panel.grid = element_blank(),
          axis.title = element_text(color = ink), axis.text.x = element_text(color = ink))
}
```

:::: z-cards
::: z-card
::: z-card-title
Newborn weight
:::

::: z-card-q
Is a 4.3 kg baby big? ($\mu = 3.4$, $\sigma = 0.5$)
:::

::: panel-tabset
### Raw (kg)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(3.4, .5, 4.3, "Birth weight (kg)")
```

### z-score

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(3.4, .5, 4.3, "", std = TRUE)
```

$z = \dfrac{4.3-3.4}{0.5} = 1.8$. Growth charts report exactly this.
:::
:::

::: z-card
::: z-card-title
Daily screen time
:::

::: z-card-q
Is 7 hours a lot? ($\mu = 4.5$, $\sigma = 1.5$)
:::

::: panel-tabset
### Raw (hours)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(4.5, 1.5, 7, "Hours per day")
```

### z-score

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(4.5, 1.5, 7, "", std = TRUE)
```

$z = \dfrac{7-4.5}{1.5} = 1.67$
:::
:::

::: z-card
::: z-card-title
Bolt diameter
:::

::: z-card-q
Is 10.10 mm out of spec? ($\mu = 10.00$, $\sigma = 0.05$)
:::

::: panel-tabset
### Raw (mm)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(10, .05, 10.1, "Diameter (mm)")
```

### z-score

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(10, .05, 10.1, "", std = TRUE)
```

$z = \dfrac{10.10-10.00}{0.05} = 2.0$. Quality control flags this.
:::
:::

::: z-card
::: z-card-title
One-day stock return
:::

::: z-card-q
Is a $-2.15\%$ day unusual? ($\mu = 0.05\%$, $\sigma = 1.1\%$)
:::

::: panel-tabset
### Raw (%)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(.05, 1.1, -2.15, "Daily return (%)")
```

### z-score

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 1.9
ex_plot(.05, 1.1, -2.15, "", std = TRUE)
```

$z = \dfrac{-2.15-0.05}{1.1} = -2.0$
:::
:::
::::
