---
title: "Within-Subjects ANOVA"
subtitle: "PSY 348: Lecture 14"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Within-Subjects ANOVA | ⬡⬢⬡⬢⬡⬢"
format:
  revealjs:
    theme: ../custom.scss
    scrollable: true
    width: 1600
    height: 900
    incremental: true
    touch: true
    chalkboard: true
    lightbox: true
    code-fold: true
    slide-number: c/t
    include-after-body: ../title-bg.html
    drop:
      engine: webr
revealjs-plugins:
  - drop
  - codewindow
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(broom)
library(gt)
library(afex)

# Site tokens (see custom.scss / DESIGN_SYSTEM.md)
sage  <- "#5C715E"  # $accent
sage2 <- "#8FAF9E"  # $accent2
sage3 <- "#A7C7B5"  # $accent3
ink   <- "#2F3E46"  # $body-color
hot   <- "#B5483B"  # "look here" highlight
amber <- "#E8A33D"

theme_deck <- function(base_size = 22) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(color = ink),
      axis.text = element_text(color = ink),
      plot.title = element_text(color = ink, face = "bold"),
      legend.position = "none"
    )
}

f2 <- function(x) formatC(x, format = "f", digits = 2)
fp <- function(p) ifelse(p < .001, "< .001", paste("=", sub("^0", "", sprintf("%.3f", p))))

# ---- Helpers ----------------------------------------------------------
# Simulate a repeated-measures design: condition mean + person effect + noise
sim_rm <- function(n, means, sd_subj, sd_err, seed, lo = -Inf, hi = Inf) {
  set.seed(seed)
  u <- rnorm(n, 0, sd_subj)
  tibble(id = factor(rep(1:n, each = length(means))),
         cond = factor(rep(names(means), times = n), levels = names(means)),
         y = rep(means, times = n) + rep(u, each = length(means)) +
           rnorm(n * length(means), 0, sd_err)) |>
    mutate(y = pmin(pmax(y, lo), hi))
}

# Classic repeated-measures ANOVA table (Subjects / Time / Error)
rm_table <- function(d) {
  a <- tidy(aov(y ~ cond + Error(id), data = d))
  sub <- a |> filter(stratum == "id", term == "Residuals")
  eff <- a |> filter(term == "cond")
  err <- a |> filter(stratum == "Within", term == "Residuals")
  tibble(Source = c("Time (condition)", "Subjects", "Error"),
         df = c(eff$df, sub$df, err$df),
         SS = c(eff$sumsq, sub$sumsq, err$sumsq),
         MS = c(eff$meansq, sub$meansq, err$meansq),
         `F` = c(eff$statistic, NA, NA),
         p = c(eff$p.value, NA, NA))
}
rm_gt <- function(tab, size = 26) {
  tab |>
    mutate(p = ifelse(is.na(p), NA, ifelse(p < .001, "<.001", sprintf("%.3f", p)))) |>
    gt() |>
    fmt_number(columns = c(SS, MS, `F`), decimals = 2) |>
    fmt_number(columns = df, decimals = 0) |>
    sub_missing(missing_text = "--") |>
    tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
    opt_row_striping() |>
    cols_align(align = "center") |>
    tab_options(table.width = pct(100), table.font.size = size)
}
rm_apa <- function(tab) {
  sprintf("F(%d, %d) = %s, p %s", tab$df[1], tab$df[3], f2(tab$`F`[1]), fp(tab$p[1]))
}
means_sd <- function(d) d |> group_by(cond) |> summarise(M = mean(y), SD = sd(y))
rm_plot <- function(d, ylab, title) {
  ggplot(d, aes(cond, y, group = id)) +
    geom_line(color = sage3, alpha = .6) +
    geom_point(color = sage, alpha = .6, size = 1.8) +
    stat_summary(aes(group = 1), fun = mean, geom = "line", color = hot, linewidth = 1.6) +
    stat_summary(aes(group = 1), fun = mean, geom = "point", color = hot, size = 4) +
    labs(title = title, x = NULL, y = ylab) +
    theme_deck(20)
}

# ---- Pizza and productivity (n = 30) ---------------------------------
pizza <- sim_rm(30, c("Before lunch" = 5, "10 min after" = 7, "1 hour after" = 4.5),
                sd_subj = 1.5, sd_err = 1, seed = 123)
pz_tab <- rm_table(pizza)
pz_ms  <- means_sd(pizza)
# the same data analysed as if they were different people
pz_between <- tidy(aov(y ~ cond, data = pizza))
pz_F_between <- pz_between$statistic[1]
pz_post <- pairwise.t.test(pizza$y, pizza$cond, paired = TRUE, p.adjust.method = "bonferroni")
pz_ges <- {
  fit <- aov_ez("id", "y", pizza, within = "cond", anova_table = list(es = c("ges", "pes"), correction = "none"),
                return = "afex_aov")
  as.data.frame(anova(fit))
}

# ---- Engagement intervention (n = 100) -------------------------------
set.seed(2025)
n1 <- 100
b1 <- rnorm(n1, 4.8, 1.6); w1a <- b1 + rnorm(n1, .7, .5); w1b <- w1a + rnorm(n1, .8, .5)
clip <- function(x) pmin(pmax(x, 1), 10)
eng <- tibble(id = factor(rep(1:n1, each = 3)),
              cond = factor(rep(c("baseline", "week1", "week2"), times = n1),
                            levels = c("baseline", "week1", "week2")),
              y = as.vector(rbind(clip(b1), clip(w1a), clip(w1b))))
eng_tab <- rm_table(eng)

# ---- Active learning (n = 80) ----------------------------------------
set.seed(2025)
n2 <- 80
b2 <- rnorm(n2, 4.5, 1.4); m2 <- b2 + rnorm(n2, 1.0, .6); e2 <- m2 + rnorm(n2, .8, .5)
act <- tibble(id = factor(rep(1:n2, each = 3)),
              cond = factor(rep(c("beginning", "mid", "end"), times = n2),
                            levels = c("beginning", "mid", "end")),
              y = as.vector(rbind(clip(b2), clip(m2), clip(e2))))
act_tab <- rm_table(act)
act_ms  <- means_sd(act)

# ---- Practice studies (n = 30 each) ----------------------------------
mem  <- sim_rm(30, c(Immediate = 8, After24Hours = 6, After1Week = 4), 1.2, 1.1, seed = 301)
clin <- sim_rm(30, c(LowDose = 5, MediumDose = 6.5, HighDose = 8), 1.2, 1.0, seed = 302)
ther <- sim_rm(30, c(Baseline = 10, MidTreatment = 7, PostTreatment = 5), 1.8, 1.4, seed = 303)
prod <- sim_rm(30, c(ProductA = 6, ProductB = 7, ProductC = 5), 0.9, 0.9, seed = 304)
strs <- sim_rm(30, c(Rest = 50, Task = 70, Recovery = 55), 4, 4, seed = 305)
mem_t <- rm_table(mem); clin_t <- rm_table(clin); ther_t <- rm_table(ther)
prod_t <- rm_table(prod); strs_t <- rm_table(strs)
```

# When to Use Repeated Measures ANOVA

Use Repeated Measures ANOVA when:

- You have the **same participants measured more than once** (e.g., pre-test/post-test).

- You're interested in **changes over time** or **under different conditions**.

  - Testing whether students' enjoyment of statistics improves after receiving lollipops for 2 weeks.

# Between vs. Within Subjects ANOVA

::::::: columns
:::: {.column width="50%"}
::: callout-tip
## Between-subjects ANOVA

- Compares means **between groups** (Group A vs. B vs. C).

- Error comes from individual differences **between people**.
:::
::::

:::: {.column width="50%"}
::: callout-important
## Within-subjects (repeated measures) ANOVA

- Compares means **within the same participants** over time or conditions.

- **Controls for** individual differences.
:::
::::
:::::::

# Pizza and Productivity Study

## A Repeated-Measures Design

:::::: columns
:::: {.column width="50%"}
::: callout-tip
## Design

- **Objective:** test whether productivity changes over time.

- **IV:** time

  1.  Before lunch
  2.  10 minutes after pizza
  3.  1 hour after pizza

- The **same participants** are measured at each timepoint.

- **DV:** number of emails written in 10 minutes.
:::
::::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.2
rm_plot(pizza, "Emails written (10 min)", "Each line is one person") +
  scale_x_discrete(labels = c("Before", "10 min", "1 hour"))
```
:::
::::::

## Visualizing the Data

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4.6
rm_plot(pizza, "Productivity (emails sent)", "Individual and mean productivity over time") +
  labs(subtitle = "red = group mean")
```

## Why Repeated-Measures ANOVA?

- Can't use multiple paired-samples t-tests.

  - Alpha inflation increases the Type I error rate.

- Repeated-Measures ANOVA allows us to:

  - Compare within-subject changes across time.

  - **Control for individual variability.**

  - Test for an overall **time effect**.

## Within-Subjects ANOVA

### Where the variance comes from

- **Between-subjects** ANOVA splits variance into **between-group** and **error** (within-group) parts.

- **Within-subjects** ANOVA has a third source, because each person contributes several scores:

:::::: term-list
::: fragment

[Time]{.term-pill}

: How the **means** vary across the time points (the effect we care about).
:::

::: fragment

[Subjects]{.term-pill}

: How people differ from **each other overall**.
:::

::: fragment

[Error]{.term-pill}

: What is left over once time and subjects are accounted for.
:::
::::::

## Within-Subjects ANOVA

### Partitioning the variability

```{dot}
digraph {
  node[shape = "box", style = "rounded,filled", fillcolor = "white", color = "#5C715E", fontname = "Fira Sans"]
  edge[color = "#5C715E"]
  bgcolor = "#E9F1EC"

  A[label="Total Variability"]
  B[label="Time Variability"]
  C[label="Within-Groups Variability"]
  D[label="Subject Variability"]
  E[label="Error Variability"]

  A -> B
  A -> C
  C -> D
  C -> E
}
```

## Within-Subjects ANOVA

### The key idea

::: callout-tip
## Removing the person

$SS_{W} = SS_{Subjects} + SS_{Error}$

$SS_{Error} = SS_{W} - SS_{Subjects}$
:::

- In between-subjects ANOVA, differences between people sit **inside** the error term.

- In within-subjects ANOVA we **subtract them out**, so the error is smaller and the test is more **powerful**.

## Calculating Repeated-Measures ANOVA

::: callout-tip
## The F ratio

$F = \dfrac{MS_{Time}}{MS_{Error}}$

$MS_{Time} = \dfrac{SS_{Time}}{df_{Time}} = \dfrac{SS_{Time}}{k - 1}$

$MS_{Error} = \dfrac{SS_{W} - SS_{Subjects}}{df_{Error}} = \dfrac{SS_{W} - SS_{Subjects}}{(n-1)(k-1)}$
:::

- $k$ is the number of time points (or conditions) and $n$ is the number of participants.

## Degrees of Freedom

::: callout-tip
## Rules

- Time: $df_{Time} = k - 1$

- Subjects: $df_{Subjects} = n - 1$

- Error: $df_{Error} = (n - 1) \times (k - 1)$
:::

## Degrees of Freedom

### Example

> **20 students** rate their **stress** at **three time points** during finals week: three days before the exam, the night before, and after the exam.

What are the df's for a one-way repeated-measures ANOVA on Time?

<details>

<summary>Reveal answer</summary>

- $df_{Time} = 3 - 1 = 2$

- $df_{Subjects} = 20 - 1 = 19$

- $df_{Error} = 19 \times 2 = 38$

</details>

## Pizza Study Results

### The ANOVA table

```{r}
#| echo: false
rm_gt(pz_tab)
```

::: {.callout-tip appearance="minimal"}
$F(`r pz_tab$df[1]`, `r pz_tab$df[3]`) = `r f2(pz_tab$F[1])`$, $p$ `r fp(pz_tab$p[1])`: productivity changes significantly across the three time points.
:::

## Pizza Study Results

### In R

::: {.codewindow .r .codewindow-dark}
repeated_measures.R

```{r}
#| echo: true
#| code-fold: false
library(afex)

# id = who; within = the repeated factor
fit <- aov_ez(id = "id", dv = "y", data = pizza,
              within = "cond", return = "afex_aov")

# the corrected ANOVA table (Greenhouse-Geisser if needed)
anova(fit)
```
:::

- `summary(fit)` also prints Mauchly's test and the other corrections.


## Pizza Study Results

### Why within-subjects helps

::::::::: side-by-cols
::::: stat-card
::: stat-num
`r f2(pz_tab$F[1])`
:::

::: stat-sub
**Repeated-measures** F (removes person differences)
:::
:::::

::::: stat-card
::: stat-num
`r f2(pz_F_between)`
:::

::: stat-sub
**Between-subjects** F (same data, ignoring the person)
:::
:::::
:::::::::

- The same scores give a much larger F once we account for the fact that each person was measured three times.

## Repeated-Measures ANOVA

### Assumptions

::: callout-important
## Assumptions

- **Normality** of the differences.

- **Sphericity:** the variances of all pairwise differences are about equal.

  - Test: **Mauchly's Test**.

- If sphericity is violated, use the **Greenhouse-Geisser correction**.

  - Always check sphericity when the within-subject factor has **more than two** levels.
:::

## Repeated-Measures ANOVA

### Interpretation

- $F(`r pz_tab$df[1]`, `r pz_tab$df[3]`) = `r f2(pz_tab$F[1])`$, $p$ `r fp(pz_tab$p[1])`

  - Main effect of time on productivity.

  - Productivity changes significantly across the timepoints.

    - What would be your next step?

## Repeated-Measures ANOVA

### Effect size

::: callout-tip
## Partial eta squared

$\eta_p^2 = \dfrac{SS_{Time}}{SS_{Time} + SS_{Error}} = \dfrac{`r f2(pz_tab$SS[1])`}{`r f2(pz_tab$SS[1])` + `r f2(pz_tab$SS[3])`} =`r f2(pz_tab$SS[1] / (pz_tab$SS[1] + pz_tab$SS[3]))`$
:::

- Benchmarks are the same as other ANOVAs: about .01 small, .06 medium, .14 large.

## Follow-Up

### Post-hoc comparisons

- A significant F says that **some** time points differ, not which.

- Run paired t-tests with a **Bonferroni correction**:

::: {.codewindow .r .codewindow-dark}
posthoc.R

```{r}
#| echo: true
#| code-fold: false
pairwise.t.test(pizza$y, pizza$cond,
                paired = TRUE, p.adjust.method = "bonferroni")
```
:::

## Follow-Up

### Reading the comparisons

```{r}
#| echo: false
pz_pw <- as.data.frame(as.table(pz_post$p.value)) |>
  filter(!is.na(Freq)) |>
  transmute(Comparison = paste(Var2, "vs.", Var1),
            `Adjusted p` = ifelse(Freq < .001, "<.001", sprintf("%.3f", Freq)))
pz_pw |>
  gt() |>
  opt_row_striping() |>
  tab_options(table.font.size = 28, table.width = pct(60))
```

- The adjusted p-values already account for running three comparisons.

## Within-Subjects ANOVA

### Reporting F

If asked to report findings in terms of the Null Hypothesis (H~0~), you should report findings as:

- Reject $H_0$, or

- Fail to Reject $H_0$

- If asked to report findings in general or for publication, you need to report $F(df_{Time}, df_{Error}) = F, p$, an effect size, and the means and standard deviations at each time point.

## Within-Subjects ANOVA

### Reporting the pizza study

> Productivity differed significantly across the three time points, *`r rm_apa(pz_tab)`*, $\eta_p^2$ = `r f2(pz_tab$SS[1] / (pz_tab$SS[1] + pz_tab$SS[3]))`. Productivity was highest 10 minutes after pizza (*M* = `r f2(pz_ms$M[2])`, *SD* = `r f2(pz_ms$SD[2])`) and lowest an hour after pizza (*M* = `r f2(pz_ms$M[3])`, *SD* = `r f2(pz_ms$SD[3])`); before lunch it was *M* = `r f2(pz_ms$M[1])` (*SD* = `r f2(pz_ms$SD[1])`).

## Repeated-Measures ANOVA

### Example: an engagement intervention

> Prof. Brocker evaluates a new **engagement intervention**. **100 students** rate their enjoyment of statistics (`1` to `10`) at the start of the semester, after week 1 of a statistics warm-up activity, and after week 2.

::: {.callout-tip .nonincremental}
## Find the degrees of freedom

-   $df_{Time}$ = ?

-   $df_{Subjects}$ = ?

-   $df_{Error}$ = ?

-   $n$ = ?
:::

## Repeated-Measures ANOVA

### Engagement intervention: answers

::: {.callout-important .nonincremental}
## Degrees of freedom

-   $df_{Time} = k - 1 = 3 - 1 = 2$

-   $df_{Subjects} = n - 1 = 100 - 1 = 99$

-   $df_{Error} = (n - 1)(k - 1) = 99 \times 2 = 198$

-   $n = 100$
:::

## Repeated-Measures ANOVA

### Engagement intervention: results

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(eng_tab)
```

*`r rm_apa(eng_tab)`*
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4.6
eng |>
  group_by(cond) |>
  summarise(m = mean(y), se = sd(y) / sqrt(n())) |>
  ggplot(aes(cond, m, group = 1)) +
  geom_line(color = sage, linewidth = 1.4) +
  geom_point(color = sage, size = 3.5) +
  geom_errorbar(aes(ymin = m - se, ymax = m + se), width = .1, color = ink) +
  labs(title = "Enjoyment of statistics", x = NULL, y = "Mean enjoyment (1-10)") +
  theme_deck(20)
```
:::
:::::

## Repeated-Measures ANOVA

### Example: active-learning practices

> Professor Brocker tests whether structured **active-learning practices** improve `80` students' attitudes toward statistics. He collects a `1-10` enjoyment rating at the **beginning** of the semester, **mid-semester** (after four modules), and the **end** (after eight modules).

::: {.callout-tip .nonincremental}
## Find the degrees of freedom

1.  $df_{Time}$ = ?
2.  $df_{Participants}$ = ?
3.  $df_{Error}$ = ?
:::

## Repeated-Measures ANOVA

### Active-learning practices: answers

::: {.callout-important .nonincremental}
## Degrees of freedom

1.  $df_{Time} = k - 1 = 3 - 1 = 2$
2.  $df_{Participants} = n - 1 = 80 - 1 = 79$
3.  $df_{Error} = (n - 1)(k - 1) = 79 \times 2 = 158$
:::

## Repeated-Measures ANOVA

### Active-learning practices: results

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(act_tab)
```

*`r rm_apa(act_tab)`*
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4.6
act |>
  group_by(cond) |>
  summarise(m = mean(y), se = sd(y) / sqrt(n())) |>
  ggplot(aes(cond, m, group = 1)) +
  geom_line(color = sage, linewidth = 1.4) +
  geom_point(color = sage, size = 3.5) +
  geom_errorbar(aes(ymin = m - se, ymax = m + se), width = .1, color = ink) +
  labs(title = "Attitudes across the semester", x = NULL, y = "Mean attitude (1-10)") +
  theme_deck(20)
```
:::
:::::

## Interpreting F

::: callout-tip
## What F means

$MS_{Time}$ is made up of $MS_{Error}$ + the theoretical change over time:

- $MS_{Time} = \text{change over time} + MS_{Error}$

- If there is no change: $MS_{Time} = 0 + MS_{Error}$, so $F = \dfrac{MS_{Error}}{MS_{Error}} = 1$.

- If there is a change: $F = 1 + \dfrac{\text{change}}{MS_{Error}}$, which is greater than 1.
:::

- When $F \leq 1$, it's not likely to be significant.

## Hypothesis Testing: ANOVA

::: callout-tip
## Decision rule

- $F \leq 1$: fail to reject the null hypothesis.

- $F > 1$: compare to the critical value (or use the *p*-value). Reject $H_0$ only if $p < \alpha$.
:::

## Reporting F

### Active-learning example

> Attitudes toward statistics differed significantly across the semester, *`r rm_apa(act_tab)`*. Attitudes were highest at the end of the semester (*M* = `r f2(act_ms$M[act_ms$cond == "end"])`, *SD* = `r f2(act_ms$SD[act_ms$cond == "end"])`) and lowest at the beginning (*M* = `r f2(act_ms$M[act_ms$cond == "beginning"])`, *SD* = `r f2(act_ms$SD[act_ms$cond == "beginning"])`). Mid-semester attitudes were *M* = `r f2(act_ms$M[act_ms$cond == "mid"])` (*SD* = `r f2(act_ms$SD[act_ms$cond == "mid"])`).

## Data Manipulation

### Wide and long format

- When collected, data can be in either **wide** or **long** format.

- Below are the same five participants in both formats.

```{r}
#| echo: false
set.seed(7)
wide <- tibble(Participant = factor(1:5),
               Immediate = rnorm(5, 8, 1.5), After24Hours = rnorm(5, 6, 1.5),
               After1Week = rnorm(5, 4, 1.5))
long <- wide |> pivot_longer(cols = !Participant, names_to = "Time", values_to = "Memory")

wide |>
  gt() |>
  fmt_number(decimals = 1) |>
  tab_header(title = "Wide data", subtitle = "Each row is one participant") |>
  opt_row_striping() |>
  tab_options(table.font.size = 22, table.width = pct(60))
```

## Data Manipulation

### Wide and long format

:::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
long |>
  gt() |>
  fmt_number(columns = Memory, decimals = 1) |>
  tab_header(title = "Long data", subtitle = "Each row is one participant at one time") |>
  opt_row_striping() |>
  tab_options(table.font.size = 18, table.width = pct(90))
```
:::

:::: {.column width="50%"}
::: {.codewindow .r .codewindow-dark}
pivot.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
library(tidyr)

# wide -> long: one row per person per time
long <- wide |>
  pivot_longer(cols = !Participant,
               names_to = "Time",
               values_to = "Memory")
```
:::

- `aov()` and `aov_ez()` need the **long** format.
::::
::::::

## Practice

### 1. Memory recall

> 30 participants study a word list and recall it **immediately**, **after 24 hours**, and **after 1 week**. Does recall change over time?

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(mem_t, size = 20)
```

<details>

<summary>Reveal answer</summary>

*`r rm_apa(mem_t)`*: recall decreased significantly over time.

</details>
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2.5
rm_plot(mem, "Words recalled", "Memory recall")
```
:::
:::::


## Practice

### 2. Clinical trial

> 30 patients each receive a **low**, **medium**, and **high** dose of a medication (in separate sessions) and report symptom improvement. Does dose matter?

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(clin_t, size = 20)
```

<details>

<summary>Reveal answer</summary>

*`r rm_apa(clin_t)`*: improvement increased significantly with dose.

</details>
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2.5
rm_plot(clin, "Symptom improvement", "Clinical trial")
```
:::
:::::


## Practice

### 3. Behavioral therapy progress

> 30 participants rate their anxiety at **baseline**, **mid-treatment**, and **post-treatment**.

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(ther_t, size = 20)
```

<details>

<summary>Reveal answer</summary>

*`r rm_apa(ther_t)`*: anxiety decreased.

</details>
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2.5
rm_plot(ther, "Anxiety", "Behavioral therapy")
```
:::
:::::


## Practice

### 4. Consumer product ratings

> 30 consumers each rate **three products** (A, B, and C).

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(prod_t, size = 20)
```

<details>

<summary>Reveal answer</summary>

*`r rm_apa(prod_t)`*: the ratings differed by product. A post-hoc test would show which products differ.

</details>
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2.5
rm_plot(prod, "Rating", "Product ratings")
```
:::
:::::


## Practice

### 5. Physiological stress

> 30 participants' heart rate is measured at **rest**, during a **task**, and during **recovery**.

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
rm_gt(strs_t, size = 20)
```

<details>

<summary>Reveal answer</summary>

*`r rm_apa(strs_t)`*: heart rate differed significantly across the phases.

</details>
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2.5
rm_plot(strs, "Heart rate (bpm)", "Physiological stress")
```
:::
:::::


## Summary: Within-Subjects ANOVA

- Compares **three or more** related means (the same people measured repeatedly).

- Splits the variability into time, subjects, and error, and **removes the subject differences** from the error term.

- $F = MS_{Time} / MS_{Error}$ with $df = (k - 1,\ (n - 1)(k - 1))$.

- Check sphericity (Mauchly's Test) and use the Greenhouse-Geisser correction if it is violated.

- Follow a significant $F$ with post-hoc paired comparisons (Bonferroni).

- Report $F(df_{Time}, df_{Error})$, $p$, an effect size, and the means and standard deviations.