---
title: "Exam 1 Review"
subtitle: "Chapters 1 - 4"
author: "Dave Brocker"
institute: "Farmingdale State College"
format: 
  revealjs:
    theme: ../custom.scss
    incremental: true   
    touch: true
    chalkboard: true
    width: 1600
    height: 900
    scrollable: true
    lightbox: true
    drop:
      engine: webr
revealjs-plugins:
  - drop
---

## Learning Goals

- Compute sample mean and standard deviation by hand.
- Interpret what the mean and SD tell us about data.
- Standardize raw scores into Z-scores.
- Interpret Z-scores (direction and distance from mean).
- Relate Z-scores to probability under a normal curve.

## 1. Mean Calculation

### Sample Mean Calculation

> A psychology researcher collects the following data on the number of hours five participants spent practicing a task: `3, 5, 8, 2, 7`.

**Question:** What is the average number of hours spent practicing?

::: panel-tabset
## Solution

<p>$\bar{x} = \frac{\sum(3+5+8+2+7)}{5} = \frac{25}{5} = 5\text{ hours}$</p>

## In R

```{r}
#| echo: true
mean(c(3,5,8,2,7))
```
:::

## 2. Mean Calculation

### Sample Mean Calculation

> A study measures the changes in mood scores before and after an intervention. The changes in mood scores are: `-4, -2, 0, 3, 5`.

**Question:** Calculate the mean change in mood score.

::: panel-tabset
## Solution

<p>$\bar{x} = \frac{\sum(-4 + (-2) + 0 + 3 + 5)}{5} = \frac{2}{5} = 0.4$</p>

## In R

```{r}
#| echo: true
mean(c(-4,-2,0,3,5))
```
:::

## 3. Standard Deviation Calculation

### Sample Standard Deviation

> A psychologist measures the number of stressful events experienced by a small group of participants over a week: `2, 3, 6, 4, 5`.

**Question:** Calculate the standard deviation of stressful events.

::: panel-tabset
## Solution

-   Mean: $\bar{x} = \frac{2+3+6+4+5}{5} = 4$
-   Sum of squares: $(2-4)^2+(3-4)^2+(6-4)^2+(4-4)^2+(5-4)^2 = 4+1+4+0+1 = 10$
-   $s = \sqrt{\frac{10}{5-1}} = \sqrt{2.5} \approx 1.58$

## In R

```{r}
#| echo: true
sd(c(2,3,6,4,5))
```
:::


## 4. Standard Deviation Calculation

### Sample Standard Deviation

> **Sample Standard Deviation** Participants' anxiety scores in a clinical trial are recorded as: `4, 7, 3, 5, 8, 6, 2`.

**Question:** Compute the standard deviation of these anxiety scores.

::: panel-tabset
## Solution

-   Mean: $\bar{x} = \frac{4+7+3+5+8+6+2}{7} = \frac{35}{7} = 5$
-   Sum of squares: $1+4+4+0+9+1+9 = 28$
-   $s = \sqrt{\frac{28}{7-1}} = \sqrt{4.67} \approx 2.16$

## In R

```{r}
#| echo: true
sd(c(4,7,3,5,8,6,2))
```
:::

## 5. Standard Deviation Calculation

### Interpreting Standard Deviation in Psychology

> A researcher finds that the standard deviation of reaction times in a cognitive task is `1.5` seconds. Another researcher finds a standard deviation of `0.5` seconds for a similar task.

**Question:** Which task shows more variability in participants' reaction times, and what might this indicate about task difficulty?

<details>

<summary>Solution</summary>

<p>A standard deviation of `1.5` is is larger than `0.5` is, so the first task shows **more variability**.</p>

</details>

## 6. Z-Score Calculation

### Z-Score for a Single Value

> In a personality study, the average score on an extraversion scale is `50`, with a standard deviation of `10`. One participant scores `70`.

**Question:** What is the Z-score for this participant?

::: panel-tabset
## Solution

<p>$z = \frac{X - \mu}{\sigma} = \frac{70 - 50}{10} = \frac{20}{10} = 2$</p>

## In R

```{r}
#| echo: true

(70-50)/10
```
:::

## 7. Z-Score Calculation:

### Interpreting Z-Scores

> A participant in a self-esteem survey has a Z-score of `-1.5`. The mean self-esteem score for the survey is `40`, and the standard deviation is `5`.

**Question:** Did this participant score above or below the mean, and by how many standard deviations?

<details>

<summary>Solution</summary>

<p>A Z-score of `-1.5` is **below** the mean. In raw score terms: $X = \mu + z\sigma = 40 + (-1.5)(5) = 40 - 7.5 = 32.5.$ So the participant scored `32.5`, which is `1.5` standard deviations **below** the mean.</p>

</details>

## 8. Z-Score Calculation

### Z-Score with Negative Value

> The mean stress score of a group is `25`, with a standard deviation of `4`. If a participant has a stress score of `18`...

**Question:** Calculate their Z-score and interpret its meaning in the context of stress levels.

::: panel-tabset
## Solution

<p>$z = \frac{18 - 25}{4} = \frac{-7}{4} = -1.75.$ The Z-score is $z \approx -1.75$, meaning this participant's stress score is `1.75` **standard deviations below** the group mean — they have substantially lower stress than average.</p>

## In R

```{r}
#| echo: true
(18-25)/4
```
:::

## 9. Z-Score Calculation

### Multiple Z-Scores

> In a clinical assessment, the average depression score is `60`, with a standard deviation of `8`. Two participants have scores of `72` and `48`.

**Question:** Calculate the Z-scores for both participants and explain what the values tell us about their relative positions in the group.

::: panel-tabset
## Solution

-   Participant 1 (72): $z_1 = \frac{72 - 60}{8} = 1.5$. Participant 2 (48): $z_2 = \frac{48 - 60}{8} = -1.5$.
-   The first is `1.5` SD **above** the mean (higher depression than average), the second `1.5` SD **below** (lower). They are symmetric around the mean.

## In R

```{r}
#| echo: true
(72-60)/8;(48-60)/8
```
:::


## 10. Z-Scores to Raw Scores

### Working Backwards

> The distribution of scores on a math test is approximately normal with mean $\mu = 100$ and standard deviation $\sigma = 15$.

**Question:** What raw score corresponds to $z = 1.2$?

::: panel-tabset
## Solution

<p>Solve $z = \frac{x-\mu}{\sigma}$ for $x$: $x = \mu + z\sigma = 100 + 1.2(15) = 118$</p>

## In R

```{r}
#| echo: true
1.2*15 + 100
```
:::

## Reading the Z-Table

::: callout-tip
## The table only ever gives the area to the **left** of z

1.  **Sketch** the curve and shade what the question asks for.
2.  **Left of z:** read the table value directly.
3.  **Right of z:** 1 minus the table value (subtract the *area*, not $z$).
4.  **Between two z's:** big area minus small area.
:::

Anxiety scores are approximately normal with $\mu = 50$ and $\sigma = 10$.

## 11. Probability: Left Tail

### Below a Score

**Question:** What proportion of people score **below 42**?

::: panel-tabset
## Solution

<ul>

<li>$z = \frac{42-50}{10} = -0.80$</li>

<li>Shade the left tail. The table gives the left area directly: $.2119$</li>

<li>About **21.2%** score below 42.</li>

</ul>

## In R

```{r}
#| echo: true
pnorm(-0.80)                       # z-table
pnorm(42, mean = 50, sd = 10)      # unrounded z
```
:::

## 12. Probability: Right Tail

### Above a Score

**Question:** What proportion of people score **above 65**?

::: panel-tabset
## Solution

<ul>

<li>$z = \frac{65-50}{10} = 1.50$</li>

<li>Shade the right tail. The table gives the area to the *left* ($.9332$), so subtract from 1: $1 - .9332 = .0668$</li>

<li>About **6.7%** score above 65.</li>

</ul>

## In R

```{r}
#| echo: true
1 - pnorm(1.50)                                       # z-table
pnorm(65, mean = 50, sd = 10, lower.tail = FALSE)     # unrounded z
```
:::

## 13. Probability: Between Two Scores

### Between Two Values

**Question:** What proportion of people score **between 45 and 60**?

::::: columns
::: {.column width="50%"}
::: panel-tabset
## Solution

<ul>

<li>$z_{45} = \frac{45-50}{10} = -0.50$ and $z_{60} = \frac{60-50}{10} = 1.00$</li>

<li>Table: left of $1.00$ is $.8413$; left of $-0.50$ is $.3085$</li>

<li>Big area minus small area: $.8413 - .3085 = .5328$</li>

<li>About **53.3%** score between 45 and 60.</li>

</ul>

## In R

```{r}
#| echo: true
pnorm(1.00) - pnorm(-0.50)                            # z-table
pnorm(60, 50, 10) - pnorm(45, 50, 10)                 # unrounded z
```
:::

:::
:::::