---
title: "Paired-Samples T-Test"
subtitle: "PSY 348: Lecture 10"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "Paired-Samples T-Test | ⬡⬢⬡⬢⬡⬢"
format:
  revealjs:
    theme: ../custom.scss
    incremental: true
    touch: true
    chalkboard: true
    lightbox: true
    scrollable: true
    code-fold: true
    slide-number: c/t
    include-after-body: ../title-bg.html
    width: 1600
    height: 900
    drop:
      engine: webr
revealjs-plugins:
  - drop
  - codewindow
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(broom)
library(gt)

# Site tokens (see custom.scss / DESIGN_SYSTEM.md)
sage  <- "#5C715E"  # $accent
sage2 <- "#8FAF9E"  # $accent2
sage3 <- "#A7C7B5"  # $accent3
ink   <- "#2F3E46"  # $body-color
hot   <- "#B5483B"  # "look here" highlight
amber <- "#E8A33D"

theme_deck <- function(base_size = 22) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(color = ink),
      axis.text = element_text(color = ink),
      plot.title = element_text(color = ink, face = "bold"),
      legend.text = element_text(color = ink)
    )
}

# Formatting helpers so every number on a slide comes from the data.
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fp <- function(p) ifelse(p < .001, "< .001", paste("=", sub("^0", "", sprintf("%.3f", p))))

# --- Running example: stress before/after a mindfulness program ---------
# Scores are whole numbers on a 0-20 stress survey. Post scores track pre
# scores (people who start more stressed stay somewhat more stressed) and
# drop by about 2 points on average, so the pairs are positively correlated.
set.seed(20)
n    <- 30
pre  <- round(rnorm(n, 8, 2))
post <- round(6 + .7 * (pre - 8) + rnorm(n, 0, 1.2))
stress <- tibble(id = paste0("P", 1:n), Pre = pre, Post = post, D = Post - Pre)

tt    <- t.test(stress$Post, stress$Pre, paired = TRUE)   # post - pre
dbar  <- mean(stress$D)
sd_D  <- sd(stress$D)
se_D  <- sd_D / sqrt(n)
tstat <- unname(tt$statistic)
df_   <- unname(tt$parameter)
pval  <- tt$p.value
ci    <- tt$conf.int
dz    <- dbar / sd_D
r_pp  <- cor(stress$Pre, stress$Post)
ind   <- t.test(stress$Post, stress$Pre, var.equal = TRUE)  # ignores the pairing
tcrit <- qt(.975, df_)
apa_main <- sprintf("t(%d) = %s, p %s", df_, f2(tstat), fp(pval))
```

## Introduction to Paired-Samples t-Test

### What is a paired-samples t-test?

::: hero-def

[Paired-Samples t-Test]{.term-pill}

: A statistical method for comparing the means of **two related** sets of scores, such as the same people measured twice.
:::

## Purpose of Paired-Samples t-Test

### Why use it?

- To determine whether there is a statistically significant difference between two related group means.

- Useful in psychology for assessing change due to an intervention, therapy, or experiment.

## Application of Paired-Samples t-Test

### When to apply this test

:::::: badge-row
::: content-badge
[trending_up]{.material-icons}

#### Pretest-Posttest

The same group measured before and after
:::

::: content-badge
[people]{.material-icons}

#### Matched Pairs

Twins or matched participants
:::

::: content-badge
[repeat]{.material-icons}

#### Repeated Measures

The same group under different conditions
:::
::::::

## Connection to the One-Sample *t*-Test

::: {.callout-important appearance="minimal"}
A **paired-samples t-test** is a **one-sample t-test on the difference scores**.
:::

::::::: columns
:::: {.column width="50%"}
::: callout-tip
## One-sample

$t = \dfrac{\bar{x} - \mu_0}{s / \sqrt{n}}$

Is the sample mean different from a known value?
:::
::::

:::: {.column width="50%"}
::: callout-tip
## Paired

$t = \dfrac{\bar{D} - 0}{s_D / \sqrt{n}}$

Is the **mean difference** $\bar{D}$ different from 0?
:::
::::
:::::::

## Formula for Paired-Samples t-Test

### Mathematical representation

::: callout-tip
## Paired-Samples t-Test

$\large{t = \dfrac{\color{blue}{\bar{D}}}{\color{orange}{s_D}/\sqrt{n}}}$
:::

- $\color{blue}{\bar{D}}$: mean of the differences between paired observations.

- $\color{orange}{s_D}$: standard deviation of the differences.

- $n$: number of pairs.

- Degrees of freedom: $df = n - 1$.

## Theoretical Formula Explanation

### Understanding each component

:::::: term-list
::: fragment

[Mean difference $\bar{D}$]{.term-pill}

: The average of the differences between paired scores.
:::

::: fragment

[SD of differences $s_D$]{.term-pill}

: How much individual differences vary around the mean difference.
:::

::: fragment

[Number of pairs $n$]{.term-pill}

: A larger $n$ shrinks the standard error $s_D/\sqrt{n}$, so smaller differences can reach significance.
:::
::::::

## Paired-Samples t-Test

### Application in psychology research

> A psychologist measures stress levels of a group before and after a mindfulness program. Participants first answer a stress survey, next they participate in a mindfulness exercise. After the exercise, their stress scores are measured again and compared.

::: callout-tip
## Steps

1.  State the null and alternative hypotheses.
2.  Calculate the difference between post- and pre-intervention scores.
3.  Find the mean $\bar{D}$ and standard deviation $s_D$ of the differences.
4.  Apply the formula to calculate the t-value and check for significance.
:::

## Paired-Samples t-Test

### Step 1: hypotheses

::::::: columns
:::: {.column width="50%"}
::: callout-tip
## Null hypothesis $H_0$

There will be **no difference** in stress before and after the mindfulness exercise.

$\mu_\text{post} = \mu_\text{pre}$, or equivalently $\mu_D = 0$
:::
::::

:::: {.column width="50%"}
::: callout-important
## Alternative hypothesis $H_A$

There will be a **decrease** in stress after the mindfulness exercise.

$\mu_\text{post} < \mu_\text{pre}$, or equivalently $\mu_D < 0$
:::
::::
:::::::

## Paired-Samples t-Test

### Each person has two scores

::::: columns
::: {.column width="40%"}
```{r}
#| echo: false
stress |>
  select(Participant = id, Pre, Post) |>
  gt_preview(top_n = 6, bottom_n = 1) |>
  opt_row_striping() |>
  tab_options(table.font.size = 24, table.width = pct(95)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels())
```
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5
ggplot(stress, aes(Pre, Post)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = ink) +
  geom_point(size = 3, color = sage, alpha = .8) +
  coord_equal(xlim = c(0, 13), ylim = c(0, 13)) +
  labs(x = "Pre-mindfulness stress", y = "Post-mindfulness stress",
       title = paste0("Pre and post are related (r = ", f2(r_pp), ")"),
       subtitle = "dashed = no change") +
  theme_deck(18)
```
:::
:::::

## Paired-Samples t-Test

### Step 2: the difference scores

::::: columns
::: {.column width="55%"}
```{r}
#| echo: false
stress |>
  select(Participant = id, Pre, Post, `D = Post - Pre` = D) |>
  gt_preview(top_n = 6, bottom_n = 1) |>
  opt_row_striping() |>
  tab_options(table.font.size = 26, table.width = pct(95)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels())
```
:::

::: {.column width="45%"}
- $D$ is **post minus pre**.

- A **negative** $D$ means stress went **down**.

- From here on, the test only looks at the column of $D$ values.
:::
:::::

## Paired-Samples t-Test

### Steps 3 and 4: construct the formula (manually)

::: callout-tip
## Mean and SD of the differences

$\bar{D} = `r f2(dbar)`$ and $s_D = `r f2(sd_D)`$ across $n = 30$ pairs
:::

::: callout-tip
## Compute t

$t = \dfrac{\bar{D}}{s_D/\sqrt{n}} = \dfrac{`r f2(dbar)`}{`r f2(sd_D)`/\sqrt{30}} = `r f2(tstat)`$
:::

::: callout-tip
## Decide

At $\alpha = .05$ (two-tailed) and $df = 29$, $t_{critical} = \pm `r f3(tcrit)`$. Since $|t| = `r f2(abs(tstat))` > `r f3(tcrit)`$, we **reject** $H_0$.
:::

## Paired-Samples t-Test

### Construct the formula (computationally)

::: {.codewindow .r .codewindow-dark}
paired_t.R

```{r}
#| echo: true
#| code-fold: false
# post - pre, so a decrease in stress is a negative difference
t.test(stress$Post, stress$Pre, paired = TRUE)
```
:::

## Paired-Samples t-Test

### Reading the output

```{r}
#| echo: false
tt |>
  tidy() |>
  rename(d = estimate, t = statistic, df = parameter) |>
  mutate(`CI[ll,uu]` = paste0("[", round(conf.low, 2), ", ", round(conf.high, 2), "]"),
         p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) |>
  select(d, t, df, p, `CI[ll,uu]`) |>
  gt() |>
  fmt_number(columns = c(d, t), decimals = 2) |>
  fmt_number(columns = df, decimals = 0) |>
  opt_row_striping() |>
  tab_options(table.font.size = 32, table.width = pct(70)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels())
```

- d = mean difference (post minus pre, $\bar{D}$)

- t = t-test statistic

- df = $n - 1$ (30 - 1)

- p = probability of a difference at least this extreme if $H_0$ is true

- CI\[ll,uu\] = 95% confidence interval for the mean difference

## One-Tailed or Two-Tailed?

### Matching the test to the hypothesis

- Our alternative is **directional**: $H_A: \mu_\text{post} < \mu_\text{pre}$.

- `t.test()` is two-tailed by default. A directional test uses `alternative`:

::: {.codewindow .r .codewindow-dark}
one_tailed.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
# post - pre, testing for a DEcrease
t.test(stress$Post, stress$Pre, paired = TRUE, alternative = "less")
```
:::

- The t-statistic is the same. The one-tailed p-value is **half** the two-tailed p-value (when the effect is in the predicted direction), and the confidence interval becomes one-sided.

- Choose the direction **before** you see the data. Here the conclusion is the same either way: p `r fp(pval)`.

## Why Pairing Helps

### The same data, analyzed two ways

::::::::: side-by-cols
::::: stat-card
::: stat-num
`r f2(tstat)`
:::

::: stat-sub
**Paired** t-test (uses the pairing)
:::
:::::

::::: stat-card
::: stat-num
`r f2(unname(ind$statistic))`
:::

::: stat-sub
**Independent** t-test (ignores the pairing)
:::
:::::
:::::::::

- People differ a lot from each other (some start stressed, some start calm). That **between-person** variability swamps the average change in an independent-samples test.

- Pairing compares each person to **themselves**, removing that variability. Because pre and post scores are correlated (r = `r f2(r_pp)`), the differences are much less spread out than the raw scores.

## Effect Size: Cohen's *d*

### Why report effect size?

- Statistical significance ($p$) tells us **if** there is a difference.

- Effect size tells us **how big** that difference is.

- For a paired-samples design, the effect size is $d_z$:

::: callout-tip
## $d_z$

$d_z = \dfrac{\bar{D}}{s_D}$: the mean difference in units of the standard deviation of the differences.
:::

```{r}
#| echo: false
tibble(
  `|d|` = c("0.20", "0.50", "0.80"),
  Interpretation = c("Small effect", "Medium effect", "Large effect")
) |>
  gt() |>
  opt_row_striping() |>
  tab_options(table.width = pct(50), table.font.size = 22)
```

## Cohen's d

### Mindfulness example

::::::: columns
:::: {.column width="50%"}
::: callout-tip
## By hand

$d_z = \dfrac{\bar{D}}{s_D} = \dfrac{`r f2(dbar)`}{`r f2(sd_D)`} = `r f2(dz)`$
:::

The size is $|d_z| = `r f2(abs(dz))`$, a **large** effect. The negative sign just means scores went **down**.
::::

:::: {.column width="50%"}
::: {.codewindow .r .codewindow-dark}
cohens_d.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
library(rstatix)
stress |>
  pivot_longer(c(Post, Pre),
               names_to = "time", values_to = "score") |>
  cohens_d(score ~ time, paired = TRUE)
```
:::

`rstatix` reports the same value: $d_z = `r f2(dz)`$.
::::
:::::::

## Paired T-Tests

### Visualizing results

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5.4
long <- stress |>
  pivot_longer(c(Pre, Post), names_to = "time", values_to = "score") |>
  mutate(time = factor(time, levels = c("Pre", "Post")))

p1 <- ggplot(long, aes(time, score, group = id)) +
  geom_line(color = sage3, linewidth = .8, alpha = .8) +
  geom_point(color = sage, size = 2.5) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", color = hot, linewidth = 1.8) +
  stat_summary(aes(group = 1), fun = mean, geom = "point", color = hot, size = 5) +
  labs(x = NULL, y = "Stress score", title = "Each line is one person",
       subtitle = "red = group mean") +
  theme_deck(20)

p2 <- ggplot(stress, aes(D)) +
  geom_histogram(binwidth = 1, fill = sage3, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = ink) +
  geom_vline(xintercept = dbar, color = hot, linewidth = 1.4) +
  labs(x = "Difference (post - pre)", y = "Number of people",
       title = "Distribution of the differences",
       subtitle = "red = mean difference") +
  theme_deck(20)

patchwork::wrap_plots(p1, p2, nrow = 1)
```

## Reporting Results

### APA-style

> A [paired-samples t-test]{.hl} was conducted to [compare stress before and after a mindfulness program]{.hl}. Results showed a [significant decrease]{.hl} in stress after the program, [*`r apa_main`*]{.hl}, 95% CI \[`r f2(ci[1])`, `r f2(ci[2])`\], $d_z$ = `r f2(dz)`.

- **What kind of test?** paired-samples t-test
- **What is being compared?** stress before and after the program
- **Is it significant?** significant decrease
- **What are the test results?** t, df, p, confidence interval, and effect size

## Interpreting Results

- We found a **statistically significant** decrease in stress ($p$ `r fp(pval)`).

- The **effect size** ($|d_z| =$ `r f2(abs(dz))`) indicates a *large* change.

- Combining significance and effect size gives us a fuller picture of the data.

> Always ask: is this difference not just *statistically*, but *practically* meaningful?

## Summary of Paired-Samples t-Test

### Key takeaways

- Paired-samples t-tests are ideal for comparing **two related** sets of scores.

- They help determine if interventions or conditions produce **significant changes**.

- Remember to check the **assumptions**:

::: callout-important
## Assumptions

1.  **The differences are normally distributed.** Check with a histogram, QQ plot, or Shapiro-Wilk test.
2.  **The pairs are dependent (related measures).** Same participants or matched pairs.
3.  **No major outliers in the differences.**
:::

## Practice

### Pre/post stress

```{r}
#| include: false
set.seed(101)
ex_1 <- tibble(
  ID = 1:30,
  Stress_Pre  = rnorm(30, mean = 20, sd = 4)
) |> mutate(Stress_Post = Stress_Pre - 4 + rnorm(30, 0, 3))

set.seed(102)
ex_2 <- tibble(
  ID = 1:25,
  RT_NoMusic = rnorm(25, mean = 520, sd = 60)
) |> mutate(RT_Music = RT_NoMusic - 20 + rnorm(25, 0, 25))

t1 <- t.test(ex_1$Stress_Post, ex_1$Stress_Pre, paired = TRUE)
dz1 <- mean(ex_1$Stress_Post - ex_1$Stress_Pre) / sd(ex_1$Stress_Post - ex_1$Stress_Pre)
t2 <- t.test(ex_2$RT_Music, ex_2$RT_NoMusic, paired = TRUE, alternative = "less")
dz2 <- mean(ex_2$RT_Music - ex_2$RT_NoMusic) / sd(ex_2$RT_Music - ex_2$RT_NoMusic)
apa_1 <- sprintf("t(%d) = %s, p %s", unname(t1$parameter), f2(unname(t1$statistic)), fp(t1$p.value))
apa_2 <- sprintf("t(%d) = %s, p %s (one-tailed)", unname(t2$parameter), f2(unname(t2$statistic)), fp(t2$p.value))
```

> A researcher measures 30 participants' stress before and after a mindfulness program (`ex_1`: `Stress_Pre`, `Stress_Post`). She predicts that stress will be **lower** afterward.

1.  State $H_0$ and $H_A$.
2.  Compute the difference scores (post minus pre).
3.  Run the paired-samples t-test, report $d_z$, and write an APA sentence.

::: {.codewindow .r .codewindow-dark}
practice_1.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
set.seed(101)
ex_1 <- tibble(
  ID = 1:30,
  Stress_Pre  = rnorm(30, mean = 20, sd = 4)
) |> mutate(Stress_Post = Stress_Pre - 4 + rnorm(30, 0, 3))
```
:::

<details>

<summary>Reveal answer</summary>

$H_0: \mu_D = 0$; $H_A: \mu_D < 0$. *`r apa_1`*, $d_z$ = `r f2(dz1)`: a significant, large decrease in stress.

</details>

## Practice

### Reaction time and music

> Do people respond faster with music playing? 25 participants complete a reaction-time task with no music and with music (`ex_2`: `RT_NoMusic`, `RT_Music`). The prediction is that reaction times will be **faster (lower)** with music.

1.  State $H_0$ and $H_A$.
2.  Compute the difference scores (music minus no music).
3.  Run a **one-tailed** paired-samples t-test (`alternative = "less"`) and decide.

::: {.codewindow .r .codewindow-dark}
practice_2.R

```{r}
#| echo: true
#| code-fold: false
#| eval: false
set.seed(102)
ex_2 <- tibble(
  ID = 1:25,
  RT_NoMusic = rnorm(25, mean = 520, sd = 60)
) |> mutate(RT_Music = RT_NoMusic - 20 + rnorm(25, 0, 25))
```
:::

<details>

<summary>Reveal answer</summary>

$H_0: \mu_D = 0$; $H_A: \mu_D < 0$. *`r apa_2`*, $d_z$ = `r f2(dz2)`. The decision depends on the result: compare p to $\alpha = .05$ and interpret both the decision and the effect size.

</details>